---
title: "ASAP Clinical Microbiome: 540 individuals - Nature Medicine 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
  word_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File import, filter for patients to exclude
```{r}
library(reshape2)
library(ggplot2)
library(dplyr)
library(car)
library(tidyr)
library(ggpubr)
library(nycflights13)
library(DT)
library(forcats)
library(ggtext)
library(datawizard)
library(PMCMRplus)
library(fabricatr)
library(rstatix)
library(Hmisc)
library(factoextra)
library(devtools)
library(RColorBrewer)
library(broom)
library(emmeans)
library(questionr)
library(MASS)
```

```{r}
#Load clinical data file (available through the UCL Research Data Repository at the link: https://doi.org/10.5522/04/30710741)
metadata_540ind <- read.csv("/.../ASAP_SCHAPIRA_clinical_dataset_540ind_2026_v2.csv", sep=",", dec = '.', stringsAsFactors = FALSE, header = TRUE)

dim(metadata_540ind)
```

# COHORT
```{r}
metadata_540ind = metadata_540ind %>% reorder_levels(three_groups, order = c("HC", "GBA_NMC", "PD"))

table(metadata_540ind$group)
table(metadata_540ind$three_groups)
table(metadata_540ind$cohort, metadata_540ind$three_groups)

metadata_540ind$three_groups = as.character(metadata_540ind$three_groups)
```

# SUBSETS
```{r}
# Create subset of data for comparison HC vs GBA-NMC
metadata_nmc_hc = filter(metadata_540ind, three_groups == "GBA_NMC" | three_groups == "HC")
dim(metadata_nmc_hc)
table(metadata_nmc_hc$three_groups)
```


```{r}
# Create subset of data for comparison GBA-NMC vs PD
metadata_nmc_pd = filter(metadata_540ind, three_groups == "GBA_NMC" | three_groups == "PD")
dim(metadata_nmc_pd)
table(metadata_nmc_pd$three_groups)
```


```{r}
# Create subset of data for comparison GBA-PD vs idiopathic PD
onlyPD = filter(metadata_540ind, three_groups == "PD")
dim(onlyPD)
table(onlyPD$group)
```


```{r}
# Create subset of data with HC and PD
metadata_hc_pd = filter(metadata_540ind, three_groups == "HC" | three_groups == "PD")
dim(metadata_hc_pd)
table(metadata_hc_pd$three_groups)
```

# RESULTS SUMMARY

```{r}
metadata_540ind = metadata_540ind %>% reorder_levels(three_groups, order = c("HC", "GBA_NMC", "PD"))

table_summ_540ind <- group_by(metadata_540ind, "group"=three_groups) %>% 
  summarise(

    mean_age = round(mean(age, na.rm=T),1), sd_age = round(sd(age, na.rm=T), 1),
    
    mean_bmi = round(mean(bmi, na.rm=T),1), sd_bmi = round(sd(bmi, na.rm=T), 1),
            
    mean_pd_onset = round(mean(pd_onset_calculated, na.rm=T),1), sd_pd_onset = round(sd(pd_onset_calculated, na.rm=T), 1),
            
    mean_pd_dur = round(mean(pd_dur, na.rm=T),1), sd_pd_dur = round(sd(pd_dur, na.rm=T), 1),
            
    mean_hy = round(median(hoehn_yahr, na.rm=T),1), sd_hy = round(sd(hoehn_yahr, na.rm=T), 1),
            
    mean_ledd = round(mean(ledd_total, na.rm=T),1), sd_ledd = round(sd(ledd_total, na.rm=T), 1),
            
    mean_updrs_i = round(mean(updrs_i_total, na.rm=T),1), sd_updrs_i = round(sd(updrs_i_total, na.rm=T), 1),
            
    mean_updrs_ii_total = round(mean(updrs_ii_total, na.rm=T),1), sd_updrs_ii_total = round(sd(updrs_ii_total, na.rm=T), 1),
            
    mean_updrs_iii_total = round(mean(updrs_iii_total, na.rm=T),1), sd_updrs_iii_total = round(sd(updrs_iii_total, na.rm=T), 1),
            
    mean_updrs_iv_total = round(mean(updrs_iv_total, na.rm=T),1), sd_updrs_iv_total = round(sd(updrs_iv_total, na.rm=T), 1),
            
    mean_updrs_total = round(mean(updrs_total, na.rm=T),1), sd_updrs_total = round(sd(updrs_total, na.rm=T), 1),
             
    mean_scopa_total = round(mean(scopa_total, na.rm=T),1), sd_scopa_total = round(sd(scopa_total, na.rm=T), 1),
  
       
    mean_scopa_gastrointestinal = round(mean(scopa_gastrointestinal, na.rm=T),1), sd_scopa_gastrointestinal = round(sd(scopa_gastrointestinal, na.rm=T), 1),
       
    mean_scopa_urinary = round(mean(scopa_urinary, na.rm=T),1), sd_scopa_urinary = round(sd(scopa_urinary, na.rm=T), 1),
       
    mean_scopa_cardiovascular = round(mean(scopa_cardiovascular, na.rm=T),1), sd_scopa_cardiovascular = round(sd(scopa_cardiovascular, na.rm=T), 1),
       
    mean_scopa_thermoregulatory = round(mean(scopa_thermoregulatory, na.rm=T),1), sd_scopa_thermoregulatory = round(sd(scopa_thermoregulatory, na.rm=T), 1),
       
    mean_scopa_pupillomotor = round(mean(scopa_pupillomotor, na.rm=T),1), sd_scopa_pupillomotor = round(sd(scopa_pupillomotor, na.rm=T), 1),
    
    mean_scopa_sexual = round(mean(scopa_sexual, na.rm=T),1), sd_scopa_sexual = round(sd(scopa_sexual, na.rm=T), 1),
    
    mean_wcs_total = round(mean(wcs_total, na.rm=T),1), sd_wcs_total = round(sd(wcs_total, na.rm=T), 1),
    
    mean_rbdsq_total = round(mean(rbdsq_total, na.rm=T),1), sd_rbdsq_total = round(sd(rbdsq_total, na.rm=T), 1),
    
    mean_upsit_total = round(mean(upsit_total, na.rm=T),1), sd_upsit_total = round(sd(upsit_total, na.rm=T), 1),
    
    mean_hads_anxiety_total = round(mean(hads_anxiety_total, na.rm=T),1), sd_hads_anxiety_total = round(sd(hads_anxiety_total, na.rm=T), 1),
    
    mean_hads_depression_total = round(mean(hads_depression_total, na.rm=T),1), sd_hads_depression_total = round(sd(hads_depression_total, na.rm=T), 1),
    
    mean_bdi_total = round(mean(bdi_total, na.rm=T),1), sd_bdi_total = round(sd(bdi_total, na.rm=T), 1),
    
    mean_moca_total = round(mean(moca_total, na.rm=T),1), sd_moca_total = round(sd(moca_total, na.rm=T), 1)
            )

glimpse(table_summ_540ind)
```


# DEMOGRAPHICS
```{r}
# Analyses are performed separately for HC vs GBA-NMC, and GBA-NMC vs PD 
```

## AGE
```{r}
#stats
tapply(metadata_540ind$age, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$age, metadata_540ind$three_groups, FUN = sd, na.rm=T)

res.aov.age <- aov(age ~ three_groups, data = metadata_540ind)
summary(res.aov.age)
pairwise.t.test(metadata_540ind$age, metadata_540ind$three_groups, p.adjust.method = "none", paired = FALSE)

t.test(age ~ three_groups, data = metadata_nmc_hc)
t.test(age ~ three_groups, data = metadata_nmc_pd)
```

## SEX
```{r}
###calculate proportions 
sex_summary = metadata_540ind %>%
              group_by(three_groups, sex) %>%
              summarise("count"= n())
sex_summary

sex_summary$proportion = (sex_summary$count / ave(sex_summary$count, sex_summary$three_groups, FUN = sum))*100
sex_summary = as.data.frame (sex_summary)
sex_summary
```

```{r}
table_sex_nmc_hc <- table(metadata_nmc_hc$sex, metadata_nmc_hc$three_groups)
table_sex_nmc_hc
chisq.test(table_sex_nmc_hc)
```

```{r}
table_sex_nmc_pd <- table(metadata_nmc_pd$sex, metadata_nmc_pd$three_groups)
table_sex_nmc_pd
chisq.test(table_sex_nmc_pd)
```

```{r}
table_sex_hc_pd <- table(metadata_hc_pd$sex, metadata_hc_pd$three_groups)
table_sex_hc_pd
chisq.test(table_sex_hc_pd)
```


## FAMILY HISTORY
```{r}
###calculate proportions for combined positive (first/second degree) vs negative FH
fhx2_summary = metadata_540ind %>%
              group_by(three_groups, fhx2) %>%
              summarise("count"= n())
fhx2_summary

fhx2_summary$proportion = (fhx2_summary$count / ave(fhx2_summary$count, fhx2_summary$three_groups, FUN = sum))*100
fhx2_summary = as.data.frame (fhx2_summary)
fhx2_summary
```

```{r}
table_fhx_nmc_hc <- table(metadata_nmc_hc$fhx2, metadata_nmc_hc$three_groups)
table_fhx_nmc_hc
chisq.test(table_fhx_nmc_hc)
```

```{r}
table_fhx_nmc_pd <- table(metadata_nmc_pd$fhx2, metadata_nmc_pd$three_groups)
table_fhx_nmc_pd
chisq.test(table_fhx_nmc_pd)
```


## EDUCATION
```{r}
#stats
tapply(metadata_540ind$ed_yrs, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$ed_yrs, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(ed_yrs ~ three_groups, data = metadata_nmc_hc)
wilcox.test(ed_yrs ~ three_groups, data = metadata_nmc_pd)
```

## PARTNERS
```{r}
partner_summary = metadata_540ind %>% filter(metadata_540ind$partner_updated != "NA") %>%
              group_by(three_groups, partner_updated) %>%
              summarise("count"= n())
partner_summary

partner_summary$proportion = (partner_summary$count / ave(partner_summary$count, partner_summary$three_groups, FUN = sum))*100
partner_summary = as.data.frame (partner_summary)
partner_summary
```


```{r}
table_partner = table(metadata_nmc_hc$partner_updated, metadata_nmc_hc$three_groups)
table_partner
chisq.test(table_partner)
```

```{r}
table_partner = table(metadata_nmc_pd$partner_updated, metadata_nmc_pd$three_groups)
table_partner
chisq.test(table_partner)
```

## FAMILY MEMBERS
```{r}
fm_summary = metadata_540ind %>% filter(metadata_540ind$blood_related_updated != "NA") %>%
              group_by(three_groups, blood_related_updated) %>%
              summarise("count"= n())
fm_summary

fm_summary$proportion = (fm_summary$count / ave(fm_summary$count, fm_summary$three_groups, FUN = sum))*100
fm_summary = as.data.frame (fm_summary)
fm_summary
```

```{r}
table_fm_nmc_hc = table(metadata_nmc_hc$three_groups, metadata_nmc_hc$blood_related_updated)
table_fm_nmc_hc
chisq.test(table_fm_nmc_hc)
```

```{r}
table_fm_nmc_pd = table(metadata_nmc_pd$three_groups, metadata_nmc_pd$blood_related_updated)
table_fm_nmc_pd
chisq.test(table_fm_nmc_pd)
```


## BMI
```{r}
tapply(metadata_540ind$bmi, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$bmi, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(bmi ~ three_groups, data = metadata_nmc_hc)
wilcox.test(bmi ~ three_groups, data = metadata_nmc_pd)
```


# DISEASE FEATURES: ONLY FOR PD PATIENTS

## Age at onset
```{r}
#Age at onset
#stats
mean(onlyPD$pd_onset_calculated)
sd(onlyPD$pd_onset_calculated)
```

## Disease duration
```{r}
mean(onlyPD$pd_dur)
sd(onlyPD$pd_dur)
```

## Levodopa equivalent daily dose
```{r}
#stats
mean(onlyPD$ledd_total)
sd(onlyPD$ledd_total)
```

## DRUG NAIVE PD 
```{r}
table(onlyPD$pd_drug_naive)
```

## Advanced therapies
```{r}
table(onlyPD$pd_dbs) 
table(onlyPD$pd_apomorphine) 
table(onlyPD$pd_lcig) 
```

# CLINICAL FEATURES
```{r}
# Analyses are performed separately for HC vs GBA-NMC, and GBA-NMC vs PD; adjustment for multiple testing with BH correction is performed
```

## MDS-UPDRS part I (var1)
```{r}
#stats
tapply(metadata_540ind$updrs_i_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_i_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

updrs_i_total_compa_nmc_hc = wilcox.test(updrs_i_total ~ three_groups, data = metadata_nmc_hc)
updrs_i_total_compa_nmc_hc

updrs_i_total_compa_nmc_pd = wilcox.test(updrs_i_total ~ three_groups, data = metadata_nmc_pd)
updrs_i_total_compa_nmc_pd
```

## MDS-UPDRS part II (var2)
```{r}
#stats
tapply(metadata_540ind$updrs_ii_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_ii_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

updrs_ii_total_compa_nmc_hc = wilcox.test(updrs_ii_total ~ three_groups, data = metadata_nmc_hc)
updrs_ii_total_compa_nmc_hc

updrs_ii_total_compa_nmc_pd = wilcox.test(updrs_ii_total ~ three_groups, data = metadata_nmc_pd)
updrs_ii_total_compa_nmc_pd
```

## MDS-UPDRS part III (var3)
```{r}
#stats
tapply(metadata_540ind$updrs_iii_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_iii_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

updrs_iii_total_compa_nmc_hc = wilcox.test(updrs_iii_total ~ three_groups, data = metadata_nmc_hc)
updrs_iii_total_compa_nmc_hc

updrs_iii_total_compa_nmc_pd = wilcox.test(updrs_iii_total ~ three_groups, data = metadata_nmc_pd)
updrs_iii_total_compa_nmc_pd
```

## MDS-UPDRS part IV (not included as not applicable to HC or GBA-NMC)

## MDS-UPDRS total score (var4)
```{r}
#stats
tapply(metadata_540ind$updrs_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

updrs_total_compa_nmc_hc = wilcox.test(updrs_total ~ three_groups, data = metadata_nmc_hc)
updrs_total_compa_nmc_hc

updrs_total_compa_nmc_pd = wilcox.test(updrs_total ~ three_groups, data = metadata_nmc_pd)
updrs_total_compa_nmc_pd
```

## HOEHN & YAHR STAGE (var5)
```{r}
tapply(metadata_540ind$hoehn_yahr, metadata_540ind$three_groups, FUN = median, na.rm=T)
tapply(metadata_540ind$hoehn_yahr, metadata_540ind$three_groups, FUN = sd, na.rm=T)

hoehn_yahr_compa_nmc_hc = wilcox.test(hoehn_yahr ~ three_groups, data = metadata_nmc_hc)
hoehn_yahr_compa_nmc_hc

hoehn_yahr_compa_nmc_pd = wilcox.test(hoehn_yahr ~ three_groups, data = metadata_nmc_pd)
hoehn_yahr_compa_nmc_pd
```

## SCOPA (var6)
```{r}
#stats
tapply(metadata_540ind$scopa_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

scopa_total_compa_nmc_hc = wilcox.test(scopa_total ~ three_groups, data = metadata_nmc_hc)
scopa_total_compa_nmc_hc

scopa_total_compa_nmc_pd = wilcox.test(scopa_total ~ three_groups, data = metadata_nmc_pd)
scopa_total_compa_nmc_pd
```

## SCOPA gastrointestinal (var7)
```{r}
#stats
tapply(metadata_540ind$scopa_gastrointestinal, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_gastrointestinal, metadata_540ind$three_groups, FUN = sd, na.rm=T)

scopa_gastrointestinal_compa_nmc_hc = wilcox.test(scopa_gastrointestinal ~ three_groups, data = metadata_nmc_hc)
scopa_gastrointestinal_compa_nmc_hc

scopa_gastrointestinal_compa_nmc_pd = wilcox.test(scopa_gastrointestinal ~ three_groups, data = metadata_nmc_pd)
scopa_gastrointestinal_compa_nmc_pd
```

## SCOPA urinary (var8)
```{r}
#stats
tapply(metadata_540ind$scopa_urinary, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_urinary, metadata_540ind$three_groups, FUN = sd, na.rm=T)

scopa_urinary_compa_nmc_hc = wilcox.test(scopa_urinary ~ three_groups, data = metadata_nmc_hc)
scopa_urinary_compa_nmc_hc

scopa_urinary_compa_nmc_pd = wilcox.test(scopa_urinary ~ three_groups, data = metadata_nmc_pd)
scopa_urinary_compa_nmc_pd
```

## SCOPA cardiovascular (var9)
```{r}
#stats
tapply(metadata_540ind$scopa_cardiovascular, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_cardiovascular, metadata_540ind$three_groups, FUN = sd, na.rm=T)

scopa_cardiovascular_compa_nmc_hc = wilcox.test(scopa_cardiovascular ~ three_groups, data = metadata_nmc_hc)
scopa_cardiovascular_compa_nmc_hc

scopa_cardiovascular_compa_nmc_pd = wilcox.test(scopa_cardiovascular ~ three_groups, data = metadata_nmc_pd)
scopa_cardiovascular_compa_nmc_pd
```

## SCOPA thermoregulatory (var10)
```{r}
#stats
tapply(metadata_540ind$scopa_thermoregulatory, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_thermoregulatory, metadata_540ind$three_groups, FUN = sd, na.rm=T)

scopa_thermoregulatory_compa_nmc_hc = wilcox.test(scopa_thermoregulatory ~ three_groups, data = metadata_nmc_hc)
scopa_thermoregulatory_compa_nmc_hc

scopa_thermoregulatory_compa_nmc_pd = wilcox.test(scopa_thermoregulatory ~ three_groups, data = metadata_nmc_pd)
scopa_thermoregulatory_compa_nmc_pd
```

## SCOPA pupillomotor (var11)
```{r}
#stats
tapply(metadata_540ind$scopa_pupillomotor, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_pupillomotor, metadata_540ind$three_groups, FUN = sd, na.rm=T)

scopa_pupillomotor_compa_nmc_hc = wilcox.test(scopa_pupillomotor ~ three_groups, data = metadata_nmc_hc)
scopa_pupillomotor_compa_nmc_hc

scopa_pupillomotor_compa_nmc_pd = wilcox.test(scopa_pupillomotor ~ three_groups, data = metadata_nmc_pd)
scopa_pupillomotor_compa_nmc_pd
```

## SCOPA sexual (var12)
```{r}
#stats
tapply(metadata_540ind$scopa_sexual, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_sexual, metadata_540ind$three_groups, FUN = sd, na.rm=T)

scopa_sexual_compa_nmc_hc = wilcox.test(scopa_sexual ~ three_groups, data = metadata_nmc_hc)
scopa_sexual_compa_nmc_hc

scopa_sexual_compa_nmc_pd = wilcox.test(scopa_sexual ~ three_groups, data = metadata_nmc_pd)
scopa_sexual_compa_nmc_pd
```

## WCSS (var13)
```{r}
#stats
tapply(metadata_540ind$wcs_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$wcs_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wcs_total_compa_nmc_hc = wilcox.test(wcs_total ~ three_groups, data = metadata_nmc_hc)
wcs_total_compa_nmc_hc 

wcs_total_compa_nmc_pd = wilcox.test(wcs_total ~ three_groups, data = metadata_nmc_pd)
wcs_total_compa_nmc_pd
```

## RBD (var14)
```{r}
## Comparison of total score - not included in multiple comparison testing
tapply(metadata_540ind$rbdsq_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$rbdsq_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

rbdsq_total_compa_nmc_hc = wilcox.test(rbdsq_total ~ three_groups, data = metadata_nmc_hc)
rbdsq_total_compa_nmc_hc

rbdsq_total_compa_nmc_pd = wilcox.test(rbdsq_total ~ three_groups, data = metadata_nmc_pd)
rbdsq_total_compa_nmc_pd
```

```{r}
table(metadata_540ind$three_groups, metadata_540ind$rbd_affected_diff)

###calculate proportion
rbd_summary = metadata_540ind[!is.na(metadata_540ind$rbd_affected_diff),] %>%
              group_by(three_groups, rbd_affected_diff) %>%
              summarise("count"= n())
rbd_summary

rbd_summary$proportion = (rbd_summary$count / ave(rbd_summary$count, rbd_summary$three_groups, FUN = sum))*100
rbd_summary = as.data.frame (rbd_summary)
rbd_summary
```

```{r}
rbd.model_nmc_hc = glm(as.factor(rbd_affected_diff) ~ three_groups, data = metadata_nmc_hc, family = binomial())
summary(rbd.model_nmc_hc)
summary(rbd.model_nmc_hc)$coefficients
odds.ratio(rbd.model_nmc_hc, level = 0.95, signif.stars = TRUE)

rbd_compa_nmc_hc = as.data.frame(summary(rbd.model_nmc_hc)$coef)
rbd_compa_nmc_hc_p = as.data.frame(rbd_compa_nmc_hc$`Pr(>|z|)`)
rbd_compa_nmc_hc_pval = rbd_compa_nmc_hc_p[2,]
rbd_compa_nmc_hc_pval
```

```{r}
rbd.model_nmc_pd = glm(as.factor(rbd_affected_diff) ~ three_groups, data = metadata_nmc_pd, family = binomial())
summary(rbd.model_nmc_pd)
summary(rbd.model_nmc_pd)$coefficients
odds.ratio(rbd.model_nmc_pd, level = 0.95, signif.stars = TRUE)

rbd_compa_nmc_pd = as.data.frame(summary(rbd.model_nmc_pd)$coef)
rbd_compa_nmc_pd_p = as.data.frame(rbd_compa_nmc_pd$`Pr(>|z|)`)
rbd_compa_nmc_pd_pval = rbd_compa_nmc_pd_p[2,]
rbd_compa_nmc_pd_pval
```

## UPSIT (var15)
```{r}
#stats
tapply(metadata_540ind$upsit_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$upsit_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)
```

```{r}
res.regression.upsit_compa_nmc_hc <- lm(upsit_total ~ three_groups + age + sex, 
            data = metadata_nmc_hc)

summary(res.regression.upsit_compa_nmc_hc)
summary(res.regression.upsit_compa_nmc_hc)$coef
vif(res.regression.upsit_compa_nmc_hc)

upsit_compa_nmc_hc_res = as.data.frame(summary(res.regression.upsit_compa_nmc_hc)$coef)
upsit_compa_nmc_hc_p = as.data.frame(upsit_compa_nmc_hc_res$`Pr(>|t|)`)
upsit_compa_nmc_hc_p
upsit_compa_nmc_hc_pval = upsit_compa_nmc_hc_p[2,]
upsit_compa_nmc_hc_pval
```

```{r}
res.regression.upsit_compa_nmc_pd <- lm(upsit_total ~ three_groups + age + sex, 
            data = metadata_nmc_pd)

summary(res.regression.upsit_compa_nmc_pd)
summary(res.regression.upsit_compa_nmc_pd)$coef
vif(res.regression.upsit_compa_nmc_pd)

upsit_compa_nmc_pd_res = as.data.frame(summary(res.regression.upsit_compa_nmc_pd)$coef)
upsit_compa_nmc_pd_p = as.data.frame(upsit_compa_nmc_pd_res$`Pr(>|t|)`)
upsit_compa_nmc_pd_p
upsit_compa_nmc_pd_pval = upsit_compa_nmc_pd_p[2,]
upsit_compa_nmc_pd_pval
```

## HADS anxiety (var16)
```{r}
#stats
tapply(metadata_540ind$hads_anxiety_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$hads_anxiety_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

hads_anxiety_total_compa_nmc_hc = wilcox.test(hads_anxiety_total ~ three_groups, data = metadata_nmc_hc)
hads_anxiety_total_compa_nmc_hc

hads_anxiety_total_compa_nmc_pd = wilcox.test(hads_anxiety_total ~ three_groups, data = metadata_nmc_pd)
hads_anxiety_total_compa_nmc_pd
```

## HADS depression (var17)
```{r}
#stats
tapply(metadata_540ind$hads_depression_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$hads_depression_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

hads_depression_total_compa_nmc_hc = wilcox.test(hads_depression_total ~ three_groups, data = metadata_nmc_hc)
hads_depression_total_compa_nmc_hc

hads_depression_total_compa_nmc_pd = wilcox.test(hads_depression_total ~ three_groups, data = metadata_nmc_pd)
hads_depression_total_compa_nmc_pd
```

## BDI (var18)
```{r}
#stats
tapply(metadata_540ind$bdi_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$bdi_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

bdi_total_compa_nmc_hc = wilcox.test(bdi_total ~ three_groups, data = metadata_nmc_hc)
bdi_total_compa_nmc_hc

bdi_total_compa_nmc_pd = wilcox.test(bdi_total ~ three_groups, data = metadata_nmc_pd)
bdi_total_compa_nmc_pd
```

## MOCA (var19)
```{r}
#stats
tapply(metadata_540ind$moca_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$moca_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)
```

```{r}
# Comparison of total score - not included in multiple comparison testing
res.regression.moca_model_compa_nmc_hc = lm(formula = moca_total ~ three_groups + age + sex + ed_yrs, data = metadata_nmc_hc)
summary(res.regression.moca_model_compa_nmc_hc)
summary(res.regression.moca_model_compa_nmc_hc)$coefficients
```

```{r}
res.regression.moca_model_compa_nmc_pd = lm(formula = moca_total ~ three_groups + age + sex + ed_yrs, data = metadata_nmc_pd)
summary(res.regression.moca_model_compa_nmc_pd)
summary(res.regression.moca_model_compa_nmc_pd)$coefficients
```

```{r}
moca_summary = metadata_540ind %>% filter(metadata_540ind$moca_range != "NA") %>%
              group_by(three_groups, moca_range) %>%
              summarise("count"= n())
moca_summary

moca_summary$proportion = (moca_summary$count / ave(moca_summary$count, moca_summary$three_groups, FUN = sum))*100
moca_summary = as.data.frame (moca_summary)
moca_summary
```

```{r}
metadata_nmc_hc = metadata_nmc_hc %>% reorder_levels(three_groups, order = c("HC", "GBA_NMC"))
metadata_nmc_hc = metadata_nmc_hc %>% reorder_levels(moca_range, order = c("Normal", "Abnormal"))
metadata_nmc_hc$moca_range = as.factor(metadata_nmc_hc$moca_range)

moca_binary_model_compa_nmc_hc = glm(moca_range ~ three_groups + age + sex + ed_yrs, data = metadata_nmc_hc, family = binomial())
summary(moca_binary_model_compa_nmc_hc)
summary(moca_binary_model_compa_nmc_hc)$coefficients
odds.ratio(moca_binary_model_compa_nmc_hc, level = 0.95, signif.stars = TRUE)

moca_compa_nmc_hc = as.data.frame(summary(moca_binary_model_compa_nmc_hc)$coef)
moca_compa_nmc_hc_p = as.data.frame(moca_compa_nmc_hc$`Pr(>|z|)`)
moca_compa_nmc_hc_pval = moca_compa_nmc_hc_p[2,]
moca_compa_nmc_hc_pval
```

```{r}
metadata_nmc_pd = metadata_nmc_pd %>% reorder_levels(three_groups, order = c("GBA_NMC", "PD"))
metadata_nmc_pd = metadata_nmc_pd %>% reorder_levels(moca_range, order = c("Normal", "Abnormal"))
metadata_nmc_pd$moca_range = as.factor(metadata_nmc_pd$moca_range)

moca_binary_model_compa_nmc_pd = glm(moca_range ~ three_groups + age + sex + ed_yrs, data = metadata_nmc_pd, family = binomial())
summary(moca_binary_model_compa_nmc_pd)
summary(moca_binary_model_compa_nmc_pd)$coefficients
odds.ratio(moca_binary_model_compa_nmc_pd, level = 0.95, signif.stars = TRUE)

moca_compa_nmc_pd = as.data.frame(summary(moca_binary_model_compa_nmc_pd)$coef)
moca_compa_nmc_pd_p = as.data.frame(moca_compa_nmc_pd$`Pr(>|z|)`)
moca_compa_nmc_pd_pval = moca_compa_nmc_pd_p[2,]
moca_compa_nmc_pd_pval
```

# HC vs GBA-NMC: Clinical variables: adjustment for multiple comparison
```{r}
pvalues_clinical_nmc_hc = c(
                     round(updrs_i_total_compa_nmc_hc$p.value,4), #var1
                     round(updrs_ii_total_compa_nmc_hc$p.value,4), #var2 
                     round(updrs_iii_total_compa_nmc_hc$p.value,4), #var3
                     round(updrs_total_compa_nmc_hc$p.value,4), #var4
                     round(hoehn_yahr_compa_nmc_hc$p.value,4), #var5
                     
                     round(scopa_total_compa_nmc_hc$p.value,4), #var6
                     round(scopa_gastrointestinal_compa_nmc_hc$p.value,4), #var7
                     round(scopa_urinary_compa_nmc_hc$p.value,4), #var8
                     round(scopa_cardiovascular_compa_nmc_hc$p.value,4), #var9
                     round(scopa_thermoregulatory_compa_nmc_hc$p.value,4), #var10
                     round(scopa_pupillomotor_compa_nmc_hc$p.value,4), #var11
                     round(scopa_sexual_compa_nmc_hc$p.value,4), #var12
                     
                     round(wcs_total_compa_nmc_hc$p.value,4), #var13
                     round(rbd_compa_nmc_hc_pval, 4), #var14
                     round(upsit_compa_nmc_hc_pval, 4), #var15
                     round(hads_anxiety_total_compa_nmc_hc$p.value,4), #var16
                     round(hads_depression_total_compa_nmc_hc$p.value,4), #var17
                     round(bdi_total_compa_nmc_hc$p.value,4), #var18
                     round(moca_compa_nmc_hc_pval,4) #var19
                     )
print(pvalues_clinical_nmc_hc)
```

```{r}
pvalues_clinical_nmc_hc_ordered = pvalues_clinical_nmc_hc[order(pvalues_clinical_nmc_hc)]
print(pvalues_clinical_nmc_hc_ordered)
```

```{r}
fdrs_nmc_hc <-p.adjust(pvalues_clinical_nmc_hc_ordered, method="BH")
print(fdrs_nmc_hc)
```

# GBA-NMC vs PD: Clinical variables: adjustment for multiple comparison
```{r}
pvalues_clinical_nmc_pd = c(
                     round(updrs_i_total_compa_nmc_pd$p.value,4), #var1
                     round(updrs_ii_total_compa_nmc_pd$p.value,4), #var2 
                     round(updrs_iii_total_compa_nmc_pd$p.value,4), #var3
                     round(updrs_total_compa_nmc_pd$p.value,4), #var4
                     round(hoehn_yahr_compa_nmc_pd$p.value,4), #var5
                     
                     round(scopa_total_compa_nmc_pd$p.value,4), #var6
                     round(scopa_gastrointestinal_compa_nmc_pd$p.value,4), #var7
                     round(scopa_urinary_compa_nmc_pd$p.value,4), #var8
                     round(scopa_cardiovascular_compa_nmc_pd$p.value,4), #var9
                     round(scopa_thermoregulatory_compa_nmc_pd$p.value,4), #var10
                     round(scopa_pupillomotor_compa_nmc_pd$p.value,4), #var11
                     round(scopa_sexual_compa_nmc_pd$p.value,4), #var12
                     
                     round(wcs_total_compa_nmc_pd$p.value,4), #var13
                     round(rbd_compa_nmc_pd_pval, 4), #var14
                     round(upsit_compa_nmc_pd_pval, 4), #var15
                     round(hads_anxiety_total_compa_nmc_pd$p.value,4), #var16
                     round(hads_depression_total_compa_nmc_pd$p.value,4), #var17
                     round(bdi_total_compa_nmc_pd$p.value,4), #var18
                     round(moca_compa_nmc_pd_pval,4) #var19
                     )
print(pvalues_clinical_nmc_pd)
str(pvalues_clinical_nmc_pd)
test = as.data.frame(pvalues_clinical_nmc_pd)
```

```{r}
pvalues_clinical_nmc_pd_ordered = pvalues_clinical_nmc_pd[order(pvalues_clinical_nmc_pd)]
print(pvalues_clinical_nmc_pd_ordered)
```

```{r}
fdrs_nmc_pd <-p.adjust(pvalues_clinical_nmc_pd_ordered, method="BH")
print(fdrs_nmc_pd)
```

## PRODROMAL PROBABILITY
```{r}
table(metadata_540ind$three_groups, metadata_540ind$prob_prodromalPD)
```

# PD ONLY: GBA-PD vs iPD
# RESULTS SUMMARY
```{r}
table(onlyPD$group)
table(onlyPD$cohort, onlyPD$group)
```

```{r}
onlyPD = onlyPD %>% reorder_levels(group, order = c("GBA_PD", "sPD"))

table_summ_pd <- group_by(onlyPD, "group"=group) %>% 
  summarise(

    mean_age = round(mean(age, na.rm=T),1), sd_age = round(sd(age, na.rm=T), 1),
    
    mean_bmi = round(mean(bmi, na.rm=T),1), sd_bmi = round(sd(bmi, na.rm=T), 1),
            
    mean_pd_onset = round(mean(pd_onset_calculated, na.rm=T),1), sd_pd_onset = round(sd(pd_onset_calculated, na.rm=T), 1),
            
    mean_pd_dur = round(mean(pd_dur, na.rm=T),1), sd_pd_dur = round(sd(pd_dur, na.rm=T), 1),
            
    mean_hy = round(median(hoehn_yahr, na.rm=T),1), sd_hy = round(sd(hoehn_yahr, na.rm=T), 1),
            
    mean_ledd = round(mean(ledd_total, na.rm=T),1), sd_ledd = round(sd(ledd_total, na.rm=T), 1),
            
    mean_updrs_i = round(mean(updrs_i_total, na.rm=T),1), sd_updrs_i = round(sd(updrs_i_total, na.rm=T), 1),
            
    mean_updrs_ii_total = round(mean(updrs_ii_total, na.rm=T),1), sd_updrs_ii_total = round(sd(updrs_ii_total, na.rm=T), 1),
            
    mean_updrs_iii_total = round(mean(updrs_iii_total, na.rm=T),1), sd_updrs_iii_total = round(sd(updrs_iii_total, na.rm=T), 1),
            
    mean_updrs_iv_total = round(mean(updrs_iv_total, na.rm=T),1), sd_updrs_iv_total = round(sd(updrs_iv_total, na.rm=T), 1),
            
    mean_updrs_total = round(mean(updrs_total, na.rm=T),1), sd_updrs_total = round(sd(updrs_total, na.rm=T), 1),
             
    mean_scopa_total = round(mean(scopa_total, na.rm=T),1), sd_scopa_total = round(sd(scopa_total, na.rm=T), 1),
  
       
    mean_scopa_gastrointestinal = round(mean(scopa_gastrointestinal, na.rm=T),1), sd_scopa_gastrointestinal = round(sd(scopa_gastrointestinal, na.rm=T), 1),
       
    mean_scopa_urinary = round(mean(scopa_urinary, na.rm=T),1), sd_scopa_urinary = round(sd(scopa_urinary, na.rm=T), 1),
       
    mean_scopa_cardiovascular = round(mean(scopa_cardiovascular, na.rm=T),1), sd_scopa_cardiovascular = round(sd(scopa_cardiovascular, na.rm=T), 1),
       
    mean_scopa_thermoregulatory = round(mean(scopa_thermoregulatory, na.rm=T),1), sd_scopa_thermoregulatory = round(sd(scopa_thermoregulatory, na.rm=T), 1),
       
    mean_scopa_pupillomotor = round(mean(scopa_pupillomotor, na.rm=T),1), sd_scopa_pupillomotor = round(sd(scopa_pupillomotor, na.rm=T), 1),
    
    mean_scopa_sexual = round(mean(scopa_sexual, na.rm=T),1), sd_scopa_sexual = round(sd(scopa_sexual, na.rm=T), 1),
    
    mean_wcs_total = round(mean(wcs_total, na.rm=T),1), sd_wcs_total = round(sd(wcs_total, na.rm=T), 1),
    
    mean_rbdsq_total = round(mean(rbdsq_total, na.rm=T),1), sd_rbdsq_total = round(sd(rbdsq_total, na.rm=T), 1),
    
    mean_upsit_total = round(mean(upsit_total, na.rm=T),1), sd_upsit_total = round(sd(upsit_total, na.rm=T), 1),
    
    mean_hads_anxiety_total = round(mean(hads_anxiety_total, na.rm=T),1), sd_hads_anxiety_total = round(sd(hads_anxiety_total, na.rm=T), 1),
    
    mean_hads_depression_total = round(mean(hads_depression_total, na.rm=T),1), sd_hads_depression_total = round(sd(hads_depression_total, na.rm=T), 1),
    
    mean_bdi_total = round(mean(bdi_total, na.rm=T),1), sd_bdi_total = round(sd(bdi_total, na.rm=T), 1),
    
    mean_moca_total = round(mean(moca_total, na.rm=T),1), sd_moca_total = round(sd(moca_total, na.rm=T), 1)
            )

glimpse(table_summ_pd)
```

# DEMOGRAPHICS
## COHORT
```{r}
table(onlyPD$group)
table(onlyPD$group, onlyPD$cohort)
```

## AGE
```{r}
#stats
tapply(onlyPD$age, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$age, onlyPD$group, FUN = sd, na.rm=T)

t.test(age ~ group, data = onlyPD)
```

## SEX
```{r}
###calculate proportions 
sex_summary_onlyPD = onlyPD %>%
              group_by(group, sex) %>%
              summarise("count"= n())
sex_summary_onlyPD

sex_summary_onlyPD$proportion = (sex_summary_onlyPD$count / ave(sex_summary_onlyPD$count, sex_summary_onlyPD$group, FUN = sum))*100
sex_summary_onlyPD = as.data.frame (sex_summary_onlyPD)
sex_summary_onlyPD
```

```{r}
table_sex_onlyPD <- table(onlyPD$sex, onlyPD$group)
table_sex_onlyPD
chisq.test(table_sex_onlyPD)
```

## FAMILY HISTORY
```{r}
###calculate proportions for combined positive (first/second degree) vs negative FH
fhx2_summary_onlyPD = onlyPD %>%
              group_by(group, fhx2) %>%
              summarise("count"= n())
fhx2_summary_onlyPD

fhx2_summary_onlyPD$proportion = (fhx2_summary_onlyPD$count / ave(fhx2_summary_onlyPD$count, fhx2_summary_onlyPD$group, FUN = sum))*100
fhx2_summary_onlyPD = as.data.frame (fhx2_summary_onlyPD)
fhx2_summary_onlyPD
```

```{r}
table_fhx <- table(onlyPD$fhx2, onlyPD$group)
table_fhx
chisq.test(table_fhx)
```

## EDUCATION
```{r}
#stats
tapply(onlyPD$ed_yrs, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$ed_yrs, onlyPD$group, FUN = sd, na.rm=T)

wilcox.test(ed_yrs ~ group, data = onlyPD)
```

## PARTNERS
```{r}
partner_summary_onlyPD = onlyPD %>% filter(onlyPD$partner_updated != "NA") %>%
              group_by(group, partner_updated) %>%
              summarise("count"= n())
partner_summary_onlyPD

partner_summary_onlyPD$proportion = (partner_summary_onlyPD$count / ave(partner_summary_onlyPD$count, partner_summary_onlyPD$group, FUN = sum))*100
partner_summary_onlyPD = as.data.frame (partner_summary_onlyPD)
partner_summary_onlyPD
```

```{r}
table_partner = table(onlyPD$partner_updated, onlyPD$group)
table_partner
chisq.test(table_partner)
```

## FAMILY MEMBERS
```{r}
fm_summary_onlyPD = onlyPD %>% filter(onlyPD$blood_related_updated != "NA") %>%
              group_by(group, blood_related_updated) %>%
              summarise("count"= n())
fm_summary_onlyPD

fm_summary_onlyPD$proportion = (fm_summary_onlyPD$count / ave(fm_summary_onlyPD$count, fm_summary_onlyPD$group, FUN = sum))*100
fm_summary_onlyPD = as.data.frame (fm_summary_onlyPD)
fm_summary_onlyPD
```

```{r}
table_fm = table(onlyPD$group, onlyPD$blood_related_updated)
table_fm
chisq.test(table_fm)
```

## BMI
```{r}
tapply(onlyPD$bmi, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$bmi, onlyPD$group, FUN = sd, na.rm=T)

wilcox.test(bmi ~ group, data = onlyPD)
```

# DISEASE FEATURES
## Age at onset
```{r}
#Age at onset
tapply(onlyPD$pd_onset_calculated, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$pd_onset_calculated, onlyPD$group, FUN = sd, na.rm=T)

t.test(pd_onset_calculated ~ group, data = onlyPD)
```

## Disease duration
```{r}
mean(onlyPD$pd_dur)
sd(onlyPD$pd_dur)

wilcox.test(pd_dur ~ group, data = onlyPD)
```

## Levodopa equivalent daily dose
```{r}
#stats
mean(onlyPD$ledd_total)
sd(onlyPD$ledd_total)
wilcox.test(ledd_total ~ group, data = onlyPD)
```

## DRUG NAIVE PD 
```{r}
#calculate proportions of patients who underwent DBS in various groups
naive_summary = onlyPD %>%
                group_by(group, pd_drug_naive) %>% 
                summarise("count"= n())
naive_summary

naive_summary$proportion = (naive_summary$count / ave(naive_summary$count, naive_summary$group, FUN = sum))*100
naive_summary = as.data.frame(naive_summary)
naive_summary

table(onlyPD$pd_drug_naive)
table_naive = table(onlyPD$pd_drug_naive, onlyPD$group)
chisq.test(table_naive)
```

## DBS
```{r}
#calculate proportions of patients who underwent DBS in various groups
dbs_summary = onlyPD %>%
                group_by(group, pd_dbs) %>% 
                summarise("count"= n())
dbs_summary

dbs_summary$proportion = (dbs_summary$count / ave(dbs_summary$count, dbs_summary$group, FUN = sum))*100
dbs_summary = as.data.frame(dbs_summary)
dbs_summary

table(onlyPD$pd_dbs) # 40 patients with DBS
table_dbs = table(onlyPD$pd_dbs, onlyPD$group)
chisq.test(table_dbs)
```

# APOMORPHINE
```{r}
#calculate proportions of patients who underwent DBS in various groups
apo_summary = onlyPD %>%
                group_by(group, pd_apomorphine) %>% 
                summarise("count"= n())
apo_summary

apo_summary$proportion = (apo_summary$count / ave(apo_summary$count, apo_summary$group, FUN = sum))*100
apo_summary = as.data.frame(apo_summary)
apo_summary

table(onlyPD$pd_apomorphine) 
table_apo = table(onlyPD$pd_apomorphine, onlyPD$group)
fisher.test(table_apo)
```

# LCIG 
```{r}
#calculate proportions of patients who underwent DBS in various groups
lcig_summary = onlyPD %>%
                group_by(group, pd_lcig) %>% 
                summarise("count"= n())
lcig_summary

lcig_summary$proportion = (lcig_summary$count / ave(lcig_summary$count, lcig_summary$group, FUN = sum))*100
lcig_summary = as.data.frame(lcig_summary)
lcig_summary

table(onlyPD$pd_lcig) 
table_lcig = table(onlyPD$pd_lcig, onlyPD$group)
fisher.test(table_lcig)
```

# CLINICAL FEATURES
```{r}
# Analyses with adjustment for multiple testing with BH correction are performed
```

## MDS-UPDRS part I (var1)
```{r}
#stats
tapply(onlyPD$updrs_i_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$updrs_i_total, onlyPD$group, FUN = sd, na.rm=T)

updrs_i_total_compa_pd = wilcox.test(updrs_i_total ~ group, data = onlyPD)
updrs_i_total_compa_pd
```

## MDS-UPDRS part II (var2)
```{r}
#stats
tapply(onlyPD$updrs_ii_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$updrs_ii_total, onlyPD$group, FUN = sd, na.rm=T)

updrs_ii_total_compa_pd = wilcox.test(updrs_ii_total ~ group, data = onlyPD)
updrs_ii_total_compa_pd
```

## MDS-UPDRS part III (var3)
```{r}
#stats
tapply(onlyPD$updrs_iii_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$updrs_iii_total, onlyPD$group, FUN = sd, na.rm=T)

updrs_iii_total_compa_pd = wilcox.test(updrs_iii_total ~ group, data = onlyPD)
updrs_iii_total_compa_pd
```

## MDS-UPDRS part IV (var4)
```{r}
tapply(onlyPD$updrs_iv_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$updrs_iv_total, onlyPD$group, FUN = sd, na.rm=T)

updrs_iv_total_compa_pd = wilcox.test(updrs_iv_total ~ group, data = onlyPD)
updrs_iv_total_compa_pd
```

## MDS-UPDRS total score (var5)
```{r}
#stats
tapply(onlyPD$updrs_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$updrs_total, onlyPD$group, FUN = sd, na.rm=T)

updrs_total_compa_pd = wilcox.test(updrs_total ~ group, data = onlyPD)
updrs_total_compa_pd
```

## HOEHN & YAHR STAGE (var6)
```{r}
tapply(onlyPD$hoehn_yahr, onlyPD$three_groups, FUN = median, na.rm=T)
tapply(onlyPD$hoehn_yahr, onlyPD$three_groups, FUN = sd, na.rm=T)

hoehn_yahr_compa_pd = wilcox.test(hoehn_yahr ~ group, data = onlyPD)
hoehn_yahr_compa_pd
```

## SCOPA (var7)
```{r}
#stats
tapply(onlyPD$scopa_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$scopa_total, onlyPD$group, FUN = sd, na.rm=T)

scopa_total_compa_pd = wilcox.test(scopa_total ~ group, data = onlyPD)
scopa_total_compa_pd
```

## SCOPA gastrointestinal (var8)
```{r}
#stats
tapply(onlyPD$scopa_gastrointestinal, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$scopa_gastrointestinal, onlyPD$group, FUN = sd, na.rm=T)

scopa_gastrointestinal_compa_pd = wilcox.test(scopa_gastrointestinal ~ group, data = onlyPD)
scopa_gastrointestinal_compa_pd
```

## SCOPA urinary (var9)
```{r}
#stats
tapply(onlyPD$scopa_urinary, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$scopa_urinary, onlyPD$group, FUN = sd, na.rm=T)

scopa_urinary_compa_pd = wilcox.test(scopa_urinary ~ group, data = onlyPD)
scopa_urinary_compa_pd
```

## SCOPA cardiovascular (var10)
```{r}
#stats
tapply(onlyPD$scopa_cardiovascular, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$scopa_cardiovascular, onlyPD$group, FUN = sd, na.rm=T)

scopa_cardiovascular_compa_pd = wilcox.test(scopa_cardiovascular ~ group, data = onlyPD)
scopa_cardiovascular_compa_pd
```

## SCOPA thermoregulatory (var11)
```{r}
#stats
tapply(onlyPD$scopa_thermoregulatory, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$scopa_thermoregulatory, onlyPD$group, FUN = sd, na.rm=T)

scopa_thermoregulatory_compa_pd = wilcox.test(scopa_thermoregulatory ~ group, data = onlyPD)
scopa_thermoregulatory_compa_pd
```

## SCOPA pupillomotor (var12)
```{r}
#stats
tapply(onlyPD$scopa_pupillomotor, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$scopa_pupillomotor, onlyPD$group, FUN = sd, na.rm=T)

scopa_pupillomotor_compa_pd = wilcox.test(scopa_pupillomotor ~ group, data = onlyPD)
scopa_pupillomotor_compa_pd
```

## SCOPA sexual (var13)
```{r}
#stats
tapply(onlyPD$scopa_sexual, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$scopa_sexual, onlyPD$group, FUN = sd, na.rm=T)

scopa_sexual_compa_pd = wilcox.test(scopa_sexual ~ group, data = onlyPD)
scopa_sexual_compa_pd
```

## WCSS (var14)
```{r}
#stats
tapply(onlyPD$wcs_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$wcs_total, onlyPD$group, FUN = sd, na.rm=T)

wcs_total_compa_pd = wilcox.test(wcs_total ~ group, data = onlyPD)
wcs_total_compa_pd
```

## RBD (var15)
```{r}
# Comparison of total score - not included in multiple comparison testing
tapply(onlyPD$rbdsq_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$rbdsq_total, onlyPD$group, FUN = sd, na.rm=T)

rbdsq_total_compa_pd = wilcox.test(rbdsq_total ~ group, data = onlyPD)
rbdsq_total_compa_pd
```

```{r}
table(onlyPD$group, onlyPD$rbd_affected_diff)

###calculate proportion
rbd_summary_onlyPD = onlyPD[!is.na(onlyPD$rbd_affected_diff),] %>%
              group_by(group, rbd_affected_diff) %>%
              summarise("count"= n())
rbd_summary_onlyPD

rbd_summary_onlyPD$proportion = (rbd_summary_onlyPD$count / ave(rbd_summary_onlyPD$count, rbd_summary_onlyPD$group, FUN = sum))*100
rbd_summary_onlyPD = as.data.frame (rbd_summary_onlyPD)
rbd_summary_onlyPD
```

```{r}
rbd.model_pd = glm(as.factor(rbd_affected_diff) ~ group, data = onlyPD, family = binomial())
summary(rbd.model_pd)
summary(rbd.model_pd)$coefficients
odds.ratio(rbd.model_pd, level = 0.95, signif.stars = TRUE)

rbd_compa_pd = as.data.frame(summary(rbd.model_pd)$coef)
rbd_compa_pd_p = as.data.frame(rbd_compa_pd$`Pr(>|z|)`)
rbd_compa_pd_pval = rbd_compa_pd_p[2,]
rbd_compa_pd_pval
```

## UPSIT (var16)
```{r}
#stats
tapply(onlyPD$upsit_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$upsit_total, onlyPD$group, FUN = sd, na.rm=T)
```

```{r}
res.regression.upsit_compa_pd <- lm(upsit_total ~ group + age + sex, 
            data = onlyPD)

summary(res.regression.upsit_compa_pd)
summary(res.regression.upsit_compa_pd)$coef
vif(res.regression.upsit_compa_pd)

upsit_compa_pd_res = as.data.frame(summary(res.regression.upsit_compa_pd)$coef)
upsit_compa_pd_p = as.data.frame(upsit_compa_pd_res$`Pr(>|t|)`)
upsit_compa_pd_p
upsit_compa_pd_pval = upsit_compa_pd_p[2,]
upsit_compa_pd_pval
```

## HADS anxiety (var17)
```{r}
#stats
tapply(onlyPD$hads_anxiety_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$hads_anxiety_total, onlyPD$group, FUN = sd, na.rm=T)

hads_anxiety_total_compa_pd = wilcox.test(hads_anxiety_total ~ group, data = onlyPD)
hads_anxiety_total_compa_pd
```

## HADS depression (var18)
```{r}
#stats
tapply(onlyPD$hads_depression_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$hads_depression_total, onlyPD$group, FUN = sd, na.rm=T)

hads_depression_total_compa_pd = wilcox.test(hads_depression_total ~ group, data = onlyPD)
hads_depression_total_compa_pd
```

## BDI (var19)
```{r}
#stats
tapply(onlyPD$bdi_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$bdi_total, onlyPD$group, FUN = sd, na.rm=T)

bdi_total_compa_pd = wilcox.test(bdi_total ~ group, data = onlyPD)
bdi_total_compa_pd
```

## MOCA (var20)
```{r}
#stats
tapply(onlyPD$moca_total, onlyPD$group, FUN = mean, na.rm=T)
tapply(onlyPD$moca_total, onlyPD$group, FUN = sd, na.rm=T)
```

```{r}
# Comparison of total score - not included in multiple comparison testing
moca_model_compa_pd = lm(formula = moca_total ~ group + age + sex + ed_yrs, data = onlyPD)
summary(moca_model_compa_pd)
summary(moca_model_compa_pd)$coefficients
```

```{r}
moca_summary_onlyPD = onlyPD %>% filter(onlyPD$moca_range != "NA") %>%
              group_by(group, moca_range) %>%
              summarise("count"= n())
moca_summary_onlyPD

moca_summary_onlyPD$proportion = (moca_summary_onlyPD$count / ave(moca_summary_onlyPD$count, moca_summary_onlyPD$group, FUN = sum))*100
moca_summary_onlyPD = as.data.frame (moca_summary_onlyPD)
moca_summary_onlyPD
```

```{r}
onlyPD = onlyPD %>% reorder_levels(moca_range, order = c("Normal", "Abnormal"))
onlyPD$moca_range = as.factor(onlyPD$moca_range)

moca_binary_model_compa_pd = glm(moca_range ~ group + age + sex + ed_yrs, data = onlyPD, family = binomial())
summary(moca_binary_model_compa_pd)
summary(moca_binary_model_compa_pd)$coefficients
odds.ratio(moca_binary_model_compa_pd, level = 0.95, signif.stars = TRUE)

moca_model_compa_pd_res = as.data.frame(summary(moca_binary_model_compa_pd)$coef)
moca_compa_pd_p = as.data.frame(moca_model_compa_pd_res$`Pr(>|z|)`)
moca_compa_pd_pval = moca_compa_pd_p[2,]
moca_compa_pd_pval
```

# GBA-PD vs idiopathic PD: Clinical variables: adjustment for multiple comparison
```{r}
pvalues_clinical_pd = c(
                     round(updrs_i_total_compa_pd$p.value,4), #var1
                     round(updrs_ii_total_compa_pd$p.value,4), #var2 
                     round(updrs_iii_total_compa_pd$p.value,4), #var3
                      round(updrs_iv_total_compa_pd$p.value,4), #var4
                     round(updrs_total_compa_pd$p.value,4), #var5
                     round(hoehn_yahr_compa_pd$p.value,4), #var6
                     
                     round(scopa_total_compa_pd$p.value,4), #var7
                     round(scopa_gastrointestinal_compa_pd$p.value,4), #var8
                     round(scopa_urinary_compa_pd$p.value,4), #var9
                     round(scopa_cardiovascular_compa_pd$p.value,4), #var10
                     round(scopa_thermoregulatory_compa_pd$p.value,4), #var11
                     round(scopa_pupillomotor_compa_pd$p.value,4), #var12
                     round(scopa_sexual_compa_pd$p.value,4),  #var13
                     
                     round(wcs_total_compa_pd$p.value,4),  #var14
                     round(rbd_compa_pd_pval, 4),  #var15
                     round(upsit_compa_pd_pval, 4),  #var16
                     round(hads_anxiety_total_compa_pd$p.value,4),  #var17
                     round(hads_depression_total_compa_pd$p.value,4),  #var18
                     round(bdi_total_compa_pd$p.value,4),  #var19
                     round(moca_compa_pd_pval,4)  #var20
                     )
print(pvalues_clinical_pd)

test = as.data.frame(print(pvalues_clinical_pd))
```

```{r}
pvalues_clinical_pd_ordered = pvalues_clinical_pd[order(pvalues_clinical_pd)]
print(pvalues_clinical_pd_ordered)
```

```{r}
fdrs_compa_pd <-p.adjust(pvalues_clinical_pd_ordered, method="BH")
print(fdrs_compa_pd)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
