---
title: "Microbiome signature of Parkinson disease in healthy and genetically at-risk individuals - Figures"
subtitle: "In progress"
author: "Yani Ren, Mallia Geiger, Victoria Meslier"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, results = "hide", fig.show = "hide", warning = FALSE, message = F) 
```

This Rmarkdown script includes analyses of the gut microbiota of 464 fecal samples, from our cohort ASAP, between the HC, GBA_NMC and PD groups, focusing on species (MSP), functional potential of the modules and their links with clinical variables.

```{r, message=FALSE}
# Load necessary package
library(dplyr)
library(stringr)
library(momr)
library(rstatix)
library(ggpubr)
library(ggpmisc)
library(gtools)
library(Hmisc)
library(tidyr)
library(tibble)
library(cowplot)
library(grid)
library(ComplexHeatmap)
library(gridExtra)
library(ggplot2)
source("script_function_ASAPh_paper.R")
``` 



```{r}
# color for the 3 groups (group_bis)  
group_bis_color <- c(HC = "#009E73", GBA_NMC =  "#F0E442", PD = "#CC79A7")
color_HC_GBA_NMC = c(group_bis_color[c(1,2)])
color_HC_PD = c(group_bis_color[c(1,3)])
color_GBA_NMC_PD = c(group_bis_color[c(2,3)])
 
# color for enriched and depleted species categories 
PD_enriched_col <- "#D55E00"
PD_depleted_col <- "#0072B2"

PD_enriched_hue <- c("#E69F00", PD_enriched_col)
PD_depleted_hue <- c("#56B4E9", PD_depleted_col)
``` 



```{r}
# Load data
# Metadata 
load("./data/ASAPh_metadata_fecal_baseline_464ind_10102025_victoria.RData")
# MSP abundance matrix
load("./data/ASAPh_microbial_data_fecal_baseline_464ind_10102025_victoria.RData")
```


```{r}
temp_metadata <- metadata.fecal.baseline %>% 
    mutate(group = factor(group, levels = c("HC", "GBA_NMC", "GBA_PD", "sPD"))) 

# form PD by grouping GBA_PD and sPD individuals together 
temp_metadata <- temp_metadata %>% 
  mutate(group_bis = case_when(group == "HC" ~ "HC", 
                               group %in% c("GBA_PD", "sPD") ~ "PD", 
                               group == "GBA_NMC" ~ "GBA_NMC")) 
temp_metadata <- temp_metadata %>% mutate(group_bis = factor(group_bis, levels = c("HC", "GBA_NMC", "PD")))


NMC_ind <- temp_metadata %>% filter(group_bis == "GBA_NMC") %>% pull(MGP_ID_fecal_baseline)
PD_ind <- temp_metadata %>% filter(group_bis == "PD") %>% pull(MGP_ID_fecal_baseline)
HC_ind <- temp_metadata %>% filter(group_bis == "HC") %>% pull(MGP_ID_fecal_baseline)


dim(msp_fecal_baseline_464ind) # 2054 MSP in reality, because manually added msp_1786 with abundance = 0 for each individual in order to avoid warning 
# one MSP was removed compared to 469ind MSP matrix : msp_1130 (unclassified)

# compute prevalence of MSP - 464ind
prevalence <- (rowSums(msp_fecal_baseline_464ind>0)/ncol(msp_fecal_baseline_464ind)) * 100 
# keep MSP where prevalence > 10% 
msp_fecal.filt10 <- msp_fecal_baseline_464ind[prevalence > 10,]
dim(msp_fecal.filt10)
```




```{r}
# MSP analysis by group

# loop to perform pairwise comparison using testRelation2 
my_compa <- combn(x=levels(temp_metadata$group), m = 2, simplify = T)
for (i in 1:ncol(my_compa)){
    # select individuals belong to select groups
    select_data <- temp_metadata %>% 
      filter(group %in% c(my_compa[1,i], my_compa[2,i])) %>% 
      mutate(group = factor(group, levels = c(my_compa[1,i], my_compa[2,i])))
    
    # extract MSP abundance for selected individuals 
    select_msp_data <- msp_fecal.filt10[,rownames(select_data)]
    
    testR_name = paste("testR_msp", my_compa[1,i], my_compa[2,i], sep = '_')
    # vec_testR_name <- append(vec_testR_name, testR_name)
    
    # perform pairwise comparisons and assign output to a new variable name
    assign(testR_name, testRelations2(x = as.matrix(select_msp_data), y = select_data$group, type = "nonparametric", paired= F, multiple.adjust = "BH"))
}
```

```{r}
# check values mentionned in paper
nrow(testR_msp_GBA_PD_sPD[which(testR_msp_GBA_PD_sPD$p<0.05),]) # 44 MSP
nrow(testR_msp_GBA_PD_sPD[which(testR_msp_GBA_PD_sPD$q<0.6),]) # 0 MSP
nrow(testR_msp_HC_GBA_PD[which(testR_msp_HC_GBA_PD$q<0.1),]) # 82 MSP
nrow(testR_msp_HC_sPD[which(testR_msp_HC_sPD$q<0.1),]) # 123 MSP

# number of MSP significant according to p<0.05 and q<0.05
nrow(testR_msp_HC_GBA_PD[which(testR_msp_HC_GBA_PD$p<0.05),]) # 142 MSP
nrow(testR_msp_HC_GBA_PD[which(testR_msp_HC_GBA_PD$q<0.05),]) # 50 MSP

nrow(testR_msp_HC_sPD[which(testR_msp_HC_sPD$p<0.05),]) # 168 MSP
nrow(testR_msp_HC_sPD[which(testR_msp_HC_sPD$q<0.05),]) # 83 MSP

nrow(testR_msp_GBA_PD_sPD[which(testR_msp_GBA_PD_sPD$p<0.05),]) # 44 MSP 
nrow(testR_msp_GBA_PD_sPD[which(testR_msp_GBA_PD_sPD$q<0.05),]) # 0 MSP
```



```{r, results='hide', message=FALSE}
# MSP analysis by group_bis
vec_testR_name <- c()

# loop to perform pairwise comparison using testRelation2 
my_compa <- combn(x=levels(temp_metadata$group_bis), m = 2, simplify = T)
for (i in 1:ncol(my_compa)){
    # select individuals belong to select groups
    select_data <- temp_metadata %>% 
      filter(group_bis %in% c(my_compa[1,i], my_compa[2,i])) %>% 
      mutate(group_bis = factor(group_bis, levels = c(my_compa[1,i], my_compa[2,i])))
    
    # extract MSP abundance for selected individuals 
    select_msp_data <- msp_fecal.filt10[,rownames(select_data)]
    
    testR_name = paste("testR_msp", my_compa[1,i], my_compa[2,i], sep = '_')
    vec_testR_name <- append(vec_testR_name, testR_name)
    
    # perform pairwise comparisons and assign output to a new variable name
    assign(testR_name, testRelations2(x = as.matrix(select_msp_data), y = select_data$group_bis, type = "nonparametric", paired= F, multiple.adjust = "BH"))
}
```  


```{r, fig.width=11, fig.height=23}
# BARPLOT MSP CONTRAST - PAIRWISE COMPARISONS
tab_group <- table(temp_metadata$group_bis)

for (i in 1:ncol(my_compa)){
    pairwise_group <- paste(my_compa[1,i], my_compa[2,i], sep = "_") 

    df_testR <- get(vec_testR_name[i]) 
    df_testR$msp <- paste(rownames(df_testR), MSP_set[match(rownames(df_testR), rownames(MSP_set)), "species"], sep = "_")
    # extract output for significant test based on p<0.05
    df_testR.sign <- df_testR[which(df_testR$p<0.05),]
    
    # show number of significant MSP according to different pval
    cat(pairwise_group, ", nb msp sign p<0.05 : ", nrow(df_testR.sign),"\n") 
    cat(pairwise_group, ", nb msp sign q<0.05 : ", nrow(df_testR[which(df_testR$q<0.05),]), "\n")
    cat(pairwise_group, ", nb msp sign q<0.1 : ", nrow(df_testR[which(df_testR$q<0.1),]), "\n")
    
    assign(paste("testR_msp.sign", pairwise_group, sep = "_"), df_testR.sign)  
} 

# get the significant MSP common in both comparison 
common21 <- intersect(rownames(testR_msp.sign_HC_PD), rownames(testR_msp.sign_HC_GBA_NMC))
common_msp21 <- paste(common21, MSP_set[match(common21, rownames(MSP_set)), "species"], sep = "_")
``` 



```{r, fig.width=8}
# merge HC vs GBA_NMC and HC vs PD comparisons results
HC_vs_PD_NMC <- merge(testR_msp_HC_GBA_NMC, testR_msp_HC_PD, by = "row.names")
HC_vs_PD_NMC <- HC_vs_PD_NMC %>% tibble::column_to_rownames(var = "Row.names")
HC_vs_PD_NMC <- HC_vs_PD_NMC %>% select(p.x, HC.GBA_NMC_Cliffdelta, HC.GBA_NMC_status_Cliffdeltacorrected, p.y, HC.PD_Cliffdelta, HC.PD_status_Cliffdeltacorrected)

# add species information
HC_vs_PD_NMC$species <- MSP_set[match(rownames(HC_vs_PD_NMC), MSP_set$msp_id), "species"] 

# MSP enriched in HC relative to GBA_NMC had negative cliffdelta
all(HC_vs_PD_NMC[which(HC_vs_PD_NMC$HC.GBA_NMC_status_Cliffdeltacorrected == "HC"), "HC.GBA_NMC_Cliffdelta"]<0)
# MSP enriched in HC relative to PD had positive cliffdelta
all(HC_vs_PD_NMC[which(HC_vs_PD_NMC$HC.PD_status_Cliffdeltacorrected == "HC"), "HC.PD_Cliffdelta"]>0)

# multiply by minus 1 in order to have same direction of cliffdelta
HC_vs_PD_NMC$HC.PD_Cliffdelta <- HC_vs_PD_NMC$HC.PD_Cliffdelta*(-1)

# get significant MSP based on HC-PD comparison
HC_vs_PD_NMC.sign <- HC_vs_PD_NMC %>% filter(p.y<0.05)

# class the 176 MSP into CE, CD, NE and ND subtype specie groups 
HC_vs_PD_NMC.sign <- HC_vs_PD_NMC.sign %>% 
    mutate(type = case_when(HC.GBA_NMC_status_Cliffdeltacorrected == "GBA_NMC" & HC.PD_status_Cliffdeltacorrected =="PD" ~ "CE",
                            HC.GBA_NMC_status_Cliffdeltacorrected == "HC" & HC.PD_status_Cliffdeltacorrected == "HC" ~ "CD",
                            HC.GBA_NMC_status_Cliffdeltacorrected == "HC" & HC.PD_status_Cliffdeltacorrected == "PD" ~ "NE",
                            HC.GBA_NMC_status_Cliffdeltacorrected == "GBA_NMC" & HC.PD_status_Cliffdeltacorrected == "HC" ~ "ND")) 

# extract of MSP name for each subtype species
CE <- HC_vs_PD_NMC.sign %>% filter(type == "CE") %>% rownames()
CD <- HC_vs_PD_NMC.sign %>% filter(type == "CD") %>% rownames()
NE <- HC_vs_PD_NMC.sign %>% filter(type == "NE") %>% rownames() 
ND <- HC_vs_PD_NMC.sign %>% filter(type == "ND") %>% rownames() 

# MSP cliffdelta that belongs to CD subgroup are negative
all(HC_vs_PD_NMC.sign %>% filter(type == "CD") %>% select(HC.GBA_NMC_Cliffdelta)<0)
all(HC_vs_PD_NMC.sign %>% filter(type == "CD") %>% select(HC.PD_Cliffdelta)<0)

all(HC_vs_PD_NMC.sign %>% filter(type == "ND") %>% select(HC.GBA_NMC_Cliffdelta)>0)
all(HC_vs_PD_NMC.sign %>% filter(type == "ND") %>% select(HC.PD_Cliffdelta)<0)


# check number in subtype MSP 
length(CE) # 81
length(CD) # 61
length(NE) # 18
length(ND) # 16
```


# Figure 1
```{r, fig.width=8, fig.show='hide', results='hide'}
# BARPLOT COHERENT
# there is already enriched group column (based on cliffdelta), we create a depleted group column
HC_vs_PD_NMC.sign <- HC_vs_PD_NMC.sign %>% 
  mutate(HC.GBA_NMC_status_Cliff_depleted = ifelse(HC.GBA_NMC_status_Cliffdeltacorrected == "HC", "GBA_NMC", "HC")) %>% 
  mutate(HC.PD_status_Cliff_depleted = ifelse(HC.PD_status_Cliffdeltacorrected == "HC", "PD", "HC")) 

# build table of pvalue 
df_cliff_CE <-  get_cliff_delta(HC_vs_PD_NMC.sign, cat = "CE") %>% mutate(type = "CE") 
df_cliff_CD <- get_cliff_delta(HC_vs_PD_NMC.sign, cat = "CD") %>% mutate(type = "CD")
df_cliff_coh <- rbind(df_cliff_CE, df_cliff_CD) 
df_cliff_coh <- df_cliff_coh %>% mutate(type = factor(type, levels = c("CE", "CD")))

# build table of mean cliffdelta of enriched/depleted MSP that are coherently altered in group GBA_NMC and PD
tab_mean_cliff_coh <- df_cliff_CE %>% group_by(cliff_status) %>% summarise_at(vars(cliff), list(mean)) %>% 
  rbind(df_cliff_CD %>% group_by(cliff_status) %>% summarise_at(vars(cliff), list(mean)))
tab_mean_cliff_coh$type <- c(rep("CE", 2), rep("CD", 2))  

# Student test of cliffdelta of coherent MSP between PD and GBA_NMC
coherent_cliff_ttest <- compute_cliff_ttest(df_cliff_CE) %>% 
  rbind(compute_cliff_ttest(df_cliff_CD))
coherent_cliff_ttest$y <- c("CE", "CD")


# Boxplot
bp_cliff142 <- ggplot(df_cliff_coh, aes(x=type, y=cliff, fill = cliff_status)) + 
  geom_boxplot() +
  theme_classic() +
  scale_fill_manual(values = color_GBA_NMC_PD) +
  labs(x = "", y = "CD") +
  scale_x_discrete(labels=c("CE" = "Enriched coherent species", "CD" = "Depleted coherent species")) + 
  annotate("text", x = coherent_cliff_ttest$y, y=0.36, label = paste("Student.p:", coherent_cliff_ttest$p), size = 7) + 
  theme(legend.position = "none", 
        axis.text = element_text(size = 19, color = "black"),
        axis.title = element_text(size = 20))  
```



```{r, fig.width=8, fig.show='hide', results='hide'}
# REGRESSION CLIFFDELTA COHERENT
HC_vs_PD_NMC_coh142 <- HC_vs_PD_NMC[which(rownames(HC_vs_PD_NMC) %in% c(CE, CD)),]

# identify the 20 significant and coherently MSP altered in both comparisons 
HC_vs_PD_NMC_coh142 <- HC_vs_PD_NMC_coh142 %>% 
  mutate(common_msp_sign = ifelse(p.x<0.05 & p.y<0.05, "yes", "no")) 

# extract the 20 significant and coherently altered MSP in both comparisons and 
# compute the lm equation of MSP cliffdeltas between both comparisons 
HC_vs_PD_NMC_coh20.sign <- HC_vs_PD_NMC_coh142 %>% filter(common_msp_sign == "yes")
lm_eq_coh20.sign <- show_lm_eq(HC_vs_PD_NMC_coh20.sign, x = "HC.GBA_NMC_Cliffdelta", y = "HC.PD_Cliffdelta") 

# regression of cliffdelta HC vs NMC and PD, point in red for significant MSP common for both comparisons
reg_cliff142 <- reg_plot(HC_vs_PD_NMC_coh142, x = "HC.GBA_NMC_Cliffdelta", y = "HC.PD_Cliffdelta", xlab = "CD - HC vs GBA_NMC", ylab = "CD - HC vs PD", color_var = "common_msp_sign", formula.pos = 0.98) + expand_limits(y = c(-0.3, 0.5)) +
  geom_text(x = -0.075, y = 0.41, label = lm_eq_coh20.sign, parse = T, size = 7,  color = "red", fontface = "plain") 
reg_cliff142
```

```{r, results='hide'}
# REGRESSION CLIFFDELTA NON COHERENT 
HC_vs_PD_NMC_noncoh <- HC_vs_PD_NMC[which(rownames(HC_vs_PD_NMC) %in% c(NE, ND)),]
HC_vs_PD_NMC[which(rownames(HC_vs_PD_NMC) %in% NE),]
# MSP enriched in HC relative to GBA_NMC had negative cliffdelta
all(HC_vs_PD_NMC_noncoh[which(HC_vs_PD_NMC_noncoh$HC.GBA_NMC_status_Cliffdeltacorrected=="HC"), "HC.GBA_NMC_Cliffdelta"]<0) 
# MSP enriched in HC relative to PD had negative cliffdelta
all(HC_vs_PD_NMC_noncoh[which(HC_vs_PD_NMC_noncoh$HC.PD_status_Cliffdeltacorrected=="HC"), "HC.PD_Cliffdelta"]<0)

# get the significant MSP
HC_vs_PD_NMC_noncoh.sign <- HC_vs_PD_NMC_noncoh %>% filter(p.x<0.05 & p.y<0.05) # 0 common significant MSP 
dim(HC_vs_PD_NMC_noncoh.sign) # 1
```


```{r, fig.show='hide'}
# REGRESSION CLIFFDELTA PD vs HC/GBA_NMC
# cliffdelta HC vs PD against cliffdelta NMC vs PD 
PD_vs_HC_NMC <- merge(testR_msp_HC_PD, testR_msp_GBA_NMC_PD, by= "row.names") 

PD_vs_HC_NMC <- PD_vs_HC_NMC %>% select(Row.names, p.x, HC.PD_Cliffdelta, HC.PD_status_Cliffdeltacorrected, p.y, GBA_NMC.PD_Cliffdelta, GBA_NMC.PD_status_Cliffdeltacorrected)

PD_vs_HC_NMC_noncoh34 <- PD_vs_HC_NMC %>% filter(Row.names %in% rownames(HC_vs_PD_NMC_noncoh))
PD_vs_HC_NMC_noncoh34

# regression cliffdelta PD/HC against PD/HC
reg_PD_vs_HC_NMC_cliff34 <- reg_plot(PD_vs_HC_NMC_noncoh34, x = "HC.PD_Cliffdelta", y = "GBA_NMC.PD_Cliffdelta", color = "black", xlab = "CD - HC vs PD", ylab = "CD - GBA_NMC vs PD", formula.pos = 0.98, point_shape = 17)
```



```{r, results='hide'}
# PROPORTION - COHERENT ALTERED IN PD (140 SPECIES)
vec_group <- c("HC", "GBA_NMC", "PD") 

# compute proportion of MSP coherently enriched or depleted in samples
grp_CE_prop <- compute_prop(vec_group, temp_metadata, msp_fecal.filt10, CE)
grp_CD_prop <- compute_prop(vec_group, temp_metadata, msp_fecal.filt10, CD)

# gather proportions of MSP coherently enriched or depleted into df
df_prop_coh <- reshape2::melt(grp_CE_prop) %>% mutate(cat = "PD_enriched") %>% 
  rbind(reshape2::melt(grp_CD_prop) %>% mutate(cat = "PD_depleted")) 
colnames(df_prop_coh) <- c("value", "variable", "cat")

df_prop_coh <- df_prop_coh %>% 
  mutate(variable = factor(variable, levels = c("HC", "GBA_NMC", "PD")), 
         cat = factor(cat, levels = c("PD_enriched", "PD_depleted"))) 

# boxplot
bp_prop_coh <- show_boxplot(df_prop_coh, x = "variable", y = "value", fill = "cat", my_color = c("PD_enriched" = PD_enriched_col, "PD_depleted" = PD_depleted_col), ylab = "MSP proportion (%)", title = "Proportion of coherently altered species", limits = c(0, round(max(df_prop_coh$value))), pas = 10, show_legend = "none")
bp_prop_coh

# tab of pvalues 
prop_coh_pval <- get_tab_pval_ttest(grp_CE_prop,  grp_CD_prop) 
fig_prop_coh_pval <- show_tab(prop_coh_pval)
prop_coh_pval
```


```{r, results='hide'}
# PROPORTION - NONCOHERENT ALTERED IN PD
vec_group <- c("HC", "GBA_NMC", "PD") 

# compute the proportion for each subtype in each group
grp_NE_prop <- compute_prop(vec_group, temp_metadata, msp_fecal.filt10, NE)
grp_ND_prop <- compute_prop(vec_group, temp_metadata, msp_fecal.filt10, ND)

# gather proportions of MSP noncoherently enriched or depleted into df
df_prop_noncoh <- reshape2::melt(grp_NE_prop) %>% mutate(cat = "PD_enriched") %>% 
  rbind(reshape2::melt(grp_ND_prop) %>% mutate(cat = "PD_depleted")) 
colnames(df_prop_noncoh) <- c("value", "variable", "cat")

df_prop_noncoh <- df_prop_noncoh %>% 
  mutate(variable = factor(variable, levels = c("HC", "GBA_NMC", "PD")), 
         cat = factor(cat, levels = c("PD_enriched", "PD_depleted"))) 

# boxplot
bp_prop_noncoh <- show_boxplot(df_prop_noncoh, x = "variable", y = "value", fill = "cat", c("PD_enriched" = PD_enriched_col, "PD_depleted" = PD_depleted_col), ylab = "MSP proportion (%)", title = "Proportion of non-coherently altered species", limits = c(0, round(max(df_prop_noncoh$value))), pas = 5, show_legend = "none")

# tab of pvalues
prop_noncoh_pval <- get_tab_pval_ttest(grp_NE_prop, grp_ND_prop) 
prop_noncoh_pval
fig_prop_noncoh_pval <- show_tab(prop_noncoh_pval)
```
 

```{r, results='hide'}
# ABUNDANCE - COHERENT ALTERED IN PD
vec_group <- c("HC", "GBA_NMC", "PD") 

grp_CE_abund <- compute_abund(vec_group, temp_metadata, msp_fecal.filt10, CE)
grp_CD_abund <- compute_abund(vec_group, temp_metadata, msp_fecal.filt10, CD)

# gather abundances of MSP coherently enriched or depleted into df
df_abund_coh <- reshape2::melt(grp_CE_abund) %>% mutate(cat = "PD_enriched") %>% 
  rbind(reshape2::melt(grp_CD_abund) %>% mutate(cat = "PD_depleted")) 
colnames(df_abund_coh) <- c("value", "variable", "cat")

df_abund_coh <- df_abund_coh %>% 
  mutate(variable = factor(variable, levels = c("HC", "GBA_NMC", "PD")),
         cat = factor(cat, levels = c("PD_enriched", "PD_depleted"))) 

# boxplot
bp_abund_coh <- show_boxplot(df_abund_coh, x = "variable", y = "value", fill = "cat", my_color = c("PD_enriched" = PD_enriched_col, "PD_depleted" = PD_depleted_col), ylab = "MSP abundance (%)", title = "Abundance of coherently altered species", limits = c(0, round(max(df_abund_coh$value))), pas = 10, show_legend = "none") 

# tab of pvalues 
abund_coh_pval <- get_tab_pval_ttest(grp_CE_abund,  grp_CD_abund) 
abund_coh_pval
fig_abund_coh_pval <- show_tab(abund_coh_pval)
```
 

```{r, warning = T, results='hide'}
# ABUNDANCE - NONCOHERENT ALTERED IN PD
# The data used are the abundance of MSP which are non-coherently altered in GBA_NMC and PD relative to HC. 
vec_group <- c("HC", "GBA_NMC", "PD") 

grp_NE_abund <- compute_abund(vec_group, temp_metadata, msp_fecal.filt10, NE)
grp_ND_abund <- compute_abund(vec_group, temp_metadata, msp_fecal.filt10, ND)

# gather abundances of MSP noncoherently enriched or depleted into df
df_abund_noncoh <- reshape2::melt(grp_NE_abund) %>% mutate(cat = "PD_enriched") %>% 
  rbind(reshape2::melt(grp_ND_abund) %>% mutate(cat = "PD_depleted")) 
colnames(df_abund_noncoh) <- c("value", "variable", "cat")

df_abund_noncoh <- df_abund_noncoh %>% 
  mutate(variable = factor(variable, levels = c("HC", "GBA_NMC", "PD")), 
         cat = factor(cat, levels = c("PD_enriched", "PD_depleted")))

# boxplot
bp_abund_noncoh <- show_boxplot(df_abund_noncoh, x = "variable", y = "value", fill = "cat", my_color = c("PD_enriched" = PD_enriched_col, "PD_depleted" = PD_depleted_col), ylab = "MSP abundance (%)", title = "Abundance of non-coherently altered species", limits = c(0, round(max(df_abund_noncoh$value))), pas = 10, show_legend = "none") 

# tab of pvalues
abund_noncoh_pval <- get_tab_pval_ttest(grp_NE_abund,  grp_ND_abund) 
abund_noncoh_pval
fig_abund_noncoh_pval <- show_tab(abund_noncoh_pval)
```


```{r, fig.height=14, fig.width=6}
fig1_left <- ggarrange(reg_cliff142, bp_cliff142, reg_PD_vs_HC_NMC_cliff34, ncol = 1, labels = c("A", "B", "C"), font.label=list(size=25))
``` 

```{r, fig.height=12, fig.width=16}
# versionf - with test 
bp_pval_prop_coh <- ggarrange(bp_prop_coh, fig_prop_coh_pval, nrow = 2, heights = c(2, 0.9)) 
bp_pval_abund_coh <- ggarrange(bp_abund_coh, fig_abund_coh_pval, nrow = 2, heights = c(2, 0.9))
bp_pval_prop_noncoh <- ggarrange(bp_prop_noncoh, fig_prop_noncoh_pval, nrow = 2, heights = c(2, 0.9)) 
bp_pval_abund_noncoh <- ggarrange(bp_abund_noncoh, fig_abund_noncoh_pval, nrow = 2, heights = c(2, 0.9))

# merged fig - proportion abundance 
fig1_right <- ggarrange(bp_pval_prop_coh, 
                        bp_pval_abund_coh, 
                        bp_pval_prop_noncoh, 
                        bp_pval_abund_noncoh, nrow = 2, ncol = 2, labels = c("D", "E", "F", "G"), font.label = list(size = 25))
``` 

```{r, fig.height=15, fig.width=23, fig.show='asis'}
final_fig1 <- ggarrange(fig1_left, "", fig1_right, ncol = 3, widths = c(1.4, 0.03, 2))
final_fig1
```


# Figure 2
```{r}
# create a dataframe for gather information of PDMS16 MSP
# df contains 2 columns : type (CE/NE or CD) and cat (PD_depleted/PD_enriched)
pdms16 <- c("msp_1302","msp_0060","msp_0639","msp_0250","msp_0480","msp_0665","msp_0541","msp_1581","msp_1558","msp_0044","msp_0274","msp_0545","msp_0086","msp_0379","msp_0510","msp_0524")
pdms16_data <- HC_vs_PD_NMC.sign[pdms16,] %>% select(type)
pdms16_data <- pdms16_data %>% mutate(cat = ifelse(rownames(pdms16_data) %in% c(CE, NE), "PD_enriched", "PD_depleted")) 
```

```{r}
# extract the interest MSP abundance in HC and PD
abund_pdms_HC <- msp_fecal.filt10[rownames(pdms16_data), HC_ind]
abund_pdms_PD <- msp_fecal.filt10[rownames(pdms16_data), PD_ind]

# divided dataframe of PDMS16 into enriched and depleted
msp_enriched <- pdms16_data %>% filter(cat == "PD_enriched")
msp_depleted <- pdms16_data %>% filter(cat == "PD_depleted")

# count the number of enriched and depleted PDMS16 MSP in HC and PD individuals  
HC_count_enriched <- colSums(abund_pdms_HC[rownames(msp_enriched),]>0)
HC_count_depleted <- colSums(abund_pdms_HC[rownames(msp_depleted),]>0)

PD_count_enriched <- colSums(abund_pdms_PD[rownames(msp_enriched),]>0)
PD_count_depleted <- colSums(abund_pdms_PD[rownames(msp_depleted),]>0)

# checking 
abund_pdms_HC %>% select(ASAPh_268) %>% filter(rownames(abund_pdms_HC) %in% c(CE, NE))
abund_pdms_HC %>% select(ASAPh_384) %>% filter(rownames(abund_pdms_HC) %in% c(CD, ND))

abund_pdms_PD %>% select(ASAPh_850) %>% filter(rownames(abund_pdms_PD) %in% c(CE, NE))
abund_pdms_PD %>% select(ASAPh_850) %>% filter(rownames(abund_pdms_PD) %in% c(CD, ND))

all(names(HC_count_enriched) == names(HC_count_depleted))
all(names(PD_count_enriched) == names(PD_count_depleted))

# get the difference between enriched and depleted PDMS16 species number in HC and PD
HC_pdms <- HC_count_enriched - HC_count_depleted
PD_pdms <- PD_count_enriched - PD_count_depleted

```


```{r}
# for checking purpose I add education_n_years 
var_tab <- c("pd_dur", "ledd_total", "bdi_total", "hads_depression_total", "hads_anxiety_total", "updrs_i_total", "updrs_ii_total", "updrs_iii_total", "updrs_iv_total", "updrs_total", "hoehn_yahr", "scopa_gastrointestinal", "scopa_pupillomotor", "scopa_cardiovascular", "scopa_thermoregulatory", "scopa_urinary", "scopa_sexual", "scopa_total", "rbdsq_total", "sleepiness", "wcs_total", "appendectomy_pd_yearsdiff", "age", "bmi", "upsit_total",  "moca_total", "DQS_score", "Fruit", "Vegetables", "education_n_years", "ed_yrs") 

temp_metadata_bis <- temp_metadata[, var_tab]
```





```{r}
# abundance coherent depleted (CD) for PD 
PD_quartile_CD_abund <- get_quartile(grp_CD_abund$PD, "D") 

# proportion of coherent enriched (CE) for NMC 
NMC_quartile_CE_prop <- get_median(grp_CE_prop$GBA_NMC, "E")

# proportion of coherent enriched (CE) for HC
HC_quartile_CE_prop <- get_quartile(grp_CE_prop$HC, "E")

# quartile of HC16 is determined by their PDMS score
HC16_lower <- HC_pdms[which(HC_pdms <=(-5))]
HC16_higher <- HC_pdms[which(HC_pdms >= 3)]

# checking
length(NMC_quartile_CE_prop$below)
length(NMC_quartile_CE_prop$above)
NMC_quartile_CD_prop <- get_median(grp_CD_prop$GBA_NMC, "D")

length(HC16_lower)
length(HC16_higher)
```



```{r}
# compute mean and standard deviation of variable of each quartile in each group 
PD_quartile_clinical <- compute_mean_sd(PD_quartile_CD_abund$Q1, PD_quartile_CD_abund$Q4, temp_metadata_bis, "PD")

NMC_quartile_clinical <- compute_mean_sd(NMC_quartile_CE_prop$below, NMC_quartile_CE_prop$above, temp_metadata_bis, "NMC")

HC_quartile_clinical <- compute_mean_sd(HC_quartile_CE_prop$Q1, HC_quartile_CE_prop$Q4, temp_metadata_bis, "HC")

HC16_quartile_clinical <- compute_mean_sd(HC16_lower, HC16_higher, temp_metadata_bis, "HC16")

# checking
HC16_quartile_clinical$mean$HC16_L[c("education_n_years", "ed_yrs")]
HC16_quartile_clinical$mean$HC16_M[c("education_n_years", "ed_yrs")]
```


```{r}
# construct dataframe contains the mean for selected variables in least and more altered group of HC/NMC/PD/HC16
df_quartile_mean <- data.frame(PD_quartile_clinical$mean, NMC_quartile_clinical$mean, HC_quartile_clinical$mean, HC16_quartile_clinical$mean) %>% t() %>% as.data.frame()

# construct df containing the sd for selected variables
df_quartile_sd <- data.frame(PD_quartile_clinical$sd, NMC_quartile_clinical$sd, HC_quartile_clinical$sd, HC16_quartile_clinical$sd) %>% t() %>% as.data.frame() 
colnames(df_quartile_sd) <- paste0(colnames(df_quartile_sd), "_sd")

# bind means and sd columns  
df_quartile_mean_sd <- df_quartile_mean %>% cbind(df_quartile_sd)

# add group information
df_quartile_mean_sd$group <- c(rep(c("PD", "NMC", "HC", "HC16"), each = 2))  
df_quartile_mean_sd$group <- factor(df_quartile_mean_sd$group, levels = c("PD", "NMC", "HC", "HC16"))
df_quartile_mean_sd$sub <- factor(rownames(df_quartile_mean_sd), levels = c(rownames(df_quartile_mean_sd)))
df_quartile_mean_sd
```





```{r}
# create dataframe containing the values of variables in each group and quartile
PD_quartile_clinical_data <- PD_quartile_clinical$data1 %>% mutate(quartile = "Q1") %>% 
  rbind(PD_quartile_clinical$data2 %>% mutate(quartile = "Q4")) 


NMC_quartile_clinical_data <-  NMC_quartile_clinical$data1 %>% mutate(quartile = "Q1") %>% 
  rbind(NMC_quartile_clinical$data2 %>% mutate(quartile = "Q4")) 


HC_quartile_clinical_data <- HC_quartile_clinical$data1 %>% mutate(quartile = "Q1") %>% 
  rbind(HC_quartile_clinical$data2 %>% mutate(quartile = "Q4"))

HC16_quartile_clinical_data <- HC16_quartile_clinical$data1 %>% mutate(quartile = "Q1") %>% 
  rbind(HC16_quartile_clinical$data2 %>% mutate(quartile = "Q4"))
```



```{r}
# compute pval on different clinical variables
PD_pval <- PD_quartile_clinical_data %>% compute_pval() %>% mutate(group = "PD")


NMC_pval <- NMC_quartile_clinical_data %>% select(!c(pd_dur, ledd_total, appendectomy_pd_yearsdiff)) %>% 
  compute_pval() %>% mutate(group = "NMC")

HC_pval <- HC_quartile_clinical_data %>% select(!c(pd_dur, ledd_total, appendectomy_pd_yearsdiff)) %>%
  compute_pval() %>% mutate(group = "HC")


HC16_pval <- HC16_quartile_clinical_data %>% select(!c(pd_dur, ledd_total, appendectomy_pd_yearsdiff)) %>% 
  compute_pval() %>% mutate(group = "HC16")

# regroup all pvalue into dataframe
df_pval <- PD_pval %>% rbind(NMC_pval) %>% rbind(HC_pval) %>% rbind(HC16_pval)
df_pval
```

```{r, results='hide'}
# checking - proportion CD in GBA_NMC
test_NMC_quartile_clinical <- compute_mean_sd(NMC_quartile_CD_prop$below, NMC_quartile_CD_prop$above, temp_metadata_bis, "NMC")
test_NMC_quartile_clinical$mean

test_NMC_quartile_clinical_data <- test_NMC_quartile_clinical$data1 %>% mutate(quartile = "Q1") %>%
  rbind(test_NMC_quartile_clinical$data2 %>% mutate(quartile = "Q4"))

test_NMC_pval <- test_NMC_quartile_clinical_data %>% select(!c(pd_dur, ledd_total, appendectomy_pd_yearsdiff)) %>% compute_pval() %>% mutate(group = "NMC")
test_NMC_pval

# checking - proportion CE in GBA_NMC
test_NMC_quartile_clinical <- compute_mean_sd(NMC_quartile_CE_prop$below, NMC_quartile_CE_prop$above, temp_metadata_bis, "NMC")
test_NMC_quartile_clinical$mean

test_NMC_quartile_clinical_data <- test_NMC_quartile_clinical$data1 %>% mutate(quartile = "Q1") %>%
  rbind(test_NMC_quartile_clinical$data2 %>% mutate(quartile = "Q4"))

test_NMC_pval <- test_NMC_quartile_clinical_data %>% select(!c(pd_dur, ledd_total, appendectomy_pd_yearsdiff)) %>% compute_pval() %>% mutate(group = "NMC")
test_NMC_pval
```



```{r}
# select final variable and construct the figure 
variable_tab_select <- c("pd_dur", "ledd_total", "scopa_total", "wcs_total", "bdi_total", "updrs_i_total", "DQS_score")
variable_tab_select_sd <- paste0(variable_tab_select, sep = "_sd")

df_quartile_mean_sd_sub <- df_quartile_mean_sd %>% select(all_of(c(variable_tab_select, variable_tab_select_sd, "group", "sub")))
```



```{r}
# extract the title for selected variables 
variable_tab_select_description <- variable_tab_select 
variable_tab_select <- setNames(c("PD_duration_year", "Levodopa_dose_mg", "SCOPA_total", "WCS_total", "BDI_score", "UPDRS_part_I", "DQS_score"), variable_tab_select) 

variable_tab_select_description <- setNames(c("Disease duration", "Medications", "Autonomic dysfunction (SCOPA-AUT)", "Constipation (WCSS)", "Depression (BDI)", "Non-motor symptoms (MDS-UPDRS part I)", "Diet quality"), variable_tab_select_description)
```


```{r}
# first construct the barplot of variable in common in all group
variable_common <- variable_tab_select[!names(variable_tab_select) %in% c("pd_dur", "ledd_total")]
variable_common_description <- variable_tab_select_description[!names(variable_tab_select_description) %in% c("pd_dur", "ledd_total")]
```



```{r, fig.width=10}
# BARPLOT FOR VARIABLE IN COMMON
group_temp_fill <- c(HC = "#009E73", NMC =  "#F0E442", PD ="#CC79A7", HC16 =  "#009E73")

# barplot of mean and sd as bars errors for variables that are common in group
bp_quartile_list <- list()
for (i in 1:length(variable_common)){
  var <- names(variable_common)[i]
  print(var)
  var_sd <- paste0(var, "_sd")
  data_p <- df_pval %>% filter(y == var)
  
  data_p$x_position <- c(1.5, 3.5, 5.5, 7.5) # Milieu entre les barres
  data_p$y_position <- tapply(df_quartile_mean_sd_sub[, var], df_quartile_mean_sd_sub$group, max) + 0.8 # Au-dessus des barplots
  
 p <- ggplot(df_quartile_mean_sd_sub, aes(x = sub, y = !!sym(var), fill = group)) +
    geom_bar(stat = "identity", width = 0.8) +  # Barre principale
    # geom_errorbar(aes(ymin = !!sym(var) - !!sym(var_sd), ymax = !!sym(var) + !!sym(var_sd)), width = 0.1) + # Barres d'erreur
    geom_errorbar(aes(ymin = !!sym(var), ymax = !!sym(var) + !!sym(var_sd)), width = 0.1) + # Barres d'erreur
    geom_text(data = data_p, aes(x = x_position, y = y_position, label = p.signif), size = 4) + # ajout des p-valeurs
    scale_fill_manual(values = group_temp_fill) + 
    labs(title = variable_common_description[[i]], x = "", y = variable_common[[i]]) + 
    theme_pubr() +
    theme(panel.background = element_blank(), 
          legend.position = "none", 
          plot.title = element_text(face = "bold", size = 14))
          # axis.title = element_text(size = 13, color = "black"),
          # axis.text = element_text(size = 12, color = "black"))
  bp_quartile_list[[var]] <- p
 print(p)
} 
```


```{r}
# appendectomy stacked barplot 
# bind the row from two dataframe _L and _M 
bind_quartile_df <- function(subset1, subset2, metadata, name_grp){
  # select individuals from the quartile (subset1) 
  subset1_data <- metadata %>%
    filter(rownames(metadata) %in% names(subset1)) %>% 
    mutate(sub = paste0(name_grp, "_L")) 
  # select individuals from the quartile subset2
  subset2_data <- metadata %>% 
    filter(rownames(metadata) %in% names(subset2)) %>% 
    mutate(sub = paste0(name_grp, "_M"))
  subset_data <- subset1_data %>% rbind(subset2_data)
  return(subset_data)
}

var_cat <- c("appendectomy", "sex", "pd_drug_naive")
temp_metadata_chr <- temp_metadata[, var_cat]
# temp_metadata %>% dplyr::select_if(is.character)

# for each group we extract categorical variables information (appendectomy, sex) in each quartile (_L and _M)
PD_quartile_categorical <- bind_quartile_df(PD_quartile_CD_abund$Q1, PD_quartile_CD_abund$Q4, temp_metadata_chr, "PD")
NMC_quartile_categorical <- bind_quartile_df(NMC_quartile_CE_prop$below, NMC_quartile_CE_prop$above, temp_metadata_chr, "NMC")
HC_quartile_categorical <- bind_quartile_df(HC_quartile_CE_prop$Q1, HC_quartile_CE_prop$Q4, temp_metadata_chr, "HC")
HC16_quartile_categorical <- bind_quartile_df(HC16_lower, HC16_higher, temp_metadata_chr, "HC16")

df_quartile_categorical <- rbind(PD_quartile_categorical, NMC_quartile_categorical, HC_quartile_categorical, HC16_quartile_categorical)
df_quartile_categorical <- df_quartile_categorical %>% mutate(group = str_remove(sub, pattern = "_L|_M")) 
df_quartile_categorical
```

```{r}
# chisq test - all categorical 
var_cat <- c("appendectomy", "sex")
group_chisq = c("PD", "NMC", "HC", "HC16") 
df_chisq <- data.frame()
for (i in 1:length(var_cat)){
  for (j in 1:length(group_chisq)){
    # select specific categorical variable and individual of group (PD, HC...)
    df_select <- df_quartile_categorical %>% 
        dplyr::select(var_cat[i], group, sub) %>% 
        filter(group == as.character(group_chisq[j])) 
    
    df_select$sub <- as.character(df_select$sub)
    
    # build contingency table by MSP subtype and categorical variable
    contingency_tab <- table(df_select[,c(var_cat[i], "sub")])
    
    # perform chi2 test and save result into df 
    chisq <- chisq_test(contingency_tab, correct = T) %>% 
      mutate(y = var_cat[i], 
             group = group_chisq[j])
    df_chisq <- df_chisq %>% rbind(chisq) 
  } 
}

# extract chisq test for appendectomy
df_chisq_appendectomy <- df_chisq %>% filter(y == "appendectomy")
df_chisq_appendectomy <- df_chisq_appendectomy %>% mutate(x_position = c(1.5, 3.5, 5.5, 7.5))
df_chisq_appendectomy <- df_chisq_appendectomy %>% mutate(y_position = rep(1.05))

# add 
# barplot + chisq test sign 
sub_order <- unique(df_quartile_categorical$sub)
df_quartile_categorical <- df_quartile_categorical %>% 
  mutate(sub = factor(sub, levels = sub_order)) 

bp_appendectomy <- ggplot(df_quartile_categorical, aes(x = sub, fill = appendectomy)) +
  geom_bar(stat = "count", width = 0.8, position = "fill") + 
  labs(title = "Appendectomy", x = "", y = "appendectomy proportion") + 
  # geom_text(aes(df_chisq_appendectomy$p.signif)) + 
  annotate("text", x = df_chisq_appendectomy$x_position, y = df_chisq_appendectomy$y_position, label = df_chisq_appendectomy$p.signif, size = 4) + # ajout des p-valeurs
  scale_fill_manual(values = c("yes" = "#000000", "no" = "#999999")) + 
  theme_pubr() + 
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14)) 

bp_appendectomy
```

```{r}
# variable available only for PD 
variable_tab_PD <- variable_tab_select[names(variable_tab_select) %in% c("pd_dur", "ledd_total")]
variable_tab_PD_title <- variable_tab_select_description[names(variable_tab_select_description) %in% c("pd_dur", "ledd_total")]
```

```{r}
# barplot of mean and sd as error bar for PD 
PD_df_quartile <- df_quartile_mean_sd %>% filter(str_detect(rownames(df_quartile_mean_sd), ".*PD.*"))
PD_df_quartile <- PD_df_quartile %>% mutate(sub = factor(sub, levels = rownames(PD_df_quartile)))
# PD_pval[which(PD_pval$p.signif == "ns"), "p.signif"] <- ""

bp_PD_quartile_list <- list()
# variable_PD <- list("PD_duration_year" = "Disease duration", "Levodopa_dose_mg" = "Medications")
for (i in 1:length(variable_tab_PD)){
  var <- names(variable_tab_PD)[i]
  var_sd <- paste0(var, "_sd")
  data_p <- df_pval %>% 
    filter(group == "PD" & y == var)

  data_p$y_position <- max(PD_df_quartile[, var]) + 1 # Au-dessus des barplots
  data_p$x_position <- c(1.5) # Milieu entre les barres

  
 p <- ggplot(PD_df_quartile, aes(x = sub, y = !!sym(var), fill = group)) +
    geom_bar(stat = "identity", width = 0.8) +  # Barre principale
    geom_errorbar(aes(ymin = !!sym(var), ymax = !!sym(var) + !!sym(var_sd)), width = 0.1) + # Barres d'erreur
    geom_text(data = data_p, aes(x = x_position, y = y_position, label = p.signif), size = 4) + # ajout des p-valeurs
    scale_fill_manual(values = group_temp_fill) + 
    labs(title = variable_tab_PD_title[[i]], x = "", y = variable_tab_PD[[i]]) + 
    theme_pubr() +
    theme(panel.background = element_blank(), 
          legend.position = "none", 
          plot.title = element_text(face = "bold", size = 14))
          # axis.title = element_text(size = 13, color = "black"),
          # axis.text = element_text(size = 12, color = "black"))
  bp_PD_quartile_list[[var]] <- p
 print(p)
}
```


```{r, fig.height=4, fig.width=9}
fig_tab_top <- ggarrange(bp_PD_quartile_list$pd_dur, 
                         bp_PD_quartile_list$ledd_total, 
                         nrow = 1, labels = c("A"), font.label = list(size = 17))
fig_tab_top
```

```{r, fig.height=8, fig.width=15}
fig_tab_mid <- ggarrange(bp_quartile_list$scopa_total, 
                         bp_quartile_list$wcs_total, 
                         bp_quartile_list$bdi_total, 
                         bp_quartile_list$updrs_i_total, 
                         nrow = 2, ncol = 2, labels = c("B"),  font.label = list(size = 17, color = "black"))
fig_tab_mid
```


```{r, fig.height=4, fig.width=14}
fig_tab_bottom <- ggarrange(bp_quartile_list$DQS_score, 
                            bp_appendectomy, 
                            nrow = 1, labels = c("C"), font.label = list(size = 17))

fig_tab_bottom
```



```{r, fig.height=12, fig.width=14, fig.show='asis'}
fig_tab <- ggarrange(fig_tab_top, fig_tab_mid, fig_tab_bottom, ncol = 1, heights = c(0.8, 1.2, 0.8))
fig_tab
```


# Figure 3 

```{r, fig.height=5, fig.width=7, fig.show='asis'}
NMC_data <- msp_fecal.filt10[, NMC_ind]
NMC_enriched_data <- msp_fecal.filt10[CE, NMC_ind]
 
sum_abund_NMC_enriched <- apply(NMC_enriched_data, 2, sum)
sum_abund_NMC <- apply(NMC_data, 2, sum)

abund_rel <- sum_abund_NMC_enriched/sum_abund_NMC*100
abund_rel_rank <- sort(abund_rel)

df_enriched_NMC <- data.frame(abund_rel_rank)
df_enriched_NMC$prob_prodromalPD <- temp_metadata[match(rownames(df_enriched_NMC), rownames(temp_metadata)), "prob_prodromalPD"] 
df_enriched_NMC <- df_enriched_NMC %>% mutate(prob_prodromalPD = factor(prob_prodromalPD, levels = c("yes", "no"))) 

# final 
median_CE_NMC <- median(x=seq(1:nrow(df_enriched_NMC))) 
fig_dot_NMC_rank <- ggplot(df_enriched_NMC, aes(x=seq(1:nrow(df_enriched_NMC)), y = abund_rel_rank)) + 
  geom_point(color = "#353868", aes(shape=prob_prodromalPD), size=3) + 
  scale_shape_manual(values = c("yes" = 16, "no" = 1), name = "", labels = c("prodromal", "not prodromal")) +
  ylab("Abundance of PD enriched species (%)") +
  xlab("Rank by abundance of PD enriched species") +
  expand_limits(y = c(0, 30), x = c(1,50)) + 
  scale_y_continuous(breaks = seq(0, 30, 5)) +
  geom_segment(aes(x = median_CE_NMC, y = round(df_enriched_NMC[median_CE_NMC, "abund_rel_rank"])+6.5, xend = median_CE_NMC, yend = round(df_enriched_NMC[median_CE_NMC, "abund_rel_rank"])), linewidth = 0.2, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom ="text", x = median_CE_NMC, y = 15, label = "median", angle=90, size=5) + 
  theme_classic() + 
  theme(axis.title = element_text(size = 14), axis.text = element_text(size = 14), legend.text = element_text(size = 13))
fig_dot_NMC_rank

```

# Figure 4

First of all, we need to create the correlation table between the msp and the clinical variable we identify as being of interest, as well as the functional potential.

## MSP

This function creates a correlation table between each variable (here msp) and the clinical variable for each of the groups. We use a function that loops over each clinical variable in each group and calculates the Spearman correlation with the 176 msp isolated during (with an identified statistical difference) the PD and HC comparisons. Each correlation table for each variable is then merged into a single database containing all the correlations.
```{r}
result_cor <- list()
result_cor <- cor_176_clinic(result_cor, # empty list where all the correlation df would be stored
                             unique(temp_metadata$group_bis), ## vector of differents groups
                             msp_fecal.filt10 %>% 
                                t() %>%
                                as.data.frame() %>%
                                select(c(NE, ND, CE, CD)) %>%
                                merge(temp_metadata["group_bis"], by = "row.names") %>%
                                column_to_rownames(var = "Row.names") %>%
                                rename(group = group_bis), ## abundance table with group information
                             temp_metadata, ## metadatada
                             var_tab[!var_tab %in% c("Fruit", "Vegetables", "sleepiness",           "appendectomy","appendectomy_pd_yearsdiff","sex","education_n_years","ed_yrs" )], ## vector of variable of intesrest
                             name_var = "msp") ## name of the variable analysed
```

Now we want to add the information of the subgroup of each of msp. The output will be a dataframe with the msp and and their subgroup in columns. 
```{r}
df_subgroup <- data.frame(msp = character(), subgroup = character(), stringsAsFactors = FALSE)

## store each of the species names in each corresponding subgroup
msp_list <- list(CE = CE, CD = CD, ND = ND, NE= NE)

for (subgroup in names(msp_list)) {
  msps <- msp_list[[subgroup]]
  
  ### store the msps name for each corresponding subgroup in a dataframe
  df_subgroup <- rbind(df_subgroup, data.frame(msp = msps, subgroup = subgroup))
}
```

Add subgroup information for each msp to the correlation matrix for each group.
```{r}
## Looping through each correlation file (separate by group)
for (name_df in c("correlation_HC", "correlation_GBA_NMC", "correlation_PD")){
  
  ### loading correlation dataframe by group
  df <- result_cor[[name_df]]

  ### adding the information of subgroup for each msp to the correlation file
  df <- df %>%
          left_join(df_subgroup, by = "msp") %>%
          rename(msp_id = msp) %>%
          left_join(MSP_set, by = "msp_id") %>%
          mutate(msp_name = paste(msp_id, species, sep="_")) %>%
          select(msp_id, msp_name, clinical_variable, correlation, pvalue, qvalue, ref_MSP, subgroup)
  
  ### store the new correlation file with subgroup information in the initial list    
  result_cor[[name_df]] <- df
  
}
```

We will now combine all the correlation tables for each group into a single table.
```{r}
final_df <- as.data.frame(matrix(ncol = 9))
for (df_name in c("correlation_HC", "correlation_GBA_NMC", "correlation_PD")){
  df <- result_cor[[df_name]]
  df$group <- gsub("correlation_", "", df_name)
  if (df_name == "correlation_HC"){
    colnames(final_df) <- colnames(df)
  }
  final_df <- rbind(final_df, df)
}
```

```{r}
final_df <- final_df[-1,]
```

## Functional modules

The next step is to correlate the clinical variables of interest with each functional module calculated from the taxonomy table. 
```{r}
## creation of the empty list where all correlation table for each group would be stored.
module_result_cor <- list()

## creation of a new column which combining the information of module name and its ID.
all_modules_definition_GMM_GBM_KEGG_107$id_name <- paste(all_modules_definition_GMM_GBM_KEGG_107$module, all_modules_definition_GMM_GBM_KEGG_107$Module_pathway, sep= "_")
```

The analysis is carried out by treating each family of modules separately. 
```{r}
## Separation of module IDs according to their module family: KEGG, GMM, GBM.
kegg <- unique(all_modules_definition_GMM_GBM_KEGG_107[all_modules_definition_GMM_GBM_KEGG_107$DataBase == 'KEGG', "module"])
gmm <- unique(all_modules_definition_GMM_GBM_KEGG_107[all_modules_definition_GMM_GBM_KEGG_107$DataBase == 'GMM', "module"])
gbm <- unique(all_modules_definition_GMM_GBM_KEGG_107[all_modules_definition_GMM_GBM_KEGG_107$DataBase == 'GBM', "module"])
```

```{r}
## This loop iterates through the names we want to assign to each module table.
for (nm in c("ftmt_kegg", "ftmt_gmm", "ftmt_gbm")){
  
  ### get all the ID for the module family we iterate on
  data_mod <- get(gsub("ftmt_", "", nm))
  
  ### for each module family, filter the functional module table to keep only its rows, and name the result after the module family.
  assign(nm, ftmt.nocore.fecal.baseline[rownames(ftmt.nocore.fecal.baseline) %in% data_mod,])
}
```


As with the MSP, the functional modules will be divided into subgroups corresponding to their difference in ‘abundance’ in the HC vs PD and HC vs GBA_NMC comparison. The same subgroups will be found, coherent and non-coherent, enriched and depleted. Two other subgroups were created, corresponding to modules enriched and depleted in PD in the comparison with HC, but which did not reach a p value < 0.05.

```{r}
## Creation of a data frame to store all the subgroup information for each module ID..
module_subgroup <- as.data.frame(matrix(ncol = 1))
colnames(module_subgroup) <- "subgroup"

## This loop iterate on all functional module family table
for (mod_df in c("ftmt_gbm", "ftmt_gmm", "ftmt_kegg")) {
  
  data <- get(mod_df)
  
  ### Save the output of testRelation2 in a data frame. It include, among other, the pvalue and qvalue for each module ID from a Wilcoxon test comparing the two selected groups. First *HC vs PD*
  testrel_mod_HC_PD <- testRelations2(x = data %>%
                                            t() %>%
                                            merge(temp_metadata["group_bis"], by = "row.names") %>%
                                            column_to_rownames(var = "Row.names") %>%
                                            filter(group_bis %in% c("HC", "PD")) %>%
                                            select(-group_bis) %>%
                                            t(), # x correspond to the abundance table dataframe
                                      y = data %>%
                                            t() %>%
                                            merge(temp_metadata["group_bis"], by = "row.names") %>%
                                            column_to_rownames(var = "Row.names") %>%
                                            filter(group_bis %in% c("HC", "PD")) %>%
                                            mutate(group_bis = factor(group_bis)) %>%
                                            pull(group_bis), # y to the vector of the group
                                            type = "nonparametric", # the test we want to perform
                                            multiple.adjust = "BH", # the correction type
                                            paired = FALSE) # information about the type of comparison
  

  ### Then *HC vs GBA_NMC*
  testrel_mod_HC_GBA_NMC <- testRelations2(x = data %>%
                                            t() %>%
                                            merge(temp_metadata["group_bis"], by = "row.names") %>%
                                            column_to_rownames(var = "Row.names") %>%
                                            filter(group_bis %in% c("HC", "GBA_NMC")) %>%
                                            select(-group_bis) %>%
                                            t(),
                                           y = data %>%
                                            t() %>%
                                            merge(temp_metadata["group_bis"], by = "row.names") %>%
                                            column_to_rownames(var = "Row.names") %>%
                                            filter(group_bis %in% c("HC", "GBA_NMC")) %>%
                                            mutate(group_bis = factor(group_bis, level = c("GBA_NMC", "HC"))) %>%
                                            pull(group_bis),
                                            type = "nonparametric", 
                                            multiple.adjust = "BH", 
                                            paired = FALSE)
  
  ### creation of the variable name were *coherent enriched* functional module would be store. 
  CoherentE <- paste("Coh_E_mod", gsub("ftmt_", "", mod_df), sep = "_")
  
  assign(CoherentE, 
         testrel_mod_HC_PD %>%
         filter(p<0.05) %>% # retaining only functional modules where wilcoxon test pvalue < 0.05.
           
         filter(HC.PD_status_Cliffdeltacorrected == "PD") %>% # keeping only those functional modules whose abundance is higher in PD than in HC.
           
         filter(rownames(.) %in% rownames(testrel_mod_HC_GBA_NMC[testrel_mod_HC_GBA_NMC$GBA_NMC.HC_status_Cliffdeltacorrected == "GBA_NMC",])) %>% 
# only keeping the functional module enriched in PD if they are also enriched in GBA_NMC in the comparison HC vs GBA_NMC
           
          merge(testrel_mod_HC_GBA_NMC["GBA_NMC.HC_status_Cliffdeltacorrected"], by = "row.names") %>%
          column_to_rownames(var = "Row.names"))
  
  
  ### creation of the variable name were *coherent enriched X* functional module would be store. 
  XE <- paste("X_E_mod", gsub("ftmt_", "", mod_df), sep = "_")
  
  assign(XE, 
         testrel_mod_HC_PD %>%
         filter(p>0.05) %>%
         filter(HC.PD_status_Cliffdeltacorrected == "PD")) # XE represents all the functional modules enriched in the PD group compare to HC with a pvalue > 0.05 in the Wilcoxon test.
  
  
  ### creation of the variable name were *coherent depleted* functional module would be store.
  CoherentD <- paste("Coh_D_mod", gsub("ftmt_", "", mod_df), sep = "_")
  
  assign(CoherentD, 
         testrel_mod_HC_PD %>%
          filter(p<0.05) %>% # retaining only functional modules where wilcoxon test pvalue < 0.05.
          filter(HC.PD_status_Cliffdeltacorrected == "HC") %>% # keeping only those functional modules whose abundance is higher in HC than in PD.
           
          filter(rownames(.) %in% rownames(testrel_mod_HC_GBA_NMC[testrel_mod_HC_GBA_NMC$GBA_NMC.HC_status_Cliffdeltacorrected == "HC",])) %>%
          merge(testrel_mod_HC_GBA_NMC["GBA_NMC.HC_status_Cliffdeltacorrected"], by = "row.names") %>%
# only keeping the functional module enriched in HC if they are also enriched in HC in the comparison HC vs GBA_NMC
  
          column_to_rownames(var = "Row.names"))
  
  
  ### creation of the variable name were *coherent depleted X* functional module would be store. 
  XD <- paste("X_D_mod", gsub("ftmt_", "", mod_df), sep = "_")
  
  assign(XD, 
         testrel_mod_HC_PD %>%
         filter(p>0.05) %>%
         filter(HC.PD_status_Cliffdeltacorrected == "HC")) # XE represents all the functional modules enriched in the HC group compare to PD with a pvalue > 0.05 in the Wilcoxon test.
  
  
  ### creation of the variable name were *non-coherent enriched* functional module would be store.
  NoncoherentE <- paste("NCoh_E_mod", gsub("ftmt_", "", mod_df), sep = "_")
  
  assign(NoncoherentE, 
         testrel_mod_HC_PD %>%
          filter(HC.PD_status_Cliffdeltacorrected == "PD") %>% # only keeping the functional module enriched in PD with its comparison with HC, no matter the pvalue wilcoxon test value.
           
          filter(rownames(.) %in% rownames(testrel_mod_HC_GBA_NMC[testrel_mod_HC_GBA_NMC$GBA_NMC.HC_status_Cliffdeltacorrected == "HC",])) %>% # only keepin the functional modules enriched in HC when compared to GBA_NMC. This is considered inconsistent, as we expect the GBA_NMC group to represent an intermediate state between HC and PD. What is enriched in PD should also be enriched in GBA_NMC.
           
          merge(testrel_mod_HC_GBA_NMC["GBA_NMC.HC_status_Cliffdeltacorrected"], by = "row.names") %>%
          column_to_rownames(var = "Row.names"))
  
  
  ### creation of the variable name were *non-coherent depleted* functional module would be store.
  NoncoherentD <- paste("NCoh_D_mod", gsub("ftmt_", "", mod_df), sep = "_")
  
  assign(NoncoherentD, 
         testrel_mod_HC_PD %>%
          filter(HC.PD_status_Cliffdeltacorrected == "HC") %>% # only keeping the functional module enriched in HC with its comparison with PD, no matter the pvalue wilcoxon test value.
           
          filter(rownames(.) %in% rownames(testrel_mod_HC_GBA_NMC[testrel_mod_HC_GBA_NMC$GBA_NMC.HC_status_Cliffdeltacorrected == "GBA_NMC",])) %>% # only keepin the functional modules enriched in GBA_NMC when compared to HC.
           
          merge(testrel_mod_HC_GBA_NMC["GBA_NMC.HC_status_Cliffdeltacorrected"], by = "row.names") %>%
          column_to_rownames(var = "Row.names"))
  
  ### Each subgroup dataframe contains information about which module belongs to which subgroup, stored in the row names. A new column called 'subgroup' is added to the data, assigning a subgroup based on the module ID (from the rownames). For example, if those rownames are found in the rownames of the filtered matrix CoherentE. These functional modules are then labeled as 'CEM' in the 'subgroup' column. 
 
  data$subgroup <-  case_when(
  rownames(data) %in% rownames(get(CoherentE))  ~ "CEM",
  rownames(data) %in% rownames(get(XE)) ~ "XEM",
  rownames(data) %in% rownames(get(CoherentD))  ~ "CDM",
  rownames(data) %in% rownames(get(XD)) ~ "XDM",
  rownames(data) %in% rownames(get(NoncoherentE))  ~ "NEM",
  rownames(data) %in% rownames(get(NoncoherentD))  ~ "NDM",
  TRUE ~ "NA")
  
  data <- data["subgroup"]
  
  module_subgroup <- rbind(module_subgroup, data)
  
  names_testrel_HC_PD <- paste(gsub("ftmt_", "", all_modules_definition_GMM_GBM_KEGG_107), "testrel_HC_PD", sep = "_")
  names_testrel_HC_GBA_NMC <- paste(gsub("ftmt_", "", all_modules_definition_GMM_GBM_KEGG_107), "testrel_HC_GBA_NMC", sep = "_")
  
  assign(names_testrel_HC_PD, testrel_mod_HC_PD)
  assign(names_testrel_HC_GBA_NMC, testrel_mod_HC_GBA_NMC)
}

## add the column corresponding the the id of module with their name, and set it as rownames. 
module_subgroup <- module_subgroup %>%
                      rownames_to_column(var = "module") %>%
                      .[-1,] %>%
                      left_join(all_modules_definition_GMM_GBM_KEGG_107[, c("id_name", "module")], by = "module", copy = TRUE) %>%
                      column_to_rownames(var = "id_name")
  
```

Each module family dataframe undergoes a transformation before being input into the cor_176_clinic function, which calculates spearman correlations between each module and each clinical variable within each group.
```{r}
for (mod_df in c("ftmt_gbm", "ftmt_gmm", "ftmt_kegg")) {
  name_df <- paste(mod_df, "df", sep = "_")
  module_result_cor[[name_df]] <- list()
}
## The loop iterate on the abundance table of each module family. 
for (mod_df in c("ftmt_gbm", "ftmt_gmm", "ftmt_kegg")) {
  
  data <- get(mod_df)
  
  ### ensuring all modules from the dataframe are in the dataframe module 'dictionary'
  if (all(rownames(data) %in%  all_modules_definition_GMM_GBM_KEGG_107$module)) {
    
    ### attributing for each module ID the equivalence with name and ID
    replacement <- setNames(all_modules_definition_GMM_GBM_KEGG_107$id_name, 
                            all_modules_definition_GMM_GBM_KEGG_107$module)
    
    ### keeping only the module present in data 
    new_rownames <- replacement[rownames(data)]
    
    ### replacing the rownames of data by equivalent ID + name information
    rownames(data) <- new_rownames 
  }
  
  mod_data_metadata <- data %>% 
    t() %>%
    as.data.frame() %>%
    merge(temp_metadata["group_bis"], by = "row.names") %>%
    rename(group = group_bis) %>%
    column_to_rownames(., var = "Row.names")
  
  groups <- unique(mod_data_metadata$group)
  
  name_df <- paste(mod_df, "df", sep = "_")
  
  ### using the function cor_176_clinic to calculate correlation and store it in a variable
  list_df <- cor_176_clinic(module_result_cor[[name_df]], 
                            groups, 
                            mod_data_metadata, 
                            temp_metadata, 
                            var_tab[!var_tab %in% c("Fruit", "Vegetables", "sleepiness",           "appendectomy","appendectomy_pd_yearsdiff","sex","education_n_years","ed_yrs" )], 
                            "modules")
  
  ### iterate through each group correlation dataframe 
  for (df in seq_along(list_df)) {
    
    #### add subgroup information for each module to each of these dataframes
    list_df[[df]]$subgroup <- module_subgroup$subgroup[match(list_df[[df]]$modules, rownames(module_subgroup))]
  }
  
  ### store the correlation dataframe in the module_result_cor list
  module_result_cor[[name_df]] <- list_df
}
```

Merge all correlation dataframe to have all information in only one dataframe. 
```{r}
final_df_mod <- as.data.frame(matrix(ncol = 7))
for (mod_type in c("ftmt_gmm_df", "ftmt_gbm_df", "ftmt_kegg_df")){
  for (df_name in c("correlation_HC", "correlation_GBA_NMC", "correlation_PD")){
    df <- module_result_cor[[mod_type]][[df_name]]
    df$group <- gsub("correlation_", "", df_name)
    if (df_name == "correlation_HC"){
      colnames(final_df_mod) <- colnames(df)
    }
    final_df_mod <- rbind(final_df_mod, df)
  }
}
```

```{r}
final_df_mod <- final_df_mod[-1,]
```

Now that the dataframes have been generated, the figure can be produced. Firstly, the names of the clinical variables are changed to make them more meaningful. 
```{r}
name_correspondance <- read.csv("../data/clinical_name_corr.csv", sep = ";")
```

```{r}
## Clinical variable are divided into clinical variable link with health and the ones link with disease. 
list_clinical_variable_before <- list(
  c("moca_total", "upsit_total", "DQS_score"),
  c("updrs_i_total", "updrs_ii_total", "updrs_iii_total", "updrs_iv_total", 
    "updrs_total", "hoehn_yahr", "scopa_cardiovascular", "scopa_gastrointestinal", 
    "scopa_pupillomotor", "scopa_sexual", "scopa_thermoregulatory", "scopa_urinary", 
    "scopa_total", "rbdsq_total", "bdi_total", "hads_anxiety_total", 
    "hads_depression_total", "wcs_total", "ledd_total", "pd_dur")
)

### Create a list where the clinical variable new name will be store
list_clinical_variable <- list()

for (i in seq_along(list_clinical_variable_before)) {  
  list_clinical_variable[[i]] <- sapply(list_clinical_variable_before[[i]], function(nm) {
    if (nm %in% name_correspondance$clinical_variable) {
      name_correspondance[name_correspondance$clinical_variable == nm, "clinical_variable_bis"]
    } else {
      nm  
    }
  })
}
```

```{r}
## Separate the final_df (msp) by group
PD <- final_df %>%
        filter(group == "PD") 
HC <- final_df %>%
        filter(group == "HC")
GBA_NMC <- final_df %>%
              filter(group == "GBA_NMC")
```

Use of the prep_data_fig4 function, which generates a dataframe in a format that is easier to use for creating the figure. 
```{r}
df_reshaped_sp <- prep_data_fig4(
  names_dataframe = c("PD", "GBA_NMC", "HC"),
  numeric_columns = c("correlation", "pvalue"),
  names_dataframe_filter = c("PD_filter", "GBA_NMC_filter", "HC_filter"),
  condition_ifelse = expression(ifelse(data$subgroup %in% c("CE", "NE"), "enriched", "depleted")),
  clin_var_select = c("age", "bdi_total", "bmi", "DQS_score", "hads_anxiety_total", "hads_depression_total",
                      "hoehn_yahr", "moca_total", "rbdsq_total", "scopa_cardiovascular", "scopa_gastrointestinal",
                      "scopa_pupillomotor", "scopa_sexual", "scopa_thermoregulatory", "scopa_total", "scopa_urinary",
                      "updrs_i_total", "updrs_ii_total", "updrs_iii_total", "updrs_iv_total" ,"updrs_total", "upsit_total", "wcs_total",
                      "pd_dur", "ledd_total"),
  creating_df = as.data.frame(matrix(ncol = 4)),
  colnames_df_final = c("clinical_variable", "type", "correlation_pos", "correlation_neg")
)
```

Merge the new name of clinical variable with the old clinical variable name
```{r}
for (df in names(df_reshaped_sp)){
  df_reshaped_sp[[df]] <- df_reshaped_sp[[df]] %>%
    left_join(name_correspondance, by ="clinical_variable")
  if (df %in% c("GBA_NMC_filter", "HC_filter")){
    df_reshaped_sp[[df]] <- df_reshaped_sp[[df]] %>%
      filter(clinical_variable != "hoehn_yahr")
  }
}
```

Using the graph_figure4_paper function, which uses the dataframe reformatted with the previous function and assigns specific measurements to create a figure adapted to the dataframe data for each group. 
```{r}
list_plot_A_C <- graph_figure4_paper(
                              names_df_tobind = c("PD_filter","HC_filter", "GBA_NMC_filter"), # df to which we will add rows 
                              format_dataframe = list("PD_filter" = as.data.frame(matrix(ncol = 6, nrow = 1)), 
                                            "HC_filter" = as.data.frame(matrix(ncol = 6, nrow = 3)), 
                                            "GBA_NMC_filter" = as.data.frame(matrix(ncol = 6, nrow = 3))), # format of the rows bind with the iniatial dataframe
                              rows_tobind =  list("PD_filter" = NULL,
                                                  "HC_filter" = list(c("pd_dur", "enriched", 0, 0, 1, "PD Duration"), c("ledd_total", "enriched", 0, 0, 1, "Levodopa eq. daily dose"), c("updrs iv total", "enriched", 0, 0, 1, "MDS-UPDRS Part IV"), c("hoehn_yahr", "enriched", 0, 0, 1, "Hoehn & Yahr")), 
                                                  "GBA_NMC_filter" = list(c("pd_dur", "enriched", 0, 0, 1, "PD Duration"), c("ledd_total", "enriched", 0, 0, 1, "Levodopa eq. daily dose"), c("updrs iv total", "enriched", 0, 0, 1, "MDS-UPDRS Part IV"), c("hoehn_yahr", "enriched", 0, 0, 1, "Hoehn & Yahr"))), # the content of the lines that will be linked. The information added is that which does not exist in HC and GBA_NMC is the information relating to the disease. We add the rows with a correlation number set to 0 to have the information on the graph. 
                              numeric_columns =  c("correlation_pos", "correlation_neg", "chi2"), # indicate which column have to be numeric.
                              list_clinical_variable = list_clinical_variable, # list created previously with dclincial variable related to health and disease.
                              final_df = df_reshaped_sp,
                              clinical_var_col = "clinical_variable_bis",
                              to_exclude = c("age", "bmi"), # this variable are not of interest in for the figure. 
                              pos_val = "correlation_pos",
                              neg_val = "correlation_neg",
                              asterisk_pos = list("PD_filter" = 3, 
                                               "HC_filter" = 1, 
                                               "GBA_NMC_filter" = 1),
                              asterisk_neg = list("PD_filter" = 9, 
                                                 "HC_filter" = 4, 
                                                 "GBA_NMC_filter" = 4), # the value to add to get correct placement of the asterisk in the graph. 
                              name_type_cor_col = "correlation_type",
                              name_type_cor_val = "correlation_value",
                              rename_pos_val = "positive",
                              rename_neg_val = "negative",
                              ifelse_condition_y = ifelse_condition_y <- function(data_) {
                                                    ifelse(data_ == "PD_filter", "PD",
                                                    ifelse(data_ == "HC_filter", "HC", "GBA_NMC"))
                                                  }, # to obtain the group concerned and place it on the title of the y-axis. 
                              colors_fill = c("depletednegative" = PD_depleted_col, 
                                           "depletedpositive" = PD_depleted_col, 
                                           "enrichednegative" = PD_enriched_col, 
                                           "enrichedpositive" = PD_enriched_col), # color set for each barplot. 
                              keep_in_legend = c("depletedpositive", "enrichedpositive"),
                              labels_legend = c("depletedpositive" = "PD depleted", 
                                             "enrichedpositive" = "PD enriched"), # reformated the legend name. 
                              which_kind_plot = c("health", "disease"), #  two graphs are produced, one with the clinical variable relating to health, the other with the clinical variable relating to illness. They are combined after all the information is gathered in a single graph, but with a distinct separation. 
                              number_of = c("Number of", "species"), # what is measure here, is for completed the y axis title. 
                              label_correlation = c("Positive correlations","Negative correlations"),
                              limits_by = list("PD_filter" = seq(-90, 60, by = 30), 
                                            "HC_filter" = seq(-30, 30, by = 10), 
                                            "GBA_NMC_filter" = seq(-30, 30, by = 10)), # setting the number of thicks in the y axis. 
                              limits_ = list("PD_filter" = c(-90, 60), 
                                         "HC_filter" = c(-30, 30), 
                                         "GBA_NMC_filter" = c(-30, 30)), # setting the range of y axis
                              y_legend =  list("PD_filter" = c(0.95, 0.13), 
                                            "HC_filter" = c(0.95, 0.442), 
                                            "GBA_NMC_filter" = c(0.95, 0.13)), # placement for annotation added in the y axis. 
                              labs_plot = list("PD_filter" = labs(x = "",  fill = ""),
                                       "HC_filter" = labs(x = "Clinical variables",  fill = ""), 
                                       "GBA_NMC_filter" = labs(x = "",  fill = "")), # Name for the plot.
                              theme_size =   list("PD_filter" = theme(legend.position = "none",
                                                                        axis.text.x = element_blank(),
                                                                        legend.title = element_blank(),
                                                                        axis.title.y = element_text(size = 17, vjust = 1, color = "black"),
                                                                        axis.text.y = element_text(size = 15, color = "black"),
                                                                        axis.title.x = element_text(size = 17, vjust = (-3), color = "black"),
                                                                        plot.margin = unit(c(3,1,0,1), "cm")),
                                                  "HC_filter" = theme(legend.position = "bottom",
                                                                       legend.direction = "horizontal",
                                                                       axis.text.x = element_text(angle =45, hjust = 1, size = 14.5, color = "black"),
                                                                       legend.title = element_blank(),
                                                                       axis.title.y = element_text(size = 17, vjust = 1, color = "black"),
                                                                       axis.text.y = element_text(size = 15, color = "black"),
                                                                
                                                                       axis.title.x = element_text(size = 17, vjust = (-3), color = "black"),
                                                                       plot.margin = unit(c(0,1,0,1), "cm"),
                                                                       legend.key.size = unit(0.4, "cm"),
                                                                       legend.box.spacing = unit(0.7, "cm"),
                                                                       legend.text = element_text(size = 15), color = "black"),
                                                   "GBA_NMC_filter" = theme(legend.position = "none",
                                                                             axis.text.x = element_blank(),
                                                                             legend.title = element_blank(),
                                                                             axis.title.y = element_text(size = 17, vjust = 1, color = "black"),
                                                                             axis.text.y = element_text(size = 15, color = "black"),
                                                                            plot.margin = unit(c(1.5,1,2,1), "cm"))), # theme setting for each group
                              size_asterisk = list("PD_filter" = 2.8,
                                                "HC_filter" = 2.8, 
                                                "GBA_NMC_filter" = 2.8),
                              angle_asterisk = list("PD_filter" = 0,
                                                "HC_filter" = 0, 
                                                "GBA_NMC_filter" = 0),
                              x_legend = list("PD_filter" = c(0.25, 0.25),
                                                   "HC_filter" = c(0.25, 0.25), 
                                                   "GBA_NMC_filter" = c(0.25, 0.25)),
                              size_legend = list("PD_filter" = 5,
                                                   "HC_filter" = 5, 
                                                   "GBA_NMC_filter" = 5),
                              letter_fig = list("PD_filter" = "A", "HC_filter" = "C", "GBA_NMC_filter" = "B"),
                              theme_plot_title = list("PD_filter" = theme(plot.title = element_text(hjust = -0.02 ,size = 17, vjust = 1.1, face = "bold")),
                                                   "HC_filter" = theme(plot.title = element_text(hjust = -0.02 ,size = 17, vjust = 1.4, face = "bold")),
                                                   "GBA_NMC_filter" = theme(plot.title = element_text(hjust = -0.02 ,size = 17, vjust = 4.5, face = "bold"))),
                              theme_health_disease_1 = list("PD_filter" = theme(plot.margin = unit(c(0.1, 0, 0, 0), "cm")), 
                                                         "HC_filter" = theme(plot.margin = unit(c(0.1, 0, 3.54, 0), "cm"), 
                                                                             legend.position = "none",
                                                                             axis.title.x = element_blank()), 
                                                         "GBA_NMC_filter" = theme(plot.margin = unit(c(0.1, 0, 0, 0), "cm"))), # for the health relative clinical variable plot
                              theme_health_disease_2 = list("PD_filter" = theme(
                                                                axis.line.y = element_blank(),
                                                                axis.ticks.y = element_blank(),
                                                                axis.text.y = element_blank(),
                                                                axis.title.y = element_blank(),
                                                                plot.margin = unit(c(0.1, 0.2, 0, 0.2), "cm")
                                                              ), 
                                                        "HC_filter" = theme(
                                                                axis.title.x = element_text(size = 17, vjust = (-4), hjust = 0.3, color = "black"),
                                                                legend.text = element_text(size = 15),
                                                                legend.position = "none",
                                                                axis.line.y = element_blank(),
                                                                axis.ticks.y = element_blank(),
                                                                axis.text.y = element_blank(),
                                                                axis.title.y = element_blank(),
                                                                plot.margin = unit(c(0.1, 0.2, 0.5, 0.2), "cm"),
                                                                legend.margin = unit(c(0, 0, 0, 0), "cm")
                                                              ), 
                                                        "GBA_NMC_filter" = theme(
                                                                axis.line.y = element_blank(),
                                                                axis.ticks.y = element_blank(),
                                                                axis.text.y = element_blank(),
                                                                axis.title.y = element_blank(),
                                                                plot.margin = unit(c(0.1, 0.2, 0, 0.2), "cm"))), # for the disease relative clinical variable plot.
                              width_two_plots = c(0.27, 1)) # setting the width when associated the two plots.
```


```{r}
PD_mod <- final_df_mod %>%
            filter(group == "PD")

GBA_NMC_mod <- final_df_mod %>%
            filter(group == "GBA_NMC")

HC_mod <- final_df_mod %>%
            filter(group == "HC")

```


Reshaping the dataframe for the modules. 
```{r}
df_reshaped_mod <- prep_data_fig4(
  names_dataframe = c("PD_mod", "GBA_NMC_mod", "HC_mod"),
  numeric_columns = c("correlation", "pvalue"),
  names_dataframe_filter = c("PD_mod_filter", "GBA_NMC_mod_filter", "HC_mod_filter"),
  condition_ifelse = expression(ifelse(data$subgroup %in% c("CEM", "XEM", "NEM"), "enriched", "depleted")),
  clin_var_select = c("age", "bdi_total", "bmi", "DQS_score", "hads_anxiety_total", "hads_depression_total",
                      "hoehn_yahr", "moca_total", "rbdsq_total", "scopa_cardiovascular", "scopa_gastrointestinal",
                      "scopa_pupillomotor", "scopa_sexual", "scopa_thermoregulatory", "scopa_total", "scopa_urinary",
                      "updrs_i_total", "updrs_ii_total", "updrs_iii_total", "updrs_iv_total", "updrs_total", "upsit_total", "wcs_total",
                      "pd_dur", "ledd_total"),
  creating_df = as.data.frame(matrix(ncol = 4)),
  colnames_df_final = c("clinical_variable", "type", "correlation_pos", "correlation_neg")
)
```

Delete the ‘hoehn_yahr’ variable in the GBA_NMC and HC module database. 
```{r}
for (name_df in c("GBA_NMC_mod_filter", "HC_mod_filter")){
  
  df <- df_reshaped_mod[[name_df]]
  
  df_reshaped_mod[[name_df]] <- df %>%
                                  filter(!clinical_variable == "hoehn_yahr")
  
}
```


Replacing the clinical variable name by the new clinical variable name. 
```{r}
for (df in names(df_reshaped_mod)){
  df_reshaped_mod[[df]] <- df_reshaped_mod[[df]] %>%
    left_join(name_correspondance, by ="clinical_variable")
}
```


```{r}
list_plot_D_F <- graph_figure4_paper(names_df_tobind = c("PD_mod_filter", "HC_mod_filter", "GBA_NMC_mod_filter"),
                           format_dataframe = list("PD_mod_filter" = as.data.frame(matrix(ncol = 6, nrow = 1)), 
                                            "HC_mod_filter" = as.data.frame(matrix(ncol = 6, nrow = 3)), 
                                            "GBA_NMC_mod_filter" = as.data.frame(matrix(ncol = 6, nrow = 3))),
                           rows_tobind = list("PD_mod_filter" = NULL, 
                                            "HC_mod_filter" = list(c("pd_dur", "enriched", 0, 0, 1, "PD Duration"), c("ledd_total", "enriched", 0, 0, 1, "Levodopa eq. daily dose"), c("updrs_iv_total", "enriched", 0, 0, 1, "MDS-UPDRS Part IV"), c("hoehn_yahr", "enriched", 0, 0, 1, "Hoehn & Yahr")), 
                                            "GBA_NMC_mod_filter" = list(c("pd_dur", "enriched", 0, 0, 1, "PD Duration"), c("ledd_total", "enriched", 0, 0, 1, "Levodopa eq. daily dose"), c("updrs_iv_total", "enriched", 0, 0, 1, "MDS-UPDRS Part IV"),c("hoehn_yahr", "enriched", 0, 0, 1, "Hoehn & Yahr"))),
                           numeric_columns =  c("correlation_pos", "correlation_neg", "chi2"),
                           list_clinical_variable =  list_clinical_variable,
                           final_df = df_reshaped_mod,
                           clinical_var_col = "clinical_variable_bis",
                           to_exclude = c("age", "bmi"),
                           pos_val = "correlation_pos",
                           neg_val = "correlation_neg",
                           asterisk_pos = list("PD_mod_filter" = 7, 
                                               "HC_mod_filter" = 3, 
                                               "GBA_NMC_mod_filter" = 3),
                           asterisk_neg = list("PD_mod_filter" = 25, 
                                               "HC_mod_filter" = 12, 
                                               "GBA_NMC_mod_filter" = 7),
                           name_type_cor_col = "correlation_type",
                           name_type_cor_val = "correlation_value",
                           rename_pos_val = "positive",
                           rename_neg_val = "negative",
                           ifelse_condition_y = ifelse_condition_y <- function(data_) {
                                                  ifelse(data_ == "PD_mod_filter", "PD",
                                                  ifelse(data_ == "HC_mod_filter", "HC", "GBA_NMC"))
                                                },
                           colors_fill = c("depletednegative" = PD_depleted_col, 
                                           "depletedpositive" = PD_depleted_col, 
                                           "enrichednegative" = PD_enriched_col, 
                                           "enrichedpositive" = PD_enriched_col),
                           keep_in_legend = c("depletedpositive", "enrichedpositive"),
                           labels_legend = c("depletedpositive" = "PD depleted", 
                                             "enrichedpositive" = "PD enriched"),
                           which_kind_plot = c("health", "disease"),
                           number_of = c("Number of", "modules"),
                           label_correlation = c("Positive correlations","Negative correlations"),
                           limits_by = list("PD_mod_filter" = seq(-160, 330, by = 80),  
                                            "HC_mod_filter" = seq(-90, 80, by = 30), 
                                            "GBA_NMC_mod_filter" = seq(-60, 80, by = 20)),
                           limits_ = list("PD_mod_filter" = c(-160, 330), 
                                         "HC_mod_filter" = c(-90, 80), 
                                         "GBA_NMC_mod_filter" = c(-60, 80)),
                           y_legend =  list("PD_mod_filter" = c(0.95, 0.13), 
                                            "HC_mod_filter" = c(0.95, 0.442), 
                                            "GBA_NMC_mod_filter" = c(0.95, 0.11)),
                           labs_plot = list("PD_mod_filter" = labs(x = "",  fill = ""),
                                       "HC_mod_filter" = labs(x = "Clinical variables",  fill = ""), 
                                       "GBA_NMC_mod_filter" = labs(x = "",  fill = "")),
                           theme_size =   list("PD_mod_filter" = theme(legend.position = "none",
                                                                        axis.text.x = element_blank(),
                                                                        legend.title = element_blank(),
                                                                        axis.title.y = element_text(size = 17, vjust = 1, color = "black"),
                                                                        axis.text.y = element_text(size = 15, color = "black"),
                                                                        axis.title.x = element_text(size = 17, vjust = (-3), color = "black"),
                                                                        plot.margin = unit(c(3,1,0,1), "cm")),
                                                  "HC_mod_filter" = theme(legend.position = "bottom",
                                                                       legend.direction = "horizontal",
                                                                       axis.text.x = element_text(angle =45, hjust = 1, size = 14.5, color = "black"),
                                                                       legend.title = element_blank(),
                                                                       axis.title.y = element_text(size = 17, vjust = 1, color = "black"),
                                                                       axis.text.y = element_text(size = 15, color = "black"),
                                                                       axis.title.x = element_text(size = 17, vjust = (-3), color = "black"),
                                                                       plot.margin = unit(c(0,1,0,1), "cm"),
                                                                       legend.key.size = unit(0.4, "cm"),
                                                                       legend.box.spacing = unit(0.7, "cm"),
                                                                       legend.text = element_text(size = 15), color = "black"),
                                                   "GBA_NMC_mod_filter" = theme(legend.position = "none",
                                                                             axis.text.x = element_blank(),
                                                                             legend.title = element_blank(),
                                                                             axis.title.y = element_text(size = 17, vjust = 1, color = "black"),
                                                                             axis.text.y = element_text(size = 15, color = "black"),
                                                                            plot.margin = unit(c(1.5,1,2,1), "cm"))),
                           x_legend = list("PD_mod_filter" = c(0.263, 0.263),
                                                   "HC_mod_filter" = c(0.25, 0.25), 
                                                   "GBA_NMC_mod_filter" = c(0.25, 0.25)),
                           
                           size_legend = list("PD_mod_filter" = 5,
                                                   "HC_mod_filter" = 5, 
                                                   "GBA_NMC_mod_filter" = 5),

                           size_asterisk = list("PD_mod_filter" = 2.8,
                                                "HC_mod_filter" = 2.8, 
                                                "GBA_NMC_mod_filter" = 2.8),
                           
                           angle_asterisk = list("PD_mod_filter" = 0,
                                                "HC_mod_filter" = 0, 
                                                "GBA_NMC_mod_filter" = 0),

                           letter_fig = list("PD_mod_filter" = "D", "HC_mod_filter" = "F", "GBA_NMC_mod_filter" = "E"),
                           theme_plot_title = list("PD_mod_filter" = theme(plot.title = element_text(hjust = -0.02 ,size = 17, vjust = 1.1, face = "bold")),
                                                   "HC_mod_filter" = theme(plot.title = element_text(hjust = -0.02 ,size = 17, vjust = 1.5, face = "bold")),
                                                   "GBA_NMC_mod_filter" = theme(plot.title = element_text(hjust = -0.02 ,size = 17, vjust = 4.5, face = "bold"))),
                           theme_health_disease_1 = list("PD_mod_filter" = theme(plot.margin = unit(c(0.1, 0, 0, 0), "cm")), 
                                                         "HC_mod_filter" = theme(plot.margin = unit(c(0.1, 0, 3.54, 0), "cm"), 
                                                                             legend.position = "none",
                                                                             axis.title.x = element_blank()), 
                                                         "GBA_NMC_mod_filter" = theme(plot.margin = unit(c(0.1, 0, 0, 0), "cm"))),
                           theme_health_disease_2 = list("PD_mod_filter" = theme(
                                                                axis.line.y = element_blank(),
                                                                axis.ticks.y = element_blank(),
                                                                axis.text.y = element_blank(),
                                                                axis.title.y = element_blank(),
                                                                plot.margin = unit(c(0.1, 0.2, 0, 0.2), "cm")
                                                              ), 
                                                        "HC_mod_filter" = theme(
                                                                axis.title.x = element_text(size = 17, vjust = (-4), hjust = 0.3, color = "black"),
                                                                legend.text = element_text(size = 15),
                                                                legend.position = "none",
                                                                axis.line.y = element_blank(),
                                                                axis.ticks.y = element_blank(),
                                                                axis.text.y = element_blank(),
                                                                axis.title.y = element_blank(),
                                                                plot.margin = unit(c(0.1, 0.2, 0.5, 0.2), "cm"),
                                                                legend.margin = unit(c(0, 0, 0, 0), "cm")
                                                              ), 
                                                        "GBA_NMC_mod_filter" = theme(
                                                                axis.line.y = element_blank(),
                                                                axis.ticks.y = element_blank(),
                                                                axis.text.y = element_blank(),
                                                                axis.title.y = element_blank(),
                                                                plot.margin = unit(c(0.1, 0.2, 0, 0.2), "cm")
                                                              )),
                           width_two_plots = c(0.27, 1)
                           
)
                           
```

```{r}
plots_bis_combined <- plot_grid(list_plot_A_C[["plots"]][[1]], 
                                list_plot_D_F[["plots"]][[1]], 
                                list_plot_A_C[["plots"]][[2]], 
                                list_plot_D_F[["plots"]][[2]],
                                list_plot_A_C[["plots"]][[3]], 
                                list_plot_D_F[["plots"]][[3]],  
                                nrow = 3, ncol = 2, rel_heights = c(1, 1, 1.3))


```

```{r fig.width = 16, fig.height = 18, fig.show='asis'}
plot_final <- plot_grid(
  plots_bis_combined, list_plot_A_C[["legends"]][[1]], 
  ncol= 1, 
  rel_heights = c(1, 0.05)
)
plot_final
```



# Extended Figure 1
```{r}
# BARPLOT MSP CONTRAST - HC vs PD
my_grp_compa <- combn(vec_group, m = 2)
sign_lvl <- "p<0.05"

# sort the bars according to the cliffdeltas
testR_msp.sign_HC_PD.ord <- testR_msp.sign_HC_PD %>% arrange(desc(HC.PD_Cliffdelta))
testR_msp.sign_HC_PD.ord$common_msp_sign <- ifelse(testR_msp.sign_HC_PD.ord$msp %in% common_msp21, "bold", "plain")

bp_msp_HC_PD <- ggbarplot(testR_msp.sign_HC_PD.ord, x = "msp", y = "HC.PD_Cliffdelta", color = "white", fill = "HC.PD_status_Cliffdeltacorrected", rotate = T, xlab = "", ylab = "CD - HC.PD") + 
  scale_fill_manual(values = color_HC_PD) + 
  ggtitle(paste("Prevalence > 10% ; ", paste0(my_grp_compa[1,2], " (n=", tab_group[my_grp_compa[1,2]], ") vs ", my_grp_compa[2,2], " (n=", tab_group[my_grp_compa[2,2]], ")", "; ", sign_lvl))) +
  theme(plot.title = element_text(hjust = 1.2, size = 18), 
        axis.title = element_text(size = 13), 
        axis.text.x = element_text(size = 13), 
        axis.text.y = element_text(size = 13, face = testR_msp.sign_HC_PD.ord$common_msp_sign), 
        legend.position = "none")
```



```{r}
# BARPLOT MSP CONTRAST - HC vs GBA_NMC
my_grp_compa <- combn(vec_group, m = 2)
sign_lvl <- "p<0.05"

# sort the bars according to the cliffdeltas
testR_msp.sign_HC_GBA_NMC.ord <- testR_msp.sign_HC_GBA_NMC %>% arrange(desc(HC.GBA_NMC_Cliffdelta))
testR_msp.sign_HC_GBA_NMC.ord$common_msp_sign <- ifelse(testR_msp.sign_HC_GBA_NMC.ord$msp %in% common_msp21, "bold", "plain")

bp_msp_HC_NMC <- ggbarplot(testR_msp.sign_HC_GBA_NMC.ord, x = "msp", y = "HC.GBA_NMC_Cliffdelta", color = "white", fill = "HC.GBA_NMC_status_Cliffdeltacorrected", rotate = T, xlab = "", ylab = "CD - HC.GBA_NMC") + 
  scale_fill_manual(values = color_HC_GBA_NMC) + 
  ggtitle(paste("Prevalence > 10% ; ", paste0(my_grp_compa[1,1], " (n=", tab_group[my_grp_compa[1,1]], ") vs ", my_grp_compa[2,1], " (n=", tab_group[my_grp_compa[2,1]], ")", "; ", sign_lvl))) +
  theme(plot.title = element_text(hjust = 1, size = 18), 
        axis.title = element_text(size = 13), 
        axis.text.x = element_text(size = 13), 
        axis.text.y = element_text(size = 13, face = testR_msp.sign_HC_GBA_NMC.ord$common_msp_sign),
        legend.position = "none")
``` 


```{r, fig.height=35, fig.width=24, fig.show='asis'}
# MERGED BARPLOT HC vs PD/GBA_NMC
fig_contrast <- ggarrange(NULL, bp_msp_HC_PD, NULL, bp_msp_HC_NMC, ncol = 4, widths = c(0.1, 1.2,0.1, 1), labels = c("", "A", "", "B"), font.label = list(size = 25)) 
fig_contrast

``` 

# Extended Figure 2 - PD 

```{r}
# custom final barplot 
custom_theme <- theme(axis.text = element_text(size = 15), 
                      axis.title = element_text(size = 15), 
                      plot.title = element_text(size = 15), 
                      legend.title = element_blank(), 
                      legend.position = "none")

# name of final barplots 
name_sub_group <- c("Coherent enriched species", 
                    "Coherent depleted species", 
                    "Noncoherent enriched species", 
                    "Noncoherent depleted species") 
```


```{r, fig.width=8}
# BARPLOT ABUNDANCE QUARTILE - PD  

# Create a list of vectors containing abundances computed in PD individuals for each subtype 
PD_abund <- list(
  CE = list(data = grp_CE_abund$PD, type = "E"),
  CD = list(data = grp_CD_abund$PD, type = "D"),
  NE = list(data = grp_NE_abund$PD, type = "E"),
  ND = list(data = grp_ND_abund$PD, type = "D")
) 
# Apply "get_quartile" function to each vectors from "PD_abund"
# The function sort abundances in ascending way when it's enriched specie type and in decreasing way when it's depleted specie type  
# Then it split into quartile groups and compute mean of abundance for each quartile 
qPD_abund <- lapply(PD_abund, function(g){ get_quartile(g$data, g$type) })

# List of dataframe of that contains the mean abundances of each quartile for each subtype specie 
df_list_qPD_abund <- lapply(qPD_abund, function(g){ compute_mean(g) })


# Create list of abundance barplot for each specie type 
bp_list_qPD_abund <- list()
for (i in 1:length(df_list_qPD_abund)){
  bp_name <- name_sub_group[i]
  bp_color <- ifelse(str_detect(bp_name, ".*enriched.*"), PD_enriched_col, PD_depleted_col)
  bp <- ggbarplot(df_list_qPD_abund[[i]], x = "variable", y = "value", fill = bp_color, title = bp_name, ylab = "", xlab = "") + custom_theme
  bp_list_qPD_abund[[i]] <- bp
}

# Arrange barplot
fig_qPD_abund <- ggarrange(plotlist = bp_list_qPD_abund)
fig_qPD_abund <- annotate_figure(fig_qPD_abund, left = text_grob("Abundance (%)", color = "black", rot = 90, size = 15)) 
fig_qPD_abund
```




```{r, fig.show='hide'}
# BARPLOT ABUNDANCE FC QUARTILE Q1 Q4

# Compute fold change (FC) of mean abundance between Q1 and Q4 for each subtype specie 
# and create dataframe of FC of mean abundance Q1-Q4 for each subtype specie 
PD_abund_fc <- rbind(
  process_fc(df_list_qPD_abund$CE, "Coherent", "Enriched species"),
  process_fc(df_list_qPD_abund$CD, "Coherent", "Depleted species"),
  process_fc(df_list_qPD_abund$NE, "Non-coherent", "Enriched species"),
  process_fc(df_list_qPD_abund$ND, "Non-coherent", "Depleted species")
  )
PD_abund_fc <- PD_abund_fc %>% 
  mutate(status = factor(status, levels = c("Enriched species", "Depleted species")))

# Create barplot of FC of mean abundance Q1-Q4 for each subtype specie 
fig_fc_qPD_abund <- ggbarplot(PD_abund_fc, 
                                x ="cat", 
                                y = "fc", 
                                fill = "status",
                                position = position_dodge(), 
                                palette = c(PD_enriched_col, PD_depleted_col),
                                title = "Abundance Fold Change Q1 - Q4") + 
  theme_classic() + xlab("") + ylab("") + 
  scale_x_discrete(labels=c("Coherent" = "Coherent\nspecies", 
                            "Non-coherent" = "Noncoherent\nspecies")) + 
  expand_limits(y = c(0,4)) + custom_theme

fig_fc_qPD_abund
```


```{r, fig.width=8, fig.height=6, fig.show='hide'}
# BARPLOT PROPORTION QUARTILE

# Create a list of vectors containing proportions computed in PD individuals for each subtype 
PD_prop <- list(
  CE = list(data = grp_CE_prop$PD, type = "E"),
  CD = list(data = grp_CD_prop$PD, type = "D"),
  NE = list(data = grp_NE_prop$PD, type = "E"),
  ND = list(data = grp_ND_prop$PD, type = "D")
) 
qPD_prop <- lapply(PD_prop, function(g){ get_quartile(g$data, g$type) })

# List of dataframe of means of proportions in each quartile for each subtype 
df_list_qPD_prop <- lapply(qPD_prop, function(g){ compute_mean(g) })


# Create list of proportion barplot for each specie type 
bp_list_qPD_prop <- list()
for (i in 1:length(df_list_qPD_prop)){
  bp_name <- name_sub_group[i]
  bp_color <- ifelse(str_detect(bp_name, ".*enriched.*"), PD_enriched_col, PD_depleted_col)
  bp <- ggbarplot(df_list_qPD_prop[[i]], x = "variable", y = "value", fill = bp_color, title = bp_name, ylab = "", xlab = "") + custom_theme
  bp_list_qPD_prop[[i]] <- bp
}

# Arrange barplot
fig_qPD_prop <- ggarrange(plotlist = bp_list_qPD_prop)
fig_qPD_prop <- annotate_figure(fig_qPD_prop, left = text_grob("Proportion (%)", color = "black", rot = 90, size = 15)) 
fig_qPD_prop
```





```{r}
# PROPORTION FC QUARTILE Q1 Q4 - PD

# Compute fold change (FC) of mean proportion between Q1 and Q4 for each subtype specie 
# and create dataframe of FC of mean proportion Q1-Q4 for each subtype specie 
PD_prop_fc <- rbind(
  process_fc(df_list_qPD_prop$CE, "Coherent", "Enriched species"),
  process_fc(df_list_qPD_prop$CD, "Coherent", "Depleted species"),
  process_fc(df_list_qPD_prop$NE, "Non-coherent", "Enriched species"),
  process_fc(df_list_qPD_prop$ND, "Non-coherent", "Depleted species")
  )
PD_prop_fc <- PD_prop_fc %>% 
  mutate(status = factor(status, levels = c("Enriched species", "Depleted species")))

# Create barplot of FC of mean proportion Q1-Q4 for each subtype specie
fig_fc_qPD_prop <- ggbarplot(PD_prop_fc, 
                             x ="cat", 
                             y = "fc", 
                             fill = "status", 
                             position = position_dodge(), 
                             palette = c(PD_enriched_col, PD_depleted_col), 
                             title = "Proportion Fold Change Q1 - Q4") + 
  theme_classic() + xlab("") + ylab("") + 
  scale_x_discrete(labels=c("Coherent" = "Coherent\nspecies", 
                          "Non-coherent" = "Noncoherent\nspecies")) + 
  expand_limits(y = c(0,4)) + custom_theme

fig_fc_qPD_prop
```




```{r, fig.height=11, fig.width = 8, fig.show='hide'}
# BARPLOT PROPORTION & ABUNDANCE QUARTILE - PD
fig_qPD_left <- ggarrange(fig_qPD_abund, fig_qPD_prop, nrow = 2, labels = c("A","B"), font.label = list(size = 17))
fig_qPD_left
```


```{r, fig.height=11, fig.width = 8, fig.show='hide'}
# BARPLOT ABUND/PROP FC QUARTILE Q1 Q4 - PD
fig_qPD_right <- ggarrange(fig_fc_qPD_abund, fig_fc_qPD_prop, nrow = 2, labels = c("C","D"), font.label = list(size = 17))
fig_qPD_right
``` 


```{r, fig.height=9, fig.width = 13, fig.show='asis'}
# MERGED ABUND/PROP QUARTILE AND ABUND/PROP FC QUARTILE
fig_qPD_top <- ggarrange(fig_qPD_left, fig_qPD_right, ncol = 2, widths = c(2, 0.9))
fig_qPD_top 
``` 



# Extended Figure 3
```{r, results='hide'}
# PROPORTION - Medication on microbiome alterations in PD

# select 32 PD drug naive individuals
PD_drug_naive_ind <- temp_metadata %>% 
  dplyr::filter(group_bis == "PD" & pd_drug_naive == "yes") %>% 
  pull(MGP_ID_fecal_baseline)

# select 47 PD medicated with lowest disease duration individuals (that matched for the mean duration of disease of drug-naive)
select_PD_med_ind <- paste0("ASAPh_", c("1138", "557", "1106", "1118", "334", "1140", "230", "223", "682", "629", "359", "656", "241", "964", "801", "1165", "1091", "641", "358", "212", "094", "074", "627", "055", "1095", "831", "597", "009", "035", "598", "260", "566", "054", "1083", "609", "248", "350", "250", "967", "976", "021", "966", "1115", "071", "1089", "617", "667"))


vec_sub_group <- c("CE", "NE", "CD", "ND") 
# In PD drug naive individuals, compute proportion of all MSP and of each subgroup MSP 
pd_drug_naive_prop <- compute_prop_subtype(vec_sub_group, PD_drug_naive_ind, msp_fecal.filt10)

# In PD medicated individuals, compute proportion of all MSP and of each subgroup MSP 
pd_med_prop <- compute_prop_subtype(vec_sub_group, select_PD_med_ind, msp_fecal.filt10)

# Student test
pd_dose_prop_ttest <- compute_ttest(pd_drug_naive_prop, pd_med_prop, grp = "ledd_status", grp_label = c("No_levodopa", "Yes_levodopa"))

# compute mean proportion of MSP for PD drug naive and PD specific individuals
tab_pd_dose_prop_melt <- show_mean_tab(pd_drug_naive_prop, pd_med_prop, vec_cat = c("No_levodopa", "Yes_levodopa")) 

tab_pd_dose_prop_melt <- tab_pd_dose_prop_melt %>% 
  mutate(variable_bis = ifelse(variable %in% c("CE", "NE"), "Enriched", "Depleted")) %>% 
  mutate(cat_color = paste(variable_bis, cat, sep = "_")) %>% 
  mutate(cat_color = factor(cat_color, levels = c("Enriched_No_levodopa", "Enriched_Yes_levodopa", "Depleted_No_levodopa", "Depleted_Yes_levodopa")))

pd_dose_color <- setNames(c(rep(PD_enriched_hue, 2), rep(PD_depleted_hue, 2)),  tab_pd_dose_prop_melt$cat_color)  
```


```{r, fig.width=9}
# barplot 
my_compa <- c("No_levodopa", "Yes_levodopa")
bp_PD_med_prop <- ggbarplot(tab_pd_dose_prop_melt, x = "variable", y = "value", fill = "cat_color", position = position_dodge()) + 
  scale_fill_manual(values = pd_dose_color) +
  ylab("MSP mean proportion (%)") + 
  theme_classic()+
  ggtitle("Proportion of the altered microbiome in PD patients") + 
  theme(legend.title=element_blank(), axis.title.x = element_blank(), axis.text = element_text(size = 14)) +
  scale_x_discrete(labels=c("CE" = "Enriched coherent\nspecies", 
                            "NE" = "Enriched\nnoncoherent species",
                            "CD" = "Depleted coherent\nspecies", 
                            "ND" = "Depleted\nnoncoherent species")) + 
  scale_y_continuous(breaks=seq(0, 16, 1)) + 
  geom_text(data = pd_dose_prop_ttest, aes(x=y, y=15, label=paste0("Student.p:\n", p)), size = 5) 
``` 


```{r}
# ABUNDANCE - Medication on microbiome alterations in PD
vec_sub_group <- c("CE", "NE", "CD", "ND")

# In PD drug naive individuals (No levodopa), sum abundance of all MSP and of each subgroup MSP (CE, NE, CD, ND)
pd_dnaive_abund <- compute_abund_subtype(vec_sub_group, PD_drug_naive_ind, msp_fecal.filt10)

# In PD medicated individuals (Yes levodopa, and dose matched), sum abundance of all MSP and of each subgroup MSP (CE, NE, CD, ND)
pd_med_abund <- compute_abund_subtype(vec_sub_group, select_PD_med_ind, msp_fecal.filt10)

# Student test
pd_dose_abund_ttest <- compute_ttest(pd_dnaive_abund, pd_med_abund, grp = "ledd_status", grp_label = c("No_levodopa", "Yes_levodopa"))
```


```{r, fig.width = 10}
# BARPLOT ABUNDANCE - DRUG NAIVE VS MEDICATED IN PD 

# compute mean abundance for each PD species type according to medication and build table of mean abundance  
tab_pd_dose_abund_melt <- show_mean_tab(pd_dnaive_abund, pd_med_abund, vec_cat = c("No_levodopa", "Yes_levodopa"))

tab_pd_dose_abund_melt <- tab_pd_dose_abund_melt %>% 
  mutate(variable_bis = ifelse(variable %in% c("CE", "NE"), "Enriched", "Depleted")) %>% 
  mutate(cat_color = paste(variable_bis, cat, sep = "_")) %>% 
  mutate(cat_color = factor(cat_color, levels = c("Enriched_No_levodopa", "Enriched_Yes_levodopa", "Depleted_No_levodopa", "Depleted_Yes_levodopa")))

pd_dose_color <- setNames(c(rep(PD_enriched_hue, 2), rep(PD_depleted_hue, 2)),  tab_pd_dose_abund_melt$cat_color)
```



```{r}
# PROPORTION - DISEASE DURATION in PD
# For study the duration of disease effect on microbiome : 61 PD individuals with different disease duration (4.2 and 7.2 years on average, N=30 and N=31 respectively) that matched for the mean dose of levodopa (600 mg on average). 

# select 30 short duration PD individuals 
PD_short_ind <- paste0("ASAPh_", c("359", "212", "566", "1083", "021", "006", "020", "371", "261", "030", "266", "1149", "989", "688", "353", "267", "968", "852", "588", "606", "363", "618", "235", "585", "381", "1101", "1098", "091", "022", "010"))

# select 31 long duration PD individuals
PD_long_ind <- paste0("ASAPh_", c("335", "988", "649", "043", "352", "038", "036", "643", "987", "040", "067", "817", "098", "031", "355", "828", "033", "005", "234", "049", "603", "342", "254", "991", "251", "690", "819", "237", "624", "354", "376"))

# extract the pd_dur and ledd_total for the 61 individuals  
select_PD_data <- temp_metadata[c(PD_short_ind, PD_long_ind),] 
select_PD_data <- select_PD_data %>% 
  mutate(pd_dur_cat = ifelse(rownames(select_PD_data) %in% PD_short_ind, "shorter", "longer"))
select_PD_data <- select_PD_data %>% dplyr::select(group_bis, pd_dur, pd_dur_cat, ledd_total)
``` 



```{r}
# In PD shorter duration disease, compute the count and the proportion of all MSP and of each subgroup MSP
pd_short_prop <- compute_prop_subtype(vec_sub_group, PD_short_ind, msp_fecal.filt10)

# In PD longer duration disease, compute the count and the proportion of all MSP and of each subgroup MSP
pd_long_prop <- compute_prop_subtype(vec_sub_group, PD_long_ind, msp_fecal.filt10)

# compute mean proportion for each species type according to PD duration and build table of mean proportion
tab_pd_duration_prop_melt <- show_mean_tab(pd_short_prop, pd_long_prop, vec_cat = c("shorter", "longer"))

# Student test
pd_duration_prop_ttest <- compute_ttest(pd_short_prop, pd_long_prop, grp = "pd_duration", grp_label = c("shorter", "longer")) 
```



```{r, fig.width=9, results='markup', fig.show='hide'}
# PROPORTION - Disease duration on microbiome alterations in PD

tab_pd_duration_prop_melt <- tab_pd_duration_prop_melt %>% 
  mutate(variable_bis = ifelse(variable %in% c("CE", "NE"), "Enriched", "Depleted")) %>% 
  mutate(cat_color = paste(variable_bis, cat, sep = "_")) %>% 
  mutate(cat_color = factor(cat_color, levels = c("Enriched_shorter", "Enriched_longer", "Depleted_shorter", "Depleted_longer"))) 
pd_duration_color <- setNames(c(rep(PD_enriched_hue, 2), rep(PD_depleted_hue, 2)), tab_pd_duration_prop_melt$cat_color)
```


```{r}
# ABUNDANCE - DISEASE DURATION IN PD 
vec_sub_group <- c("CE", "NE", "CD", "ND")

# In PD shorter duration disease individuals, sum abundance of all MSP and of each subgroup MSP
pd_short_abund <- compute_abund_subtype(vec_sub_group, PD_short_ind, msp_fecal.filt10)

# In PD longer duration disease individuals, sum abundance of all MSP and of each subgroup MSP
pd_long_abund <- compute_abund_subtype(vec_sub_group, PD_long_ind, msp_fecal.filt10)

# compute mean abundance for each type according to shorter or longer duration and build table of mean 
tab_pd_duration_abund_melt <- show_mean_tab(pd_short_abund, pd_long_abund, vec_cat = c("shorter", "longer"))

# Student test
pd_duration_abund_ttest <- compute_ttest(pd_short_abund, pd_long_abund, grp = "pd_duration", grp_label = c("shorter", "longer"))  
```


    
```{r, fig.width = 8}
# ABUNDANCE - DISEASE DURATION IN PD
tab_pd_duration_abund_melt <- tab_pd_duration_abund_melt %>% 
  mutate(variable_bis = ifelse(variable %in% c("CE", "NE"), "Enriched", "Depleted")) %>% 
  mutate(cat_color = paste(variable_bis, cat, sep = "_")) %>% 
  mutate(cat_color = factor(cat_color, levels = c("Enriched_shorter", "Enriched_longer", "Depleted_shorter", "Depleted_longer")))

pd_duration_color <- setNames(c(rep(PD_enriched_hue, 2), rep(PD_depleted_hue, 2)),  tab_pd_duration_abund_melt$cat_color) 
``` 


```{r, fig.width=10}
# MIX BARPLOT - DOSE EFFECT IN PD (drug naive vs medicated)
# enriched represented by abundance and depleted represented by proportion
select_dose_abund <- tab_pd_dose_abund_melt %>% filter(variable %in% c("CE", "NE"))
select_dose_prop <- tab_pd_dose_prop_melt %>% filter(variable %in% c("CD","ND"))
tab_select_dose <- select_dose_abund %>% rbind(select_dose_prop)

# bind the abundance ttest for enriched species with the proportion ttest for depleted species 
pd_dose_select_test <- pd_dose_abund_ttest %>% filter(y %in% c("CE", "NE")) %>%
  rbind(pd_dose_prop_ttest %>% filter(y %in% c("CD", "ND")))
# round pval at 2 decimals
pd_dose_select_test <- pd_dose_select_test %>% mutate(p = round(p, 3))

# barplot
bp_dose_mix <- ggbarplot(tab_select_dose, x = "variable", y = "value", fill = "cat_color", position = position_dodge2(), title = "Effect of medication on microbiome alterations in PD", ylab = "") +
  scale_fill_manual(values = pd_dose_color) +
  theme_classic()+
  theme(title = element_text(size = 18),
        axis.title.x = element_blank(), 
        axis.text = element_text(size = 15), 
        legend.title=element_blank(), 
        legend.text = element_text(size = 14.5), 
        legend.position = "bottom", 
        legend.key.size = unit(1, 'cm'), 
        legend.key.spacing.x = unit(1, "cm")) +
  guides(fill = guide_legend(label.position = "bottom")) + 
  scale_x_discrete(labels=c("CE" = "Enriched coherent\nspecies\n(mean abund.)",
                            "NE" = "Enriched\nnoncoherent species\n(mean abund.)",
                            "CD" = "Depleted coherent\nspecies\n(mean prop.)",
                            "ND" = "Depleted\nnoncoherent species\n(mean prop.)")) +
  scale_y_continuous(breaks = seq(0, 15, 1)) +
  expand_limits(y = c(0, 17)) +
  geom_text(data = pd_dose_select_test, aes (x=y, y=16, label = paste0("Student.p:\n", p)), size = 7)

bp_dose_mix
```



```{r, fig.width=8}
# MIX BARPLOT - DISEASE DURATION IN PD 
# enriched represented by abundance and depleted represented by proportion
select_duration_abund <- tab_pd_duration_abund_melt %>% filter(variable %in% c("CE", "NE"))
select_duration_prop <- tab_pd_duration_prop_melt %>% filter(variable %in% c("CD","ND"))
tab_select_duration <- select_duration_abund %>% rbind(select_duration_prop)

# bind the abundance ttest for enriched species with the proportion ttest for depleted species 
pd_duration_select_test <- pd_duration_abund_ttest %>% filter(y %in% c("CE", "NE")) %>%
  rbind(pd_duration_prop_ttest %>% filter(y %in% c("CD", "ND")))
pd_duration_select_test <- pd_duration_select_test %>% mutate(p = round(p, 3))

# barplot
bp_duration_mix <- ggbarplot(tab_select_duration, x = "variable", y = "value", fill = "cat_color", position = position_dodge2(), title = "Effect of disease duration on microbiome alterations in PD", ylab = "") + 
  scale_fill_manual(values = pd_duration_color) + 
  theme_classic()+
  theme(title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.text = element_text(size = 15),
        legend.title=element_blank(),
        legend.text = element_text(size = 14.5),
        legend.position = "bottom",
        legend.key.size = unit(1, 'cm'),
        legend.key.spacing.x = unit(1, "cm")) +
  guides(fill = guide_legend(label.position = "bottom")) +
  scale_x_discrete(labels=c("CE" = "Enriched coherent\nspecies\n(mean abund.)",
                            "NE" = "Enriched\nnoncoherent species\n(mean abund.)",
                            "CD" = "Depleted coherent\nspecies\n(mean prop.)",
                            "ND" = "Depleted\nnoncoherent species\n(mean prop.)")) +
  scale_y_continuous(breaks = seq(0, 15, 1)) +
  expand_limits(y = c(0, 17)) +
  geom_text(data = pd_duration_select_test, aes(x=y, y=16, label = paste0("Student.p:\n", p)), size = 7)
  
bp_duration_mix
```



```{r, fig.height=9, fig.width=21, fig.show='asis'}
fig_medication_duration <- ggarrange(bp_dose_mix, NULL, bp_duration_mix, ncol = 3, labels = c("A", "", "B"), widths = c(1, 0.06, 1.04), font.label = list(size = 23))
fig_medication_duration
```
 

# Extended Figure 4
```{r}
# Create a list of vectors containing abundances/proportions computed in GBA_NMC individuals for each subtype (CE and CD) 
NMC_coh <- list(
  CE_abund = list(data = grp_CE_abund$GBA_NMC, type = "E"),
  CD_abund = list(data = grp_CD_abund$GBA_NMC, type = "D"),
  CE_prop = list(data = grp_CE_prop$GBA_NMC, type = "E"),
  CD_prop = list(data = grp_CD_prop$GBA_NMC, type = "D")
)

# Apply "get_median" function to each vectors from "NMC_coherent"
# The function sort abundances/proportions in ascending way when it's enriched specie type and in decreasing way when it's depleted specie type  
# Then it split into median groups and compute mean of abundance/proportion for each median groups  
median_NMC_coh <- lapply(NMC_coherent, function(g){ get_median(g$data, g$type)}) 


# List of dataframe of that contains the mean abundances/proportions of each subgroup for each subtype specie 
median_NMC_means <- lapply(median_NMC_coh, function(g) { compute_mean(g)})

# check 
length(median_NMC_coh$CD_abund$below)
length(median_NMC_coh$CD_abund$above)

length(median_NMC_coh$CE_abund$below)
length(median_NMC_coh$CE_abund$above)
```

 
```{r}
# Student test of median groups (below vs above) 
median_NMC_test <- lapply(median_NMC_coh, function(g) { 
  direction <- ifelse(mean(g$below) < mean(g$above), "less", "greater")  
  ttest_res <- t.test(g$below, g$above, alternative = direction, var.equal = T)
  signif(ttest_res$p.value, 3) 
  }) %>% unlist()
 
# Fold Change of means (below vs above)
median_NMC_fc <- lapply(median_NMC_means, function(g) {
  min_val <- min(g$value)
  max_val <- max(g$value)
  fc <- foldchange(max_val, min_val)
  signif(fc, 3)
}) %>% unlist() 
```
 
 
```{r}
# BARPLOT MEDIAN - GBA_NMC
name_median <- setNames(c("Abundance of coherent PD\nenriched species", 
                          "Abundance of coherent PD\ndepleted species",
                          "Proportion of coherent PD\nenriched species",
                          "Proportion of coherent PD\ndepleted species"), names(median_NMC_means))

bp_median_list <- list()
for (metric in names(median_NMC_means)){
  data <- median_NMC_means[[metric]]
  bp_title <- name_median[[metric]]
  bp_color <- ifelse(str_detect(bp_title, ".*enriched.*"), PD_enriched_col, PD_depleted_col)
  p <- show_barplot(data, 
                    x = "variable",
                    y = "value", 
                    fill = bp_color, 
                    title = bp_title, 
                    show_legend = "right", 
                    limits = c(0, round(max(data$value))+2), 
                    pas = 2, 
                    ylab = "%", 
                    size_title = 16.5, size_title_axis = 17, size_axis_text = 16) + 
    annotate(geom = "text", label = paste("FC:", median_NMC_fc[[metric]]), x = 1.5, y=round(max(data$value))+3, size = 5) + 
    annotate(geom = "text", label = paste("Student.p:", median_NMC_test[[metric]]), x=1.5, y=round(max(data$value))+1.5, size = 5)

  bp_median_list[[metric]] <- p
}

bp_median_list
```

```{r}
# BARPLOT ABUNDANCE COHERENT MEDIAN
fig_median_top <- ggarrange(bp_median_list$CE_abund, bp_median_list$CD_abund, ncol = 2, labels = c("A", "B"), font.label = list(size = 17))

# BARPLOT PROPORTION COHERENT MEDIAN
fig_median_bottom <- ggarrange(bp_median_list$CE_prop, bp_median_list$CD_prop, ncol = 2, labels = c("C", "D"), font.label = list(size = 17))

fig_median_top
fig_median_bottom
```
 
```{r, fig.height=7, fig.width=8, fig.show='asis'}
# FINAL ABUND/PROP COHERENT MEDIAN
fig_median <- ggarrange(fig_median_top, "", fig_median_bottom, nrow = 3, heights = c(1, 0.05 ,1))
fig_median
```
 

# Extended Figure 5 - HC

```{r, fig.width=8}
# BARPLOT ABUNDANCE QUARTILE  

# Create a list of vectors containing abundances computed in HC individuals for each subtype 
HC_abund <- list(
  CE = list(data = grp_CE_abund$HC, type = "E"),
  CD = list(data = grp_CD_abund$HC, type = "D"),
  NE = list(data = grp_NE_abund$HC, type = "E"),
  ND = list(data = grp_ND_abund$HC, type = "D")
) 

# Apply "get_quartile" function to each vectors from "HC_abund"
# The function sort abundances in ascending way when it's enriched specie type and in decreasing way when it's depleted specie type  
# Then it split into quartile groups and compute mean of abundance for each quartile 
qHC_abund <- lapply(HC_abund, function(g){ get_quartile(g$data, g$type)}) 

# List of dataframe of that contains the mean abundances of each quartile for each subtype specie 
df_list_qHC_abund <- lapply(qHC_abund, function(g) { compute_mean(g)})
 

# Create list of abundance barplot for each specie type 
bp_list_qHC_abund <- list()
for (i in 1:length(df_list_qHC_abund)){
  bp_name <- name_sub_group[i]
  bp_color <- ifelse(str_detect(bp_name, ".*enriched.*"), PD_enriched_col, PD_depleted_col)
  bp <- ggbarplot(df_list_qHC_abund[[i]], x = "variable", y = "value", fill = bp_color, title = bp_name, ylab = "", xlab = "") + custom_theme
  bp_list_qHC_abund[[i]] <- bp
}

# Arrange barplot
fig_qHC_abund <- ggarrange(plotlist = bp_list_qHC_abund)
fig_qHC_abund <- annotate_figure(fig_qHC_abund, left = text_grob("Abundance (%)", color = "black", rot = 90, size = 15)) 
fig_qHC_abund
```



```{r, fig.show='hide'}
# BARPLOT ABUNDANCE FC QUARTILE Q1 Q4 - HC

# Compute fold change (FC) of mean abundance between Q1 and Q4 for each subtype specie 
# and create dataframe of FC of mean abundance Q1-Q4 for each subtype specie 
HC_abund_fc <- rbind(
  process_fc(df_list_qHC_abund$CE, "Coherent", "Enriched species"),
  process_fc(df_list_qHC_abund$CD, "Coherent", "Depleted species"),
  process_fc(df_list_qHC_abund$NE, "Non-coherent", "Enriched species"),
  process_fc(df_list_qHC_abund$ND, "Non-coherent", "Depleted species"))

HC_abund_fc <- HC_abund_fc %>% 
  mutate(status = factor(status, levels = c("Enriched species", "Depleted species")))  

# Create barplot of FC of mean abundance Q1-Q4 for each subtype specie 
fig_fc_qHC_abund <- ggbarplot(HC_abund_fc, 
                              x ="cat", 
                              y = "fc", 
                              fill = "status", 
                              position = position_dodge(), 
                              palette = c(PD_enriched_col, PD_depleted_col), 
                              title = "Abundance Fold Change Q1 - Q4") + 
  theme_classic() + xlab("") + ylab("") + 
  scale_x_discrete(labels=c("Coherent" = "Coherent\nspecies", 
                            "Non-coherent" = "Noncoherent\nspecies")) + 
  expand_limits(y = c(0,4)) + custom_theme

fig_fc_qHC_abund
```


```{r, fig.width=8, fig.height=6, fig.show='hide'}
# BARPLOT PROPORTION QUARTILE - HC

# Create a list of vectors containing proportions computed in HC individuals for each subtype 
HC_prop <- list(
  CE = list(data = grp_CE_prop$HC, type = "E"),
  CD = list(data = grp_CD_prop$HC, type = "D"),
  NE = list(data = grp_NE_prop$HC, type = "E"),
  ND = list(data = grp_ND_prop$HC, type = "D")
)

# Apply "get_quartile" function to each vectors from "HC_prop"
# The function sort proportions in ascending way when it's enriched specie type and in decreasing way when it's depleted specie type  
# Then it split into quartile groups and compute mean of proportion for each quartile 
qHC_prop <- lapply(HC_prop, function(g){ get_quartile(g$data, g$type)})

# List of dataframe of that contains the mean proportions of each quartile for each subtype specie 
df_list_qHC_prop <- lapply(qHC_prop, function(g) { compute_mean(g)})


# Create list of proportion barplot for each specie type 
bp_list_qHC_prop <- list()
for (i in 1:length(df_list_qHC_prop)){
  bp_name <- name_sub_group[i]
  bp_color <- ifelse(str_detect(bp_name, ".*enriched.*"), PD_enriched_col, PD_depleted_col)
  # barplot 
  bp <- ggbarplot(df_list_qHC_prop[[i]], x = "variable", y = "value", fill = bp_color, title = bp_name, ylab = "", xlab = "") + custom_theme
  bp_list_qHC_prop[[i]] <- bp
}

# Arrange barplot
fig_qHC_prop <- ggarrange(plotlist = bp_list_qHC_prop)
fig_qHC_prop <- annotate_figure(fig_qHC_prop, left = text_grob("Proportion (%)", color = "black", rot = 90, size = 15)) 
fig_qHC_prop
```

```{r}
# PROPORTION FC QUARTILE Q1 Q4 - HC

# Compute fold change (FC) of mean proportion between Q1 and Q4 for each subtype specie 
# and create dataframe of FC of mean proportion Q1-Q4 for each subtype specie 
HC_prop_fc <- rbind(
  process_fc(df_list_qHC_prop$CE, "Coherent", "Enriched species"),
  process_fc(df_list_qHC_prop$CD, "Coherent", "Depleted species"),
  process_fc(df_list_qHC_prop$NE, "Non-coherent", "Enriched species"),
  process_fc(df_list_qHC_prop$ND, "Non-coherent", "Depleted species"))

HC_prop_fc <- HC_prop_fc %>% 
  mutate(status = factor(status, levels = c("Enriched species", "Depleted species")))  

# Create barplot of FC of mean proportion Q1-Q4 for each subtype specie 
fig_fc_qHC_prop <- ggbarplot(HC_prop_fc, 
                             x ="cat", 
                             y = "fc", 
                             fill = "status", 
                             position = position_dodge(), 
                             palette = c(PD_enriched_col, PD_depleted_col), 
                             title = "Proportion Fold Change Q1 - Q4") + 
  theme_classic() + xlab("") + ylab("") + 
  scale_x_discrete(labels=c("Coherent" = "Coherent\nspecies", 
                          "Non-coherent" = "Noncoherent\nspecies")) + 
  expand_limits(y = c(0,4)) + custom_theme

fig_fc_qHC_prop
```


```{r, fig.height=11, fig.width = 8, fig.show='hide'}
# BARPLOT PROP AND ABUND QUARTILE - HC
fig_qHC_left <- ggarrange(fig_qHC_abund, fig_qHC_prop, nrow = 2, labels = c("A","B"), font.label = list(size = 17))
fig_qHC_left
```


```{r, fig.height=11, fig.width = 8, fig.show='hide'}
# BARPLOT ABUND/PROP FC QUARTILE Q1 Q4 - HC
fig_qHC_right <- ggarrange(fig_fc_qHC_abund, fig_fc_qHC_prop, nrow = 2, labels = c("C","D"), font.label = list(size = 17))
fig_qHC_right
``` 


```{r, fig.height=9, fig.width = 13, fig.show='asis'}
# MERGED ABUND/PROP QUARTILE AND ABUND/PROP FC QUARTILE - HC
fig_qHC_top <- ggarrange(fig_qHC_left, fig_qHC_right, ncol = 2, widths = c(2, 0.9))
fig_qHC_top 
``` 


# Extended Figure 6 

```{r}
testR_msp.sign_HC_GBA_NMC_ht <- testR_msp.sign_HC_GBA_NMC
testR_msp.sign_HC_PD_ht <- testR_msp_HC_PD %>%
                              filter(q < 0.1)


MSP_set$msp_id_name <- paste(MSP_set$msp_id, MSP_set$species)

variable_select_clinic <- var_tab[!var_tab %in% c("Fruit", "Vegetables", "sleepiness",           "appendectomy","appendectomy_pd_yearsdiff","sex","education_n_years","ed_yrs" )] 


groups_select <- list(c("HC", "GBA_NMC"),
                c("HC", "PD"))

list_correlation_variable <- list()
list_plot_cor <- list()
list_legend_cor <- list()
list_merge <- list()
leg_cor <- list()
leg_gp <- list()

for (gps in groups_select){
  
  if (gps[1] == "HC" & gps[2] == "GBA_NMC"){
     size_lgd_gp = gpar(fontsize = 17)
     size_label_gp = gpar(fontsize =15)
     rownames_size = 15
     colnames_size = 15
     leg_cor[[1]] <- leg_gp[[1]] <- 15
     leg_cor[[2]] <- leg_gp[[2]] <- 17
     leg_gp[[3]] <- leg_gp[[4]] <- 0.5
     
  } else if (gps[1] == "HC" & gps[2] == "PD"){
     size_lgd_gp = gpar(fontsize = 22)
     size_label_gp = gpar(fontsize = 20)
     rownames_size = 20
     colnames_size = 25
     leg_cor[[1]] <- leg_gp[[1]] <- 25
     leg_cor[[2]] <- leg_gp[[2]] <- 27
     leg_gp[[3]] <- leg_gp[[4]] <- 0.8
  }
  
  names_file_cor <- paste("correlation", gps[1], gps[2], sep="_")
  
  if ("PD" %in% gps){
    select_data <- temp_metadata %>%
      filter(group_bis %in% c(gps[1], gps[2]))
  } else {
    select_data <- temp_metadata %>%
      filter(group_bis %in% c(gps[1], gps[2]))
  }
  
  name_to_select <- paste("sign", gps[1], gps[2], "ht",sep= "_")
  
  
  ### for file with msp 
  df_testR_msp <- get(paste("testR_msp", name_to_select, sep = "."))
  
  row_ha <-  HeatmapAnnotation(
        status = df_testR_msp[[13]],
        col = list(status = unlist(group_bis_color[c(gps[1], gps[2])])),
        which = "row",
        show_annotation_name = F,
        show_legend = FALSE
       )


  
  assign("df_testR.sign_msp", df_testR_msp)
  
  select_msp_data <- msp_fecal.filt10[rownames(df_testR.sign_msp), rownames(select_data)]
  select_msp_data.t <- t(select_msp_data)
  all(rownames(select_msp_data.t) == rownames(select_data))
 
  
  merged_data <- cbind(select_msp_data.t, select_data[match(rownames(select_msp_data.t),  rownames(select_data)), variable_select_clinic])
  

  list_correlation_variable[[names_file_cor]] <- list()
      

  res_cor <- rcorr(as.matrix(merged_data), type = "spearman")
        
  columns <- colnames(res_cor$r)[colnames(res_cor$r) %in% variable_select_clinic]
  rows <- rownames(res_cor$r)[rownames(res_cor$r) %in% unique(MSP_set$msp_id)]
        
  mat_cor <- res_cor$r[rows, columns, drop = FALSE]
  
  mat_corP <- res_cor$P[rows, columns, drop = FALSE]
  
  for (name_df in c("mat_cor", "mat_corP")){
    
    df <- get(name_df)
    
    if (sum(is.na(df)) >= nrow(df)){
    
      na_columns <- apply(df, 2, function(col) all(is.na(col)))
      df <- df[, !na_columns]
      
      assign(name_df, df)

    }
    
  }
        
        
  if (all(colnames(mat_cor) %in% variable_select_clinic)){
    
    colnames(mat_cor) <-  gsub("_", " ", colnames(mat_cor))
    colnames(mat_corP) <- gsub("_", " ", colnames(mat_corP))
  }
        

  if (all(rownames(mat_cor) %in%  MSP_set$msp_id)) {
    
    MSP_set$msp_id_name <- sapply(strsplit(MSP_set$msp_id_name, " sp\\."), `[`, 1)


    replacement <- setNames(MSP_set$msp_id_name, MSP_set$msp_id)
    
    current_rownames <- rownames(mat_cor)
    new_rownames <- replacement[current_rownames]
    
    rownames(mat_cor) <- new_rownames
    rownames(mat_corP) <- new_rownames
  }
        
  mat_cor2 <- mat_cor %>%
    as.data.frame(.) %>%
    rownames_to_column("clinic") %>%
    pivot_longer(-clinic, names_to = "msp", values_to = "val_cor")
  
  mat_corP2 <- mat_corP %>%
    as.data.frame(.) %>%
    rownames_to_column("clinic") %>%
    pivot_longer(-clinic, names_to = "msp", values_to = "val_pval")
  
  combined_mat_cor <- mat_cor2 %>%
    inner_join(mat_corP2, by = c("clinic", "msp")) 
  
  list_correlation_variable[[names_file_cor]][["ht_clinic_varibale"]] <- combined_mat_cor
        
  assign("ht_clinic_variable", heatmap_plot(mat_cor, mat_corP, lgd_title = "Correlation values", size_lgd_gp, size_label_gp,  rownames_size, colnames_size, annot = row_ha))
        
  
        
  lgd_cor = Legend(col_fun = colorRamp2(c(-1, 0, 1), c("#D60C00FF", "white","#072AC8" )),
                   title = "Correlation values",
                   title_position = "topcenter",
                   direction = "horizontal",
                   labels_gp = gpar(col = "black", fontsize = leg_cor[[1]]),
                   title_gp = gpar(col = "black", fontsize = leg_cor[[2]]),
                   border = "black")

  lgd_group = Legend(labels = c(gps[1], gps[2]),
                     legend_gp = gpar(fill=c(group_bis_color[gps[1]],group_bis_color[gps[2]])),
                     title = "Groups",
                     title_position = "topcenter",
                     nrow = 1,
                     labels_gp = gpar(col = "black", fontsize = leg_gp[[1]]),
                     title_gp = gpar(col = "black", fontsize = leg_gp[[2]]),
                     grid_height = unit(leg_gp[[3]], "cm"), grid_width = unit(leg_gp[[4]], "cm"))

  pd = packLegend(lgd_cor, lgd_group,
                  direction = "horizontal",
                  column_gap = unit(1.5, "cm"),
                  row_gap = unit(3, "cm"))


  name_ht <- paste("ht", gps[1], gps[2], sep="_")
  legend_ht <- paste("leg", gps[1], gps[2], sep="_")
  
  list_plot_cor[[name_ht]] <- list()
  list_plot_cor[[legend_ht]] <- pd
  
  list_plot_cor[[name_ht]][["ht_clinic_variable"]] <- ht_clinic_variable
  
}
```

```{r,fig.height= 25, fig.width= 28, fig.show='asis'}
draw(list_plot_cor$ht_HC_GBA_NMC[[1]],   
     annotation_legend_list = list_plot_cor$leg_HC_GBA_NMC, annotation_legend_side = "top", padding = unit(c(2, 1, 1, 9), "cm"))
grid.text("B", 
          x = unit(0.01, "npc"),  
          y = unit(0.99, "npc"), 
          gp = gpar(fontsize = 20, fontface = "bold"))
```


```{r, fig.height= 40, fig.width= 30, fig.show='asis'}
draw(list_plot_cor$ht_HC_PD[[1]], 
     annotation_legend_list = list_plot_cor$leg_HC_PD, 
     annotation_legend_side = "top",
     padding = unit(c(8, 0, 1, 25), "cm"))
grid.text("A", 
          x = unit(0.01, "npc"),  
          y = unit(0.993, "npc"), 
          gp = gpar(fontsize = 40, fontface = "bold"))
``` 
 
# Extended Figure 7
```{r}
# CORRELATIONS CLINICAL VS MSP in groups

# list of variable for PD
select_clinic_var <- c("moca_total", "upsit_total", "DQS_score", "updrs_i_total", "updrs_ii_total", "updrs_iii_total", "updrs_total", "scopa_cardiovascular", "scopa_gastrointestinal", "scopa_pupillomotor", "scopa_sexual", "scopa_thermoregulatory", "scopa_urinary", "scopa_total", "rbdsq_total", "bdi_total", "hads_anxiety_total", "hads_depression_total", "wcs_total", "ledd_total", "pd_dur", "age", "bmi", "hoehn_yahr")

# variable selection for GBA_NMC and HC 
select_clinic_var_bis <- select_clinic_var[!select_clinic_var %in% c("ledd_total", "pd_dur", "hoehn_yahr")] # keep age and bmi
```


```{r}
## PD 
# The data used are correlation of species with clinical parameters in PD patients (n = 271).

# extract the clinic parameters and the 176 MSP abundance for 271 PD individuals 
PD_data <- temp_metadata %>% filter(group_bis == "PD") %>% dplyr::select(all_of(select_clinic_var))
PD_176abund <- msp_fecal.filt10[c(CE, NE, CD, ND), rownames(PD_data)]
PD_176abund_t <- as.data.frame(t(PD_176abund))
# dim(PD_176abund_t)
# dim(PD_data)

# checking
all(rownames(PD_data) == rownames(PD_176abund_t))

# merge clinic parameters and abundance of the 176 MSP
merged_PD <- cbind(PD_176abund_t, PD_data)

# compute spearman correlation
cor_PD <- rcorr(as.matrix(merged_PD), type = "spearman") 
mat_cor <- cor_PD$r
mat_corP <- cor_PD$P

# correlation and pvalue values between MSP abundance and clinical variables
merge_mat_cor_PD <- get_corr_and_pval(cor_PD, PD_176abund)
colnames(merge_mat_cor_PD)
merge_mat_cor_PD
```

 
```{r, warning=F}
## GBA_NMC 
# The data used are correlation of species with clinical parameters in GBA_NMC patients (n = 43).  

# extract the clinical parameters and the 176 MSP abundance for 43 GBA_NMC individuals (except : pd_dur and ledd_total)
NMC_data <- temp_metadata %>% filter(group_bis == "GBA_NMC") %>% dplyr::select(all_of(select_clinic_var_bis))
NMC_176abund <- msp_fecal.filt10[c(CE,NE,CD,ND), rownames(NMC_data)]
NMC_176abund_t <- as.data.frame(t(NMC_176abund))
all(rownames(NMC_176abund_t) == rownames(NMC_data))
# dim(NMC_176abund_t)
# dim(NMC_data)

# merge clinical parameters and abundance of the 173 MSP
merged_NMC <- cbind(NMC_176abund_t, NMC_data)

# compute spearman correlations
cor_NMC <- rcorr(as.matrix(merged_NMC), type = "spearman") 
mat_cor <- cor_NMC$r
mat_corP <- cor_NMC$P

# correlation and pvalue values between MSP abundance and clinical variables
merge_mat_cor_NMC <- get_corr_and_pval(cor_NMC, NMC_176abund)
```

 
```{r, warning=F}
## HC 
# The data used are correlation of species with clinical parameters in HC patients (n = 150).  

# extract the clinic parameters and the 176 MSP abundance for 150 HC individuals 
HC_data <- temp_metadata %>% filter(group_bis == "HC") %>% dplyr::select(all_of(select_clinic_var_bis))
HC_176abund <- msp_fecal.filt10[c(CE, NE, CD, ND), rownames(HC_data)]
HC_176abund_t <- as.data.frame(t(HC_176abund))
all(rownames(HC_176abund_t) == rownames(HC_data))

dim(HC_176abund_t)
dim(HC_data)

# merge clinic parameters and abundance of the 176 MSP
merged_HC <- cbind(HC_176abund_t, HC_data)
cor_HC <- rcorr(as.matrix(merged_HC), type = "spearman") 
mat_cor <- cor_HC$r
mat_corP <- cor_HC$P

# correlation and pvalue values between MSP abundance and clinical variables
merge_mat_cor_HC <- get_corr_and_pval(cor_HC, HC_176abund)
```




```{r}
# CORRELATION PD vs GBA_NMC

# merge correlation matrix of PD and GBA_NMC and 
# extract significant correlation between clinical and MSP
merged_corr_PD_NMC <- merge(merge_mat_cor_PD, merge_mat_cor_NMC, by = c("msp_id", "variable"))
merged_corr_PD_NMC <- merged_corr_PD_NMC %>% filter(p_corr.x <0.05 & p_corr.y <0.05)

# create new column to indicate if correlations are all positive or negative
merged_corr_PD_NMC <- merged_corr_PD_NMC %>% mutate(same_corr_dir = ifelse(corr_dir.x == corr_dir.y, "yes", "no")) 
merged_corr_PD_NMC %>% filter(same_corr_dir == "yes") 

# checking
dim(merged_corr_PD_NMC)
merged_corr_PD_NMC %>% pull(msp_id) %>% unique() %>% length()
merged_corr_PD_NMC %>% filter(same_corr_dir == "yes") 
merged_corr_PD_NMC %>% filter(same_corr_dir == "yes") %>% pull(msp_id) %>% unique() %>% length()
merged_corr_PD_NMC %>% filter(variable == "hoehn_yahr")
merged_corr_PD_NMC %>% filter(same_corr_dir == "no")
```



```{r}
# regression plot - PD and NMC 
formula <- y ~ x
reg_corr_NMC_PD <- ggplot(data = merged_corr_PD_NMC, aes(x = corr.x, y = corr.y)) + 
  geom_point(aes(color= same_corr_dir), size = 3) + 
  geom_smooth(method = "lm", se = F, formula = formula, linetype = 3, linewidth=0.8, color = "black") + 
  #stat_poly_eq(use_label(c("R2")), label.x = 0.1, label.y = 0.9, size = 7) +
  stat_correlation(method = "spearman", label.x = 0.1, label.y = 0.9, size = 7) +
  xlab("Spearman correlation in PD") + 
  ylab("Spearman correlation in GBA_NMC") + 
  theme_classic() +
  theme(legend.position = 'none', axis.title = element_text(size = 16), axis.text = element_text(size = 16)) 

reg_corr_NMC_PD
```




```{r}
# CORRELATION HC vs PD 

# merge correlation matrix of PD and HC and 
# extract significant correlation between clinical and MSP
merged_corr_PD_HC <- merge(merge_mat_cor_PD, merge_mat_cor_HC, by = c("msp_id", "variable")) 
merged_corr_PD_HC <- merged_corr_PD_HC %>% filter(p_corr.x <0.05 & p_corr.y <0.05)

# create new column to indicate if correlations are all positive or negative
merged_corr_PD_HC <- merged_corr_PD_HC %>% mutate(same_corr_dir = ifelse(corr_dir.x == corr_dir.y, "yes", "no")) 

# check
dim(merged_corr_PD_HC)
merged_corr_PD_HC %>% pull(msp_id) %>% unique() %>% length()
merged_corr_PD_HC %>% filter(same_corr_dir == "yes") 
merged_corr_PD_HC %>% filter(same_corr_dir == "yes") %>% pull(msp_id) %>% unique() %>% length()
```




```{r}
# regression plot
formula <- y ~ x
reg_corr_HC_PD <- ggplot(data = merged_corr_PD_HC, aes(x = corr.x, y = corr.y)) + 
  geom_point(aes(color= same_corr_dir), size = 3) + 
  geom_smooth(method = "lm", se = F, formula = formula, linetype = 3, linewidth=0.8, color = "black") + 
  #stat_poly_eq(use_label(c("R2")), label.x=0.1, label.y = 0.9, size = 7) +
  stat_correlation(method = "spearman", label.x=0.1, label.y = 0.9, size = 7) +
  xlab("Spearman correlation in PD") + 
  ylab("Spearman correlation in HC") + 
  theme_classic() + 
  theme(legend.position = 'none', axis.title = element_text(size = 16), axis.text = element_text(size = 16)) 

reg_corr_HC_PD
```


```{r, fig.height=6, fig.width=14, fig.show='asis'}
# FINAL CORRELATION PD VS GBA_NMC/HC 
fig_corr <- ggarrange(reg_corr_NMC_PD, reg_corr_HC_PD, ncol = 2, labels = c("A", "B"), font.label = list(size = 17))
fig_corr
``` 
 

 
# Extended Figure 8

```{r}
pdms16 <- c("msp_1302","msp_0060","msp_0639","msp_0250","msp_0480","msp_0665","msp_0541","msp_1581","msp_1558","msp_0044","msp_0274","msp_0545","msp_0086","msp_0379","msp_0510","msp_0524")
pdms16_data <- HC_vs_PD_NMC.sign[pdms16,] %>% select(type)
pdms16_data <- pdms16_data %>% mutate(cat = ifelse(rownames(pdms16_data) %in% c(CE, NE), "PD_enriched", "PD_depleted")) 
```

```{r}
# extract the interest MSP abundance in HC and PD
abund_pdms_HC <- msp_fecal.filt10[rownames(pdms16_data), HC_ind]
abund_pdms_PD <- msp_fecal.filt10[rownames(pdms16_data), PD_ind]


msp_enriched <- pdms16_data %>% filter(cat == "PD_enriched")
msp_depleted <- pdms16_data %>% filter(cat == "PD_depleted")

HC_count_enriched <- colSums(abund_pdms_HC[rownames(msp_enriched),]>0)
HC_count_depleted <- colSums(abund_pdms_HC[rownames(msp_depleted),]>0)

PD_count_enriched <- colSums(abund_pdms_PD[rownames(msp_enriched),]>0)
PD_count_depleted <- colSums(abund_pdms_PD[rownames(msp_depleted),]>0)

# checking 
abund_pdms_HC %>% select(ASAPh_268) %>% filter(rownames(abund_pdms_HC) %in% c(CE, NE))
abund_pdms_HC %>% select(ASAPh_384) %>% filter(rownames(abund_pdms_HC) %in% c(CD, ND))

abund_pdms_PD %>% select(ASAPh_850) %>% filter(rownames(abund_pdms_PD) %in% c(CE, NE))
abund_pdms_PD %>% select(ASAPh_850) %>% filter(rownames(abund_pdms_PD) %in% c(CD, ND))
```



```{r, fig.show='asis'}
# checking
all(names(HC_count_enriched) == names(HC_count_depleted))
all(names(PD_count_enriched) == names(PD_count_depleted))

HC_pdms <- HC_count_enriched - HC_count_depleted
PD_pdms <- PD_count_enriched - PD_count_depleted

# define breaks 
breaks <- c(-Inf, -5, -2, 1, 4, 7, 10)
labels <- c("-7 to -5", "-4 to -2", "-1 to 1", "2 to 4", "5 to 7", "8 to 10")

# count occurence of PDMS score according to define breaks in HC and PD 
HC_pdms_occ <- table(cut(HC_pdms, breaks = breaks, labels = labels))
PD_pdms_occ <- table(cut(PD_pdms, breaks = breaks, labels = labels)) 

df_HC_pdms_occ <- as.data.frame(HC_pdms_occ) %>% mutate(group_bis = "HC")
df_PD_pdms_occ <- as.data.frame(PD_pdms_occ) %>% mutate(group_bis = "PD")
df_pdms_occ <- df_HC_pdms_occ %>% rbind(df_PD_pdms_occ)

# the barplot 
fig_pdms <- ggbarplot(df_pdms_occ, x = "Var1", y = "Freq", position = position_dodge(), xlab = "PDMS score", ylab = "Number of individuals", title = "PDMS 16 distribution", fill = "group_bis") + scale_fill_manual(values = group_bis_color[c(1,3)]) + theme(legend.position = "none", axis.title = element_text(size = 15, color = "black"), axis.text = element_text(size = 13, color = "black")) + scale_y_continuous(breaks = seq(0, 80, 10))

fig_pdms
```

 

# Extended Figure 9

```{r}
mod2 <- ftmt.nocore.fecal.baseline %>%
    t() %>%
    as.data.frame() 
    

mod_met2 <- mod2 %>%
              filter(rownames(.) %in% rownames(temp_metadata)) %>%
              merge(temp_metadata["group_bis"], by = "row.names") %>%
              column_to_rownames(var = "Row.names") 
```

```{r}
res_PD_HC <- testRelations2(x = mod_met2 %>%
                                   filter(group_bis %in% c("HC", "PD")) %>%
                                   select(-group_bis) %>%
                                   t(), 
                            y = mod_met2 %>%
                                 filter(group_bis %in% c("HC", "PD")) %>%
                                 mutate(group_bis = factor(group_bis, levels = c("HC", "PD"))) %>%
                                 pull(group_bis), 
                            type = "nonparametric", 
                            multiple.adjust = "BH")

res_GBA_NMC_HC <- testRelations2(x = t(mod_met2 %>%
                                   filter(group_bis %in% c("HC", "GBA_NMC")) %>%
                                   select(-group_bis)), 
                            y = mod_met2 %>%
                                 filter(group_bis %in% c("HC", "GBA_NMC")) %>%
                                 mutate(group_bis = factor(group_bis, levels = c("GBA_NMC","HC"))) %>%
                                 pull(group_bis), 
                            type = "nonparametric", 
                            multiple.adjust = "BH")
```

```{r}
modul_subgroup_filter <- module_subgroup %>%
                    filter(subgroup %in% c("CEM", "CDM"))
```


```{r}
res_PD_HC <- res_PD_HC[rownames(res_PD_HC) %in% modul_subgroup_filter$module,]
```

```{r}
res_GBA_NMC_HC <- res_GBA_NMC_HC[rownames(res_GBA_NMC_HC) %in% modul_subgroup_filter$module,]
```


```{r}
for (df in c("res_PD_HC", "res_GBA_NMC_HC")){
  data <- get(df)
  nm <- paste("cliff", gsub("res_", "", df), sep = "_")
  assign(nm, data[12])
  
}
df_cliff <- cbind(cliff_GBA_NMC_HC, cliff_PD_HC)
df_cliff$HC.PD_Cliffdelta <- df_cliff$HC.PD_Cliffdelta * -1
```


```{r include = FALSE}
# Compute Spearman correlation 
rho <- cor(df_cliff[,"HC.PD_Cliffdelta"], df_cliff[,"GBA_NMC.HC_Cliffdelta"], method = "spearman") 

# Construct plots 
p <- ggplot(data = df_cliff, aes(x = HC.PD_Cliffdelta, y = GBA_NMC.HC_Cliffdelta, group = NULL)) +
          geom_point(color = "black", size = 4, shape = 19) +
          theme_classic() +
          geom_smooth(method = "lm",  se = FALSE, color = "black", linetype = 3, linewidth = 1) +
          stat_poly_eq(aes(label = paste(..eq.label.., paste0("rho==", signif(rho, 2)), sep = "~~~")), 
                       vjust = 0.5, size = 7, output.type = "expression", parse  = TRUE) +
          labs(title = "A", x = "CD HC vs PD", y = "CD HC vs GBA_NMC") +
          ylim(-0.2, 0.3)+
          theme(plot.title = element_text(hjust = -0.17,
                              size = 21,
                              vjust = 3,
                              face = "bold"),
                legend.position = "none",
                axis.title = element_text(size = 21, color = "black"),
                axis.text = element_text(size = 19, color = "black"),
                axis.title.x = element_text(vjust = -3),
                axis.title.y = element_text(vjust = 4),
                plot.margin = unit(c(0.7,1,1,1.1), "cm")
              )
```

```{r include = FALSE}
df_cliff_barplot <- df_cliff %>%
                      pivot_longer(cols = everything(),
                                   names_to = "Groups", 
                                   values_to = "cliffdelta") %>%
                      group_by(Groups) %>%
                      dplyr::summarize("mean_cliffdelta" = mean(abs(cliffdelta))) %>%
                      mutate(Groups = factor(Groups, levels = c("HC.PD_Cliffdelta", "GBA_NMC.HC_Cliffdelta")))
                      


pbar <- ggbarplot(
  data = df_cliff_barplot,
  x = "Groups",
  y = "mean_cliffdelta",
  position = position_dodge(),
  color = "black",
  fill = "Groups"
) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 0.24)) +
  scale_fill_manual(values = c("HC.PD_Cliffdelta"="#3d3d3d",
                               "GBA_NMC.HC_Cliffdelta"="#737373")) +
  theme(
    plot.title = element_text(hjust = -0.22,
                              size = 21,
                              vjust = 3,
                              face = "bold"),
    legend.position = "none",
    legend.title = element_blank(),
    axis.text.x = element_text(vjust = -1),
    axis.title.x = element_text(vjust = -3),
    axis.title.y = element_text(vjust = 5),
    axis.title = element_text(size = 21, color = "black"),
    axis.text = element_text(size = 19, color = "black"),
    plot.margin = unit(c(0.7, 0.5, 1, 1.5), "cm")#, 
  ) +
  annotate("text", x = 1.5, y = 0.23, label = "****", size = 7) +
  labs(title = "B", x = "Comparisons", y = "CD absolute values") +
  scale_x_discrete(labels = c("HC vs PD", "HC vs GBA_NMC"))

```

```{r, fig.height= 7, fig.width= 17, fig.show='asis'}
exfig9 <- grid.arrange(p, pbar, nrow = 1, 
             widths= c(1.1, 0.9))
exfig9
```


# Extended Figure 10
```{r}
# USA cohort
load("./data/verified_metadata_Wallen_PRJNA834801.RData") # metadata
load("./verified_merge_msp_data_Wallen_PRJNA834801.RData") # MSP abundance matrix 
dim(metadata.wallen.PRJNA834801) # 725 ind 
table(metadata.wallen.PRJNA834801$group)
dim(msp.merged.wallen.PRJNA834801)
```


```{r}
# Korea cohort 
load("./data/verified_metadata_Jo_PRJNA743718.RData") # metadata
load("./data/verified_merge_msp_data_Jo_PRJNA743718.RData") # MSP abundance matrix 
dim(metadata.jo.PRJNA743718) # 146 ind
table(metadata.jo.PRJNA743718$group)
dim(msp.merged.jo.PRJNA743718)
```


```{r}
# Turkish cohort 
load("./data/verified.metadata.clasen.PRJEB79944.RData") # metadata
load("./data/verified.msp.merged.clasen.PRJEB79944.RData") # MSP abundance matrix
# removed 28 problematic samples 
verified.metadata.clasen.PRJEB79944 <- verified.metadata.clasen.PRJEB79944 %>% 
    filter(is.na(excluded)) 
all(verified.metadata.clasen.PRJEB79944$sampleID %in% colnames(verified.msp.merged.clasen.PRJEB79944))
verified.metadata.clasen.PRJEB79944 <- verified.metadata.clasen.PRJEB79944 %>% 
    mutate(group = ifelse(health_status == "H", "HC", "PD"))
rownames(verified.metadata.clasen.PRJEB79944) <- verified.metadata.clasen.PRJEB79944$sampleID 

# Transform into numeric for bmi 
verified.metadata.clasen.PRJEB79944$bmi <- as.numeric(str_replace_all(verified.metadata.clasen.PRJEB79944$bmi, ",", "."))

dim(verified.metadata.clasen.PRJEB79944)# 106 ind 
table(verified.metadata.clasen.PRJEB79944$group)
dim(verified.msp.merged.clasen.PRJEB79944) 


# 86 individuals selected to make it comparable with the analysis SupTab3 in paper : Microbiome signatures of virulence in the oral-gut-brain axis influence Parkinson’s disease and cognitive decline pathophysiology
select_ind_tur <- c("ERS24550623","ERS24550783","ERS24550733","ERS24550822","ERS24550677","ERS24550620","ERS24550834","ERS24550716","ERS24550838","ERS24550649","ERS24550635","ERS24550713","ERS24550714","ERS24550656","ERS24550745","ERS24550763","ERS24550660","ERS24550812","ERS24550687","ERS24550786","ERS24550755","ERS24550645","ERS24550781","ERS24550670","ERS24550818","ERS24550712","ERS24550739","ERS24550663","ERS24550810","ERS24550743","ERS24550790","ERS24550720","ERS24550738","ERS24550776","ERS24550762","ERS24550825","ERS24550605","ERS24550764","ERS24550779","ERS24550731","ERS24550841","ERS24550803","ERS24550768","ERS24550629","ERS24550689","ERS24550644","ERS24550646","ERS24550811","ERS24550664","ERS24550766","ERS24550741","ERS24550785","ERS24550816","ERS24550659","ERS24550828","ERS24550626","ERS24550732","ERS24550791","ERS24550792","ERS24550711","ERS24550614","ERS24550752","ERS24550774","ERS24550747","ERS24550616","ERS24550831","ERS24550715","ERS24550665","ERS24550651","ERS24550734","ERS24550770","ERS24550777","ERS24550754","ERS24550612","ERS24550699","ERS24550782","ERS24550718","ERS24550634","ERS24550693","ERS24550685","ERS24550725","ERS24550830","ERS24550681","ERS24550618","ERS24550820","ERS24550666")


# checking 
all(select_ind_tur %in% verified.metadata.clasen.PRJEB79944$sampleID)
all(select_ind_tur %in% colnames(verified.msp.merged.clasen.PRJEB79944))

# Select the 86 individuals on metadata 
verified.metadata.clasen.PRJEB79944 <- verified.metadata.clasen.PRJEB79944 %>% filter(sampleID %in% select_ind_tur)
dim(verified.metadata.clasen.PRJEB79944) # 86 ind 
table(verified.metadata.clasen.PRJEB79944$group) # 86 ind : 17 HC and 69 PD

# Select the 86 individuals on msp abundance 
verified.msp.merged.clasen.PRJEB79944 <- verified.msp.merged.clasen.PRJEB79944[, verified.metadata.clasen.PRJEB79944$sampleID]
all(colnames(verified.msp.merged.clasen.PRJEB79944) == verified.metadata.clasen.PRJEB79944$sampleID)
```

```{r}
# Clinical data from Dusko - received in 24/10/25 
clinic_data_tur <- openxlsx::read.xlsx("./data/Turkish_MSP_and_metadata.xlsx")
index <- which(colnames(clinic_data_tur) == "mgp_sample_alias")
clinic_data_tur <- clinic_data_tur[, index:ncol(clinic_data_tur)]
clinic_data_tur <- clinic_data_tur %>% filter(!is.na(mgp_sample_alias))
dim(clinic_data_tur) # 86 ind 

# Replace missing values by real NA 
missing_values <- c("Missing_Data", "NA", "Not_Applicable", "") 

clinic_data_tur[] <- lapply(clinic_data_tur, function(col) {
  if (is.character(col)) {
    col[col %in% missing_values] <- NA
  }
  return(col)
}) 


# Make sure numeric values recognized as numeric variables  
is_numeric_column <- function(x) {
  all(is.na(x) | grepl("^-?\\d+(\\.\\d+)?$", x))
} 
numeric_cols <- sapply(clinic_data_tur, is_numeric_column) # the detected numerical variables 

# convert of numeric column into numeric variable type  
clinic_data_tur[numeric_cols] <- lapply(clinic_data_tur[numeric_cols], as.numeric)
```

```{r}
# List of clinical variables
var_tab_tur <- c("Disease_Duration", "NPI_Frequency-Severity", "WMS-VR_Short-Term_Memory", "ST_False_Answer", "GDS", "RBD", "UMPDHQ", "WMS-VR_Long-Term_Memory", "Benton_Line_Orientation_Test", "CDRS", "MMSE", "UPDRS", "Benton_Face_Recognition_Test", "Categorical_Fluency_", "Clock_Drawing_Test", "Phonemic_Fluency", "HYE", "Apathy", "DS_Backward", "DS_Forward", "OVMPT_Coding", "OVMPT_Recalling", "OVMPT_Recognition", "ST_Time_Difference")

# Select the clinical variables for merging
clinic_data_tur <- clinic_data_tur[, c("mgp_sample_alias", var_tab_tur)]

# Merge sample information and clinical information 
verified.metadata.clasen.PRJEB79944 <- merge(verified.metadata.clasen.PRJEB79944, clinic_data_tur, by = "mgp_sample_alias") 
rownames(verified.metadata.clasen.PRJEB79944) <- verified.metadata.clasen.PRJEB79944$sampleID 

# Same order of samples in metadata and abundance matrix 
verified.msp.merged.clasen.PRJEB79944 <- verified.msp.merged.clasen.PRJEB79944[, rownames(verified.metadata.clasen.PRJEB79944)]

all(colnames(verified.msp.merged.clasen.PRJEB79944) == rownames(verified.metadata.clasen.PRJEB79944)) # T 

# Specific character like "-" in column variable names caused errors, we replace them by "_" in metadata 
new_var_tab_tur <- str_replace_all(var_tab_tur, "-", "_") 

metadata_colnames <- colnames(verified.metadata.clasen.PRJEB79944)
metadata_colnames[match(var_tab_tur, metadata_colnames)] <- new_var_tab_tur
names(verified.metadata.clasen.PRJEB79944) <- metadata_colnames 
```



```{r}
# Filter MSP abundance with prevalence > 10% 
# USA cohort
msp.merged.wallen.filt10 <- msp.merged.wallen.PRJNA834801[rowSums(msp.merged.wallen.PRJNA834801>0)>(ncol(msp.merged.wallen.PRJNA834801)*10/100),] #498 msp left
all(colnames(msp.merged.wallen.PRJNA834801) == metadata.wallen.PRJNA834801$sampleID) # TRUE 

# Korea cohort 
msp.merged.jo.filt10 <- msp.merged.jo.PRJNA743718[rowSums(msp.merged.jo.PRJNA743718>0)>(ncol(msp.merged.jo.PRJNA743718)*10/100),] 
dim(msp.merged.jo.filt10) # 572 MSP left 
all(colnames(msp.merged.jo.PRJNA743718) == metadata.jo.PRJNA743718$sampleID) # TRUE 

# Turkish cohort 
verified.msp.merged.clasen.PRJEB79944.filt10 <- verified.msp.merged.clasen.PRJEB79944[rowSums(verified.msp.merged.clasen.PRJEB79944>0)>(ncol(verified.msp.merged.clasen.PRJEB79944)*10/100),] 
dim(verified.msp.merged.clasen.PRJEB79944.filt10) # 606 MSP left 
all(colnames(verified.msp.merged.clasen.PRJEB79944.filt10) == verified.metadata.clasen.PRJEB79944$sampleID) # TRUE
```

```{r}
# HC vs PD comparison outputs 
# ASAP 
testR_asap_HC_PD <- testR_msp_HC_PD %>% invert_CD_direction()

# USA 
testR_wallen_HC_PD <- testRelations2(x = msp.merged.wallen.filt10, y = metadata.wallen.PRJNA834801$group, type = "nonparametric", paired= F, multiple.adjust = "BH")
testR_wallen_HC_PD <- testR_wallen_HC_PD %>% invert_CD_direction()

# KOR
testR_jo_HC_PD <- testRelations2(x = msp.merged.jo.filt10, y = metadata.jo.PRJNA743718$group, type = "nonparametric", paired= F, multiple.adjust = "BH")
testR_jo_HC_PD <- testR_jo_HC_PD %>% invert_CD_direction()

# TUR 
testR_clasen_HC_PD <- testRelations2(x = verified.msp.merged.clasen.PRJEB79944.filt10, y = verified.metadata.clasen.PRJEB79944$group, type = "nonparametric", paired= F, multiple.adjust = "BH")
testR_clasen_HC_PD <- testR_clasen_HC_PD %>% invert_CD_direction()
```



```{r, fig.width=19, fig.height=14}
# Build Clifdelta's plots for ASAP between 3 public cohorts 
list_plot_asap_usa <- combine_CD_plots(testR_asap_HC_PD, testR_wallen_HC_PD, xlab = "CD - ASAP", ylab = "CD - USA")
list_plot_asap_kor <- combine_CD_plots(testR_asap_HC_PD, testR_jo_HC_PD, xlab = "CD - ASAP", ylab = "CD - KOREA")
list_plot_asap_tur <- combine_CD_plots(testR_asap_HC_PD, testR_clasen_HC_PD, xlab = "CD - ASAP", ylab = "CD - TURKIYE")

# Arrange plots
ext_usa <- ggarrange(list_plot_asap_usa$testR_merged.all, 
                     list_plot_asap_usa$testR_merged.sign,
                     list_plot_asap_usa$testR_merged.bothsign, 
                     ncol = 1, nrow = 3, labels = c("A", "D", "G"), font.label = list(size = 20))

ext_kor <- ggarrange(list_plot_asap_kor$testR_merged.all, 
                     list_plot_asap_kor$testR_merged.sign,
                     list_plot_asap_kor$testR_merged.bothsign, 
                     ncol = 1, nrow = 3, labels = c("B", "E", "H"), font.label = list(size = 20)) 

ext_tur <- ggarrange(list_plot_asap_tur$testR_merged.all, 
                     list_plot_asap_tur$testR_merged.sign,
                     list_plot_asap_tur$testR_merged.bothsign, 
                     ncol = 1, nrow = 3, labels = c("C", "F", "I"), font.label = list(size = 20)) 

ext_fig10 <- ggarrange(ext_usa, ext_kor, ext_tur, ncol = 3) 
ext_fig10
```

# Extended Figure 11 

```{r}
other_metadata <- c("metadata.wallen.PRJNA834801", "metadata.jo.PRJNA743718", "verified.metadata.clasen.PRJEB79944") 
other_abund_mat <- c("msp.merged.wallen.filt10", "msp.merged.jo.filt10", "verified.msp.merged.clasen.PRJEB79944.filt10")
other_cohort_name <- c("USA", "KOR", "TUR")

# Initialize lists 
list_qHC_abund <- list()
list_qHC_prop <- list()
list_qPD_abund <- list()
list_qPD_prop <- list()

# Compute abundance/proportion of coherent enriched/depleted species in HC and PD for each cohort
vec_group = c("HC", "PD")
for (i in seq_along(other_metadata)){
  metadata <- get(other_metadata[i])
  abund_mat <- get(other_abund_mat[i])
  
  # Get the detected species from CE and CD in given cohort
  CE_bis <- CE[CE %in% rownames(abund_mat)]
  CD_bis <- CD[CD %in% rownames(abund_mat)]
  
  # Compute abundance/proportion of CE/CD in HC/PD that are divided in quartile groups for each cohort
  CE_metrics <- compute_metrics(vec_group, metadata, abund_mat, CE_bis, "group", "sampleID", "E")
  CD_metrics <- compute_metrics(vec_group, metadata, abund_mat, CD_bis, "group", "sampleID", "D")
  
  # Stock the values in list for each cohort 
  cohort <- other_cohort_name[i]
  
  list_qHC_abund[[cohort]] <- list(CE = CE_metrics$HC_abund, CD = CD_metrics$HC_abund)
  list_qHC_prop[[cohort]]  <- list(CE = CE_metrics$HC_prop,  CD = CD_metrics$HC_prop)
  list_qPD_abund[[cohort]] <- list(CE = CE_metrics$PD_abund, CD = CD_metrics$PD_abund)
  list_qPD_prop[[cohort]]  <- list(CE = CE_metrics$PD_prop,  CD = CD_metrics$PD_prop)
}
```


```{r, fig.width=10, fig.height=15}
# From created lists containing the abund/prop of CE/CD in quartiles groups of HC/PD, create the barplots of quartiles  
list_bp_qHC_abund  <- generate_barplot_list(list_qHC_abund, "Abundance (%)")
list_bp_qHC_prop   <- generate_barplot_list(list_qHC_prop, "Proportion (%)")
list_bp_qPD_abund  <- generate_barplot_list(list_qPD_abund, "Abundance (%)")
list_bp_qPD_prop   <- generate_barplot_list(list_qPD_prop, "Proportion (%)")
```


```{r, fig.width=12, fig.height=15}
# Arrange the plots
bp_arranged_PD <- ggarrange(ggarrange(list_bp_qPD_abund$USA,  list_bp_qPD_prop$USA, labels = c("A"), ncol = 1), 
                            ggarrange(list_bp_qPD_abund$KOR, list_bp_qPD_prop$KOR, labels = c("C"), ncol = 1),  
                            ggarrange(list_bp_qPD_abund$TUR, list_bp_qPD_prop$TUR, labels = c("E"), ncol = 1), 
                            ncol = 1)

bp_arranged_HC <- ggarrange(ggarrange(list_bp_qHC_abund$USA,  list_bp_qHC_prop$USA, labels = c("B"), ncol = 1), 
                            ggarrange(list_bp_qHC_abund$KOR, list_bp_qHC_prop$KOR, labels = c("D"), ncol = 1),
                            ggarrange(list_bp_qHC_abund$TUR, list_bp_qHC_prop$TUR, labels = c("F"), ncol = 1),
                            ncol = 1)


bp_arranged_HC_PD <- ggarrange(bp_arranged_PD, bp_arranged_HC)
bp_arranged_HC_PD
```
