---
title: "ASAP Clinical Microbiome: 540 individuals - Nature 2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
  word_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File import, filter for patients to exclude
```{r}
library(reshape2)
library(ggplot2)
library(dplyr)
library(car)
library(tidyr)
library(ggpubr)
library(nycflights13)
library(DT)
library(forcats)
library(ggtext)
library(datawizard)
library(PMCMRplus)
library(fabricatr)
library(rstatix)
library(Hmisc)
library(factoextra)
library(devtools)
library(RColorBrewer)
library(broom)
library(emmeans)
library(questionr)
library(MASS)
```

```{r}
#Load metadata file and prepare final metadata file
metadata_559ind <- read.csv("/Users/elisamenozzi/Library/CloudStorage/OneDrive-Personal/Documents/GBA/ASAP BASELINE CLINICAL/Manuscript/Nature_submission/Final documents/Codes_metadata/ASAP_human_metadata_file_559ind_20250520_v10.csv", sep=",", dec = '.', stringsAsFactors = FALSE, header = TRUE)

dim(metadata_559ind)
table(metadata_559ind$cohort)
table(metadata_559ind$group)
```

```{r}
##keep only groups of interest
metadata_540ind = filter(metadata_559ind, group == "GBA_NMC" | group == "GBA_PD" |group == "HC" |group == "sPD")
dim(metadata_540ind)
table(metadata_540ind$group)
#51 GBA-NMC, 128 GBA-PD, 175 HC, 186 sPD
table(metadata_540ind$group, metadata_540ind$genotype)

table(metadata_540ind$cohort)
```


```{r}
# Metagenomics data
metadata_469ind <- read.csv("/Users/elisamenozzi/Library/CloudStorage/OneDrive-Personal/Documents/PhD Literature/Upgrade/ASAP MEATADATA/baseline_cohort/final_documents/metadata_full_cohort/clinical_metadata/00_complete_metadata/MGP_sampleID_469ind_fecal_baseline_Victoria.csv", sep=",", dec = '.', stringsAsFactors = FALSE, header = TRUE)
```

```{r}
metadata_469ind_complete = metadata_540ind[match(metadata_469ind$x, metadata_540ind$MGP_ID_fecal_baseline),]
dim(metadata_469ind_complete)
table(metadata_469ind_complete$group)
```


```{r}
##Combine first and second degree family history 
metadata_540ind$fhx2 = ifelse(metadata_540ind$fhx =="no", "no", "yes")
```

```{r}
metadata_540ind$three_groups = ifelse(metadata_540ind$group == "HC", "HC",
                                      ifelse(metadata_540ind$group == "GBA_NMC", "GBA_NMC", 
                                             "PD"))
table(metadata_540ind$three_groups)
```

```{r}
#Modify initial dataset as a few participants have been excluded, so partner status has changed
metadata_540ind$partner_updated = ifelse((metadata_540ind$clinical_ID == "ASAP_clinical_477" | metadata_540ind$clinical_ID == "ASAP_clinical_485" | metadata_540ind$clinical_ID == "ASAP_clinical_518" | metadata_540ind$clinical_ID == "ASAP_clinical_518"), "no", metadata_540ind$partner)

table(metadata_540ind$partner_updated)
```

```{r}
#Modify initial dataset as a few participants have been excluded, so family member status has changed
metadata_540ind$blood_related_updated =
  ifelse(
           (metadata_540ind$clinical_ID == "ASAP_clinical_425" | metadata_540ind$clinical_ID == "ASAP_clinical_441" | metadata_540ind$clinical_ID == "ASAP_clinical_442"), "yes", 
           
                ifelse(metadata_540ind$clinical_ID == "ASAP_clinical_495", "no",
                                                metadata_540ind$blood_related))

table(metadata_540ind$blood_related_updated)
```

```{r}
family = filter(metadata_540ind, cohort == "ITA")
family = filter(family, blood_related_updated == "yes")
family = family[,c(1:2, 15:20, 27:28)]
```

```{r}
metadata_nmc_pd = filter(metadata_540ind, three_groups == "GBA_NMC" | three_groups == "PD")
dim(metadata_nmc_pd)
```

```{r}
metadata_nmc_hc = filter(metadata_540ind, three_groups == "GBA_NMC" | three_groups == "HC")
dim(metadata_nmc_hc)
```

```{r}
metadata_hc_pd = filter(metadata_540ind, three_groups == "HC" | three_groups == "HC")
dim(metadata_hc_pd)
```

# RESULTS SUMMARY

```{r}
metadata_540ind = metadata_540ind %>% reorder_levels(three_groups, order = c("HC", "GBA_NMC", "PD"))

table_summ_540ind <- group_by(metadata_540ind, "group"=three_groups) %>% 
  summarise(

    mean_age = round(mean(age, na.rm=T),1), sd_age = round(sd(age, na.rm=T), 1),
    
    mean_bmi = round(mean(bmi, na.rm=T),1), sd_bmi = round(sd(bmi, na.rm=T), 1),
            
    mean_pd_onset = round(mean(pd_onset_calculated, na.rm=T),1), sd_pd_onset = round(sd(pd_onset_calculated, na.rm=T), 1),
            
    mean_pd_dur = round(mean(pd_dur, na.rm=T),1), sd_pd_dur = round(sd(pd_dur, na.rm=T), 1),
            
    mean_hy = round(median(hoehn_yahr, na.rm=T),1), sd_hy = round(sd(hoehn_yahr, na.rm=T), 1),
            
    mean_ledd = round(mean(ledd_total, na.rm=T),1), sd_ledd = round(sd(ledd_total, na.rm=T), 1),
            
    mean_updrs_i = round(mean(updrs_i_total, na.rm=T),1), sd_updrs_i = round(sd(updrs_i_total, na.rm=T), 1),
            
    mean_updrs_ii_total = round(mean(updrs_ii_total, na.rm=T),1), sd_updrs_ii_total = round(sd(updrs_ii_total, na.rm=T), 1),
            
    mean_updrs_iii_total = round(mean(updrs_iii_total, na.rm=T),1), sd_updrs_iii_total = round(sd(updrs_iii_total, na.rm=T), 1),
            
    mean_updrs_iv_total = round(mean(updrs_iv_total, na.rm=T),1), sd_updrs_iv_total = round(sd(updrs_iv_total, na.rm=T), 1),
            
    mean_updrs_total = round(mean(updrs_total, na.rm=T),1), sd_updrs_total = round(sd(updrs_total, na.rm=T), 1),
             
    mean_scopa_total = round(mean(scopa_total, na.rm=T),1), sd_scopa_total = round(sd(scopa_total, na.rm=T), 1),
       
    mean_scopa_gastrointestinal = round(mean(scopa_gastrointestinal, na.rm=T),1), sd_scopa_gastrointestinal = round(sd(scopa_gastrointestinal, na.rm=T), 1),
       
    mean_scopa_urinary = round(mean(scopa_urinary, na.rm=T),1), sd_scopa_urinary = round(sd(scopa_urinary, na.rm=T), 1),
       
    mean_scopa_cardiovascular = round(mean(scopa_cardiovascular, na.rm=T),1), sd_scopa_cardiovascular = round(sd(scopa_cardiovascular, na.rm=T), 1),
       
    mean_scopa_thermoregulatory = round(mean(scopa_thermoregulatory, na.rm=T),1), sd_scopa_thermoregulatory = round(sd(scopa_thermoregulatory, na.rm=T), 1),
       
    mean_scopa_pupillomotor = round(mean(scopa_pupillomotor, na.rm=T),1), sd_scopa_pupillomotor = round(sd(scopa_pupillomotor, na.rm=T), 1),
    
    mean_scopa_sexual = round(mean(scopa_sexual, na.rm=T),1), sd_scopa_sexual = round(sd(scopa_sexual, na.rm=T), 1),
    
    mean_wcs_total = round(mean(wcs_total, na.rm=T),1), sd_wcs_total = round(sd(wcs_total, na.rm=T), 1),
    
    mean_rbdsq_total = round(mean(rbdsq_total, na.rm=T),1), sd_rbdsq_total = round(sd(rbdsq_total, na.rm=T), 1),
    
    mean_upsit_total = round(mean(upsit_total, na.rm=T),1), sd_upsit_total = round(sd(upsit_total, na.rm=T), 1),
    
    mean_hads_anxiety_total = round(mean(hads_anxiety_total, na.rm=T),1), sd_hads_anxiety_total = round(sd(hads_anxiety_total, na.rm=T), 1),
    
    mean_hads_depression_total = round(mean(hads_depression_total, na.rm=T),1), sd_hads_depression_total = round(sd(hads_depression_total, na.rm=T), 1),
    
    mean_bdi_total = round(mean(bdi_total, na.rm=T),1), sd_bdi_total = round(sd(bdi_total, na.rm=T), 1),
    
    mean_moca_total = round(mean(moca_total, na.rm=T),1), sd_moca_total = round(sd(moca_total, na.rm=T), 1)
            )

glimpse(table_summ_540ind)

writexl::write_xlsx(table_summ_540ind, "/Users/elisamenozzi/Library/CloudStorage/OneDrive-Personal/Documents/PhD Literature/Upgrade/ASAP MEATADATA/baseline_cohort/final_documents/metadata_full_cohort/clinical_metadata/00_complete_metadata/Metadata_file_final/table_summ_540ind.xlsx")
```

# DEMOGRAPHICS
## COHORT
```{r}
table(metadata_540ind$three_groups)
table(metadata_540ind$three_groups, metadata_540ind$cohort)
table(metadata_540ind$group)
```

## AGE
```{r}
#stats
tapply(metadata_540ind$age, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$age, metadata_540ind$three_groups, FUN = sd, na.rm=T)

res.aov.age <- aov(age ~ three_groups, data = metadata_540ind)
summary(res.aov.age)
pairwise.t.test(metadata_540ind$age, metadata_540ind$three_groups, p.adjust.method = "none", paired = FALSE)

t.test(age ~ three_groups, data = metadata_nmc_hc)
t.test(age ~ three_groups, data = metadata_nmc_pd)
```

## SEX
```{r}
###calculate proportions 
sex_summary = metadata_540ind %>%
              group_by(three_groups, sex) %>%
              summarise("count"= n())
sex_summary

sex_summary$proportion = (sex_summary$count / ave(sex_summary$count, sex_summary$three_groups, FUN = sum))*100
sex_summary = as.data.frame (sex_summary)
sex_summary
```

```{r}
table_sex <- table(metadata_nmc_hc$sex, metadata_nmc_hc$three_groups)
table_sex
chisq.test(table_sex)
```

```{r}
table_sex <- table(metadata_nmc_pd$sex, metadata_nmc_pd$three_groups)
table_sex
chisq.test(table_sex)
```

```{r}
table_sex <- table(metadata_hc_pd$sex, metadata_hc_pd$three_groups)
table_sex
chisq.test(table_sex)
```


## FAMILY HISTORY
```{r}
###calculate proportions for combined positive (first/second degree) vs negative FH
fhx2_summary = metadata_540ind %>%
              group_by(three_groups, fhx2) %>%
              summarise("count"= n())
fhx2_summary

fhx2_summary$proportion = (fhx2_summary$count / ave(fhx2_summary$count, fhx2_summary$three_groups, FUN = sum))*100
fhx2_summary = as.data.frame (fhx2_summary)
fhx2_summary
```

```{r}
table_fhx <- table(metadata_nmc_hc$fhx2, metadata_nmc_hc$three_groups)
table_fhx
chisq.test(table_fhx)
```

```{r}
table_fhx <- table(metadata_nmc_pd$fhx2, metadata_nmc_pd$three_groups)
table_fhx
chisq.test(table_fhx)
```


## EDUCATION
```{r}
#stats
tapply(metadata_540ind$ed_yrs, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$ed_yrs, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(ed_yrs ~ three_groups, data = metadata_nmc_hc)
wilcox.test(ed_yrs ~ three_groups, data = metadata_nmc_pd)
```

## PARTNERS
```{r}
partner_summary = metadata_540ind %>% filter(metadata_540ind$partner_updated != "NA") %>%
              group_by(three_groups, partner_updated) %>%
              summarise("count"= n())
partner_summary

partner_summary$proportion = (partner_summary$count / ave(partner_summary$count, partner_summary$three_groups, FUN = sum))*100
partner_summary = as.data.frame (partner_summary)
partner_summary
```


```{r}
table_partner = table(metadata_nmc_hc$partner_updated, metadata_nmc_hc$three_groups)
table_partner
chisq.test(table_partner)
```

```{r}
table_partner = table(metadata_nmc_pd$partner_updated, metadata_nmc_pd$three_groups)
table_partner
chisq.test(table_partner)
```

## FAMILY MEMBERS
```{r}
fm_summary = metadata_540ind %>% filter(metadata_540ind$blood_related_updated != "NA") %>%
              group_by(three_groups, blood_related_updated) %>%
              summarise("count"= n())
fm_summary

fm_summary$proportion = (fm_summary$count / ave(fm_summary$count, fm_summary$three_groups, FUN = sum))*100
fm_summary = as.data.frame (fm_summary)
fm_summary
```

```{r}
table_fm = table(metadata_nmc_hc$three_groups, metadata_nmc_hc$blood_related_updated)
table_fm
chisq.test(table_fm)
```

```{r}
table_fm = table(metadata_nmc_pd$three_groups, metadata_nmc_pd$blood_related_updated)
table_fm
chisq.test(table_fm)
```


## BMI
```{r}
tapply(metadata_540ind$bmi, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$bmi, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(bmi ~ three_groups, data = metadata_nmc_hc)
wilcox.test(bmi ~ three_groups, data = metadata_nmc_pd)
```


# DISEASE FEATURES
```{r}
onlyPD = filter(metadata_540ind, three_groups == "PD")
dim(onlyPD)
```

## Age at onset
```{r}
#Age at onset
#stats
mean(onlyPD$pd_onset_calculated)
sd(onlyPD$pd_onset_calculated)
```

## Disease duration
```{r}
mean(onlyPD$pd_dur)
sd(onlyPD$pd_dur)
```

## Levodopa equivalent daily dose
```{r}
#stats
mean(onlyPD$ledd_total)
sd(onlyPD$ledd_total)
```

## DRUG NAIVE PD 
```{r}
table(onlyPD$pd_drug_naive)
```

## Advanced therapies
```{r}
table(onlyPD$pd_dbs) 
table(onlyPD$pd_apomorphine) 
table(onlyPD$pd_lcig) 
```

# CLINICAL FEATURES
## HOEHN & YAHR STAGE
```{r}
tapply(metadata_540ind$hoehn_yahr, metadata_540ind$three_groups, FUN = median, na.rm=T)
tapply(metadata_540ind$hoehn_yahr, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(hoehn_yahr ~ three_groups, data = metadata_nmc_hc)
wilcox.test(hoehn_yahr ~ three_groups, data = metadata_nmc_pd)
```

## MDS-UPDRS part I
```{r}
#stats
tapply(metadata_540ind$updrs_i_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_i_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(updrs_i_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(updrs_i_total ~ three_groups, data = metadata_nmc_pd)
```

## MDS-UPDRS part II
```{r}
#stats
tapply(metadata_540ind$updrs_ii_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_ii_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(updrs_ii_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(updrs_ii_total ~ three_groups, data = metadata_nmc_pd)
```


## MDS-UPDRS part III
```{r}
#stats
tapply(metadata_540ind$updrs_iii_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_iii_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(updrs_iii_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(updrs_iii_total ~ three_groups, data = metadata_nmc_pd)
```

## MDS-UPDRS part IV
```{r}
mean(onlyPD$updrs_iv_total, na.rm=T)
sd(onlyPD$updrs_iv_total, na.rm=T)
```

## MDS-UPDRS total score
```{r}
#stats
tapply(metadata_540ind$updrs_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$updrs_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(updrs_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(updrs_total ~ three_groups, data = metadata_nmc_pd)
```

## SCOPA
```{r}
#stats
tapply(metadata_540ind$scopa_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(scopa_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(scopa_total ~ three_groups, data = metadata_nmc_pd)
```

## SCOPA gastrointestinal
```{r}
#stats
tapply(metadata_540ind$scopa_gastrointestinal, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_gastrointestinal, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(scopa_gastrointestinal ~ three_groups, data = metadata_nmc_hc)
wilcox.test(scopa_gastrointestinal ~ three_groups, data = metadata_nmc_pd)
```

## SCOPA urinary
```{r}
#stats
tapply(metadata_540ind$scopa_urinary, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_urinary, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(scopa_urinary ~ three_groups, data = metadata_nmc_hc)
wilcox.test(scopa_urinary ~ three_groups, data = metadata_nmc_pd)
```

## SCOPA cardiovascular
```{r}
#stats
tapply(metadata_540ind$scopa_cardiovascular, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_cardiovascular, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(scopa_cardiovascular ~ three_groups, data = metadata_nmc_hc)
wilcox.test(scopa_cardiovascular ~ three_groups, data = metadata_nmc_pd)
```

## SCOPA thermoregulatory
```{r}
#stats
tapply(metadata_540ind$scopa_thermoregulatory, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_thermoregulatory, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(scopa_thermoregulatory ~ three_groups, data = metadata_nmc_hc)
wilcox.test(scopa_thermoregulatory ~ three_groups, data = metadata_nmc_pd)
```

## SCOPA pupillomotor
```{r}
#stats
tapply(metadata_540ind$scopa_pupillomotor, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_pupillomotor, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(scopa_pupillomotor ~ three_groups, data = metadata_nmc_hc)
wilcox.test(scopa_pupillomotor ~ three_groups, data = metadata_nmc_pd)
```

## SCOPA sexual
```{r}
#stats
tapply(metadata_540ind$scopa_sexual, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$scopa_sexual, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(scopa_sexual ~ three_groups, data = metadata_nmc_hc)
wilcox.test(scopa_sexual ~ three_groups, data = metadata_nmc_pd)
```

## WCSS
```{r}
#stats
tapply(metadata_540ind$wcs_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$wcs_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(wcs_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(wcs_total ~ three_groups, data = metadata_nmc_pd)
```

## RBD
```{r}
#raw score
tapply(metadata_540ind$rbdsq_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$rbdsq_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(rbdsq_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(rbdsq_total ~ three_groups, data = metadata_nmc_pd)
```


```{r}
table(metadata_540ind$three_groups, metadata_540ind$rbd_affected_diff)

###calculate proportion
rbd_summary = metadata_540ind[!is.na(metadata_540ind$rbd_affected_diff),] %>%
              group_by(three_groups, rbd_affected_diff) %>%
              summarise("count"= n())
rbd_summary

rbd_summary$proportion = (rbd_summary$count / ave(rbd_summary$count, rbd_summary$three_groups, FUN = sum))*100
rbd_summary = as.data.frame (rbd_summary)
rbd_summary
```

```{r}
rbd.model = glm(as.factor(rbd_affected_diff) ~ three_groups, data = metadata_nmc_hc, family = binomial())
summary(rbd.model)
summary(rbd.model)$coefficients
odds.ratio(rbd.model, level = 0.95, signif.stars = TRUE)
```

```{r}
rbd.model = glm(as.factor(rbd_affected_diff) ~ three_groups, data = metadata_nmc_pd, family = binomial())
summary(rbd.model)
summary(rbd.model)$coefficients
odds.ratio(rbd.model, level = 0.95, signif.stars = TRUE)
```

## UPSIT
```{r}
#stats
tapply(metadata_540ind$upsit_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$upsit_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)
```

```{r}
res.regression.upsit <- lm(upsit_total ~ three_groups + age + sex, 
            data = metadata_nmc_hc)

summary(res.regression.upsit)
summary(res.regression.upsit)$coef
vif(res.regression.upsit)
```

```{r}
res.regression.upsit <- lm(upsit_total ~ three_groups + age + sex, 
            data = metadata_nmc_pd)

summary(res.regression.upsit)
summary(res.regression.upsit)$coef
vif(res.regression.upsit)
```

## HADS depression
```{r}
#stats
tapply(metadata_540ind$hads_depression_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$hads_depression_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(hads_depression_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(hads_depression_total ~ three_groups, data = metadata_nmc_pd)
```

## HADS anxiety
```{r}
#stats
tapply(metadata_540ind$hads_anxiety_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$hads_anxiety_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(hads_anxiety_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(hads_anxiety_total ~ three_groups, data = metadata_nmc_pd)
```


## BDI
```{r}
#stats
tapply(metadata_540ind$bdi_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$bdi_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)

wilcox.test(bdi_total ~ three_groups, data = metadata_nmc_hc)
wilcox.test(bdi_total ~ three_groups, data = metadata_nmc_pd)
```

## MOCA adjusted score
```{r}
#stats
tapply(metadata_540ind$moca_total, metadata_540ind$three_groups, FUN = mean, na.rm=T)
tapply(metadata_540ind$moca_total, metadata_540ind$three_groups, FUN = sd, na.rm=T)
```

```{r}
moca_model = lm(formula = moca_total ~ three_groups + age + ed_yrs + sex, data = metadata_nmc_hc)
summary(moca_model)
summary(moca_model)$coefficients
```

```{r}
moca_model = lm(formula = moca_total ~ three_groups + age + ed_yrs + sex, data = metadata_nmc_pd)
summary(moca_model)
summary(moca_model)$coefficients
```

## MOCA ranges
```{r}
moca_summary = metadata_540ind %>% filter(metadata_540ind$moca_range != "NA") %>%
              group_by(three_groups, moca_range) %>%
              summarise("count"= n())
moca_summary

moca_summary$proportion = (moca_summary$count / ave(moca_summary$count, moca_summary$three_groups, FUN = sum))*100
moca_summary = as.data.frame (moca_summary)
moca_summary
```

```{r}
metadata_nmc_hc = metadata_nmc_hc %>% reorder_levels(three_groups, order = c("HC", "GBA_NMC"))
metadata_nmc_hc = metadata_nmc_hc %>% reorder_levels(moca_range, order = c("Normal", "Abnormal"))
metadata_nmc_hc$moca_range = as.factor(metadata_nmc_hc$moca_range)
moca.model = glm(moca_range ~ three_groups + age + education_n_years + sex, data = metadata_nmc_hc, family = binomial())
summary(moca.model)
summary(moca.model)$coefficients
odds.ratio(moca.model, level = 0.95, signif.stars = TRUE)
```


```{r}
metadata_nmc_pd = metadata_nmc_pd %>% reorder_levels(three_groups, order = c("GBA_NMC", "PD"))
metadata_nmc_pd = metadata_nmc_pd %>% reorder_levels(moca_range, order = c("Normal", "Abnormal"))
metadata_nmc_pd$moca_range = as.factor(metadata_nmc_pd$moca_range)
moca.model = glm(moca_range ~ three_groups + age + education_n_years + sex, data = metadata_nmc_pd, family = binomial())
summary(moca.model)
summary(moca.model)$coefficients
odds.ratio(moca.model, level = 0.95, signif.stars = TRUE)
```

## PRODROMAL PROBABILITY
```{r}
table(metadata_540ind$three_groups, metadata_540ind$prob_prodromalPD)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
